---
title: "Ligand–Receptor Recruitment Along the Classical Niche Ablation Response Trajectory"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
date: "2026-04-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 5)
```

```{r libraries}
suppressPackageStartupMessages({
  library(anndata)
  library(Matrix)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(ggthemes)
  library(RColorBrewer)
  library(patchwork)
  library(here)
})

# Source helper functions
source("basal_ablation/src/xenium_utils/pl/lr_heatmap_helpers.R")
RETICULATE_PYTHON='.conda/envs/xenium-rapids/bin/python'


CELLCHAT_PATH <- "ref_data/Mus_musculus/rl/CellChatDB.mouse.v2.rda"
H5AD_PATH     <- "basal_ablation/data/processed/xenium/niche_trajectory_classical_pruned_d24_filterv2_d1rm/st_data_classical_niche_DC_env.h5ad"
FIG_DIR       <- "basal_ablation/results/figures/chemokine_dc1/"

dir.create(FIG_DIR, recursive = TRUE, showWarnings = FALSE)

rose_theme <- theme(
  axis.text  = element_text(size = 13, color = "black"),
  axis.title = element_text(size = 14, face = "bold")
)

X_LAB   <- "Classical niche ablation response (DC1)"
MIN_PCT           <- 0.009   # min fraction cells expressing ligand within lineage
MIN_LIG_CELLS     <- 500     # min cells expressing ligand within lineage (absolute count)
# Ligand passes filter if EITHER condition is met (OR logic)
MIN_REC_FRAC      <- 0.01    # min fraction receptor+ cells within lineage (overall, not per-bin)

# Cell-annotation level filters (cell type mode) - more stringent than lineage
MIN_PCT_CELLTYPE       <- 0.02    # min fraction cells expressing ligand within cell type
MIN_LIG_CELLS_CELLTYPE <- 300     # min cells expressing ligand within cell type
MIN_REC_FRAC_CELLTYPE  <- 0.02    # min receptor+ fraction within cell type

# Chemokine gene lists — for annotation/filtering of results, not for restricting tested pairs
CHEMOKINE_LIGANDS   <- c("Ccl2", "Ccl5", "Ccl17",
                          "Cxcl9", "Cxcl10", "Cxcl12", "Cxcl13", "Cxcl16", "Mif", 'Ccl9', 'Cxcl1')
CHEMOKINE_RECEPTORS <- c("Ccr2", "Ccr4", "Ccr6", "Ccr7",
                          "Cx3cr1", "Cxcr3", "Cxcr4", "Cxcr5", "Cxcr6", "Xcr1")

# Type I IFN ligands — sparse but important; use relaxed filtering
IFN_I_LIGANDS <- c("Ifna2", "Ifna13", "Ifne", "Ifnb1")
MIN_IFN_CELLS <- 30  # relaxed threshold for IFN-I ligands

LINEAGE_MAP <- c(
  Classical = 'Cancer', 
  Basal = 'Cancer', 
  Mesenchymal = 'Cancer', 
  # Cancer
  Cancer_epithelial              = "Cancer",
  `Cancer_epithelial_neuronal-like` = "Cancer",
  Cancer_mesenchymal             = "Cancer",
  # Myeloid
  Macrophage             = "Mac_Mono",
  Monocyte               = "Mac_Mono",
  cDC1                   = "DC",
  cDC2                   = "DC",
  mregDC                 = "DC",
  pDC                    = "pDC",
  Neutrophil             = "Neutrophil",
  Granulocyte            = "Neutrophil",
  # Stromal
  Fibroblast             = "Fibroblast",
  `Mesothelial/apCAF` = 'Fibroblast',
  PSC                    = "Fibroblast",
  `Mesenchymal-like_stroma` = "Stroma",
  Mesothelial            = "Stroma",
  Mural                  = "Stroma",
  Smooth_muscle          = "Stroma",
  Adipocyte              = "Stroma",
  Pericyte = 'Stroma',
  # Lymphoid
  CD8_T                  = "T_NK_ILC",
  Tconv                  = "T_NK_ILC",
  NK                     = "T_NK_ILC",
  Treg                   = "T_NK_ILC",
  gdT                    = "T_NK_ILC",
  gdT_CD8                = "T_NK_ILC",
  MAIT                   = "T_NK_ILC",
  `Innate-like_T`        = "T_NK_ILC",
  ILC1                   = "T_NK_ILC",
  ILC2                   = "T_NK_ILC",
  ILC3                   = "T_NK_ILC",
  NK_ILC1 = 'T_NK_ILC', 
  Lymphoid_proliferating = "T_NK_ILC",
  B_cell                 = "B_Plasma",
  Plasma_cell = 'B_Plasma', 
  Plasmablast            = "B_Plasma",
  # Endothelial
  Endothelial_vascular   = "Endothelial",
  Endothelial_lymphatic  = "Endothelial",
  # Epithelial (normal)
  Acinar                 = "Epithelial",
  ADM                    = "Epithelial",
  Ductal                 = "Epithelial",
  Islet                  = "Epithelial",
  `Tuft-like`            = "Epithelial",
  # Neural
  Neuron                 = "Neural",
  Schwann_cell           = "Neural"
)


# Cell types for cell type mode (cell_annotation level) - populated after data loading
SENDER_CELLTYPES   <- NULL
RECEIVER_CELLTYPES <- NULL

DC1_N_BINS = 30

set.seed(42)
```

---

# Introduction

This notebook tests ligand → receptor recruitment relationships along the **classical niche ablation response trajectory** (DC\_1\_rank\_flip). This axis captures how the tumor microenvironment of the classical PDAC niche evolves in response to basal cell ablation: DC1 ≈ 0 corresponds to the untreated classical niche; DC1 ≈ 1 corresponds to the late ablation-induced immune response. 

**Statistical approach:** Tier1 (lm) and tier2 (local niche correlation) used to identify LR relationships. Environment cells are assigned DC1 bins matching cancer cell definitions, and analysis is restricted to bins 0-27. This is tested at the cell type and total niche level. 

**LR pairs tested:** All CellChatDB pairs where both the ligand and at least one receptor subunit gene are present in the 480-gene Xenium panel. Chemokine-specific pairs are annotated in results but all pairs are tested.

---

# Data Loading

```{r load_adata}
adata  <- read_h5ad(H5AD_PATH)
obs_df <- adata$obs
# set a column that will indicate if a niche was part of the original trajectory
obs_df$in_trajectory = !(is.na(obs_df$DC_1_rank))

cat("Loaded:", nrow(obs_df), "cells,", ncol(adata$X), "genes\n")
```

```{r build_lr_pairs}
e_cc   <- new.env()
load(CELLCHAT_PATH, envir = e_cc)
cc_int <- e_cc$CellChatDB.mouse$interaction

# For each interaction, find the first receptor subunit gene present in the panel
cc_int$receptor_gene <- sapply(cc_int$receptor.symbol, function(x) {
  syms <- trimws(unlist(strsplit(x, ",")))
  hit  <- syms[syms %in% adata$var_names]
  if (length(hit) == 0) NA_character_ else hit[1]
})

LR_PAIRS <- cc_int[
  (cc_int$ligand %in% adata$var_names | cc_int$receptor_gene %in% adata$var_names) &
  !is.na(cc_int$receptor_gene),
  c("ligand", "receptor", "receptor.symbol", "receptor_gene", "pathway_name")
]
LR_PAIRS <- LR_PAIRS[!duplicated(paste(LR_PAIRS$ligand, LR_PAIRS$receptor_gene)), ]
rownames(LR_PAIRS) <- NULL

# Annotate chemokine pairs
LR_PAIRS$is_chemokine <- LR_PAIRS$ligand %in% CHEMOKINE_LIGANDS &
                          LR_PAIRS$receptor_gene %in% CHEMOKINE_RECEPTORS
# Expanded: any pair with a chemokine ligand (regardless of receptor)
LR_PAIRS$is_chemokine_expanded <- LR_PAIRS$ligand %in% CHEMOKINE_LIGANDS

cat("LR pairs to test:", nrow(LR_PAIRS), "\n")
cat("  of which chemokine LR:", sum(LR_PAIRS$is_chemokine), "\n")
cat("  of which chemokine expanded:", sum(LR_PAIRS$is_chemokine_expanded), "\n")
print(LR_PAIRS[LR_PAIRS$is_chemokine, c("ligand", "receptor_gene", "pathway_name")])
```


## Load cytokine annotations

```{r load_cytokine_anno}
library(readxl)

SIGNAL_ANNO_PATH <- "basal_ablation/data/features/xenium_panel/44CXV9_mPancreas_480g_gene_list_annotated.xlsx"

# Helper function to convert GMT-like format to named list
# GMT format: col1 = program name, col2 = description (ignored), col3+ = genes
gmt_to_list <- function(df) {
  result <- list()
  for (i in seq_len(nrow(df))) {
    program_name <- as.character(df[i, 1])
    # Get genes from column 3 onwards, removing NAs
    genes <- as.character(df[i, 3:ncol(df)])
    genes <- genes[!is.na(genes) & genes != ""]
    result[[program_name]] <- genes
  }
  result
}

signal_anno <- read_excel(SIGNAL_ANNO_PATH, sheet = "signal_anno_broad", col_names = FALSE)
gene_anno <- read_excel(SIGNAL_ANNO_PATH, sheet = "44CXV9_mPancreas_480g_gene_list", col_names = T)

# Convert GMT format to named list of gene sets
cytokine_signatures <- gmt_to_list(signal_anno)
cat("Loaded", length(cytokine_signatures), "cytokine programs\n")
for (prog in names(cytokine_signatures)) {
  cat("  ", prog, ":", length(cytokine_signatures[[prog]]), "genes\n")
}

panel_cyto_genes = cc_int %>% dplyr::filter(ligand.symbol %in% gene_anno$Gene) %>% .$ligand.symbol %>% unique()

# Add Ifn_I pseudo-gene (summed IFN-I ligands) and individual IFN-I ligands to panel_cyto_genes
panel_cyto_genes <- c(panel_cyto_genes, "Ifn_I", IFN_I_LIGANDS)
panel_cyto_genes <- unique(panel_cyto_genes)
cat("Panel cytokine genes:", length(panel_cyto_genes), "(includes Ifn_I pseudo-gene and IFN-I ligands)\n")
```

---

# Data Preparation

```{r impute_dc1}
obs_df$DC_1_rank = ifelse(is.na(obs_df$DC_1_rank), obs_df$DC_1_rank_env, obs_df$DC_1_rank)
sum(is.na(obs_df$DC_1_rank))
```

```{r extract_genes}
# Extract all ligand and receptor genes in LR_PAIRS
all_lr_genes   <- unique(c(LR_PAIRS$ligand, LR_PAIRS$receptor_gene, unique(unlist(cytokine_signatures))))
genes_in_panel <- intersect(all_lr_genes, adata$var_names)
cat("LR genes in panel:", length(genes_in_panel), "/", length(all_lr_genes), "\n")

# Raw counts (for receptor thresholding)
X        <- adata$layers[["counts_segger"]] %>% as.matrix()
gene_idx <- match(genes_in_panel, adata$var_names)
for (i in seq_along(genes_in_panel))
  obs_df[[genes_in_panel[i]]] <- as.numeric(X[, gene_idx[i]])

# Log1p normalized expression (for ligand sender analysis)
X_norm   <- adata$layers[["log1p_norm"]] %>% as.matrix()
all_lr_ligands_panel <- intersect(unique(LR_PAIRS$ligand), adata$var_names)
lig_idx <- match(all_lr_ligands_panel, adata$var_names)
for (i in seq_along(all_lr_ligands_panel))
  obs_df[[paste0(all_lr_ligands_panel[i], "_norm")]] <- as.numeric(X_norm[, lig_idx[i]])

obs_df$lib_size     <- rowSums(X)
obs_df$log_lib_size <- log1p(obs_df$lib_size)
obs_df$cell_int_idx <- as.character(seq_len(nrow(obs_df)) - 1L)
obs_df$lineage      <- LINEAGE_MAP[as.character(obs_df$cell_annotation)]

# Set cell types for cell type mode analysis (cell_annotation level)
# Filter to cell types with at least MIN_CELLS_CELLTYPE cells
MIN_CELLS_CELLTYPE <- 200
ct_counts <- table(obs_df$cell_annotation[!is.na(obs_df$cell_annotation)])
SENDER_CELLTYPES   <- names(ct_counts[ct_counts >= MIN_CELLS_CELLTYPE])
RECEIVER_CELLTYPES <- SENDER_CELLTYPES
cat("Cell types for cell type mode:", length(SENDER_CELLTYPES), "of", length(ct_counts),
    "(filtered to n >=", MIN_CELLS_CELLTYPE, ")\n")

# check for NA lineage conversion
table(obs_df$lineage, useNA = 'always')
```



```{r dc1_trim}
DC1_N_BINS    <- 30L
DC1_TRIM_LOW  <- 0L
DC1_TRIM_HIGH <- 2L

obs_df$DC1_bin30 <- cut(obs_df$DC_1_rank, breaks = DC1_N_BINS, labels = FALSE)
bin_breaks <- seq(min(obs_df$DC_1_rank, na.rm = TRUE),
                  max(obs_df$DC_1_rank, na.rm = TRUE),
                  length.out = DC1_N_BINS + 1)
bin_lines  <- bin_breaks[(DC1_TRIM_LOW + 1):(DC1_N_BINS - DC1_TRIM_HIGH)]

obs_df <- obs_df[
  !is.na(obs_df$DC1_bin30) &
  obs_df$DC1_bin30 > DC1_TRIM_LOW &
  obs_df$DC1_bin30 <= (DC1_N_BINS - DC1_TRIM_HIGH), ]

DC1_T_MIN <- min(obs_df$DC_1_rank, na.rm = TRUE)
DC1_T_MAX <- max(obs_df$DC_1_rank, na.rm = TRUE)

cat("Cells after trimming:", nrow(obs_df), "\n")
cat("DC1 range:", round(DC1_T_MIN, 3), "–", round(DC1_T_MAX, 3), "\n")
print(sort(table(obs_df$lineage), decreasing = TRUE))
```

What are the cells per bin?

```{r}
table(obs_df %>% dplyr::filter(in_trajectory) %>% .$DC1_bin30) %>% unique()
```


## Precompute niche receptor counts

```{r niche_counts}
conn   <- adata$obsp[["mingle_connectivities"]]
conn@x <- rep(1.0, length(conn@x))

keep_idx  <- as.integer(obs_df$cell_int_idx) + 1L
conn_trim <- conn[keep_idx, keep_idx]

obs_df$n_niche_total <- as.numeric(conn_trim %*% rep(1.0, nrow(obs_df)))

# Precompute for all LR receptor genes × receiver lineages
all_lr_receptors <- intersect(unique(LR_PAIRS$receptor_gene), genes_in_panel)

# --- VECTORIZED: Lineage niche counts (one matrix multiplication) ---
lineage_indicators <- sapply(RECEIVER_LINEAGES, function(lin) {
  as.numeric(!is.na(obs_df$lineage) & obs_df$lineage == lin)
})
niche_lineage_counts <- as.matrix(conn_trim %*% lineage_indicators)
for (i in seq_along(RECEIVER_LINEAGES)) {
  obs_df[[paste0("n_", RECEIVER_LINEAGES[i], "_niche")]] <- niche_lineage_counts[, i]
}

# --- VECTORIZED: Receptor-positive × lineage counts (one matrix multiplication) ---
combo_list <- list()
combo_names <- character(0)
for (lin in RECEIVER_LINEAGES) {
  is_lin <- !is.na(obs_df$lineage) & obs_df$lineage == lin
  for (rec in all_lr_receptors) {
    nm <- paste0("n_", rec, "_", lin)
    combo_names <- c(combo_names, nm)
    combo_list[[nm]] <- as.numeric(is_lin & obs_df[[rec]] > 1)
  }
}
indicator_mat <- do.call(cbind, combo_list)
niche_rec_counts <- as.matrix(conn_trim %*% indicator_mat)
for (i in seq_along(combo_names)) {
  obs_df[[combo_names[i]]] <- niche_rec_counts[, i]
}

cat("Niche counts computed for", length(RECEIVER_LINEAGES), "receiver lineages,",
    length(all_lr_receptors), "receptor genes\n")
```

## Compute niche cell type counts

```{r niche_celltype_counts}
# --- Cell-type level niche counts (for cell type mode) ---
celltype_indicators <- sapply(SENDER_CELLTYPES, function(ct) {
  as.numeric(!is.na(obs_df$cell_annotation) & obs_df$cell_annotation == ct)
})
niche_celltype_counts <- as.matrix(conn_trim %*% celltype_indicators)
for (i in seq_along(SENDER_CELLTYPES)) {
  obs_df[[paste0("n_", SENDER_CELLTYPES[i], "_niche_ct")]] <- niche_celltype_counts[, i]
}

# Receptor-positive × cell type counts
combo_list_ct <- list()
combo_names_ct <- character(0)
for (ct in RECEIVER_CELLTYPES) {
  is_ct <- !is.na(obs_df$cell_annotation) & obs_df$cell_annotation == ct
  for (rec in all_lr_receptors) {
    nm <- paste0("n_", rec, "_", ct, "_ct")
    combo_names_ct <- c(combo_names_ct, nm)
    combo_list_ct[[nm]] <- as.numeric(is_ct & obs_df[[rec]] > 1)
  }
}
indicator_mat_ct <- do.call(cbind, combo_list_ct)
niche_rec_counts_ct <- as.matrix(conn_trim %*% indicator_mat_ct)
for (i in seq_along(combo_names_ct)) {
  obs_df[[combo_names_ct[i]]] <- niche_rec_counts_ct[, i]
}

cat("Cell type niche counts computed for", length(SENDER_CELLTYPES), "cell types,",
    length(all_lr_receptors), "receptor genes\n")
```

## Compute niche-summed ligand (legacy - kept for niche mode)

```{r niche_ligand_compute}
cancer_mask <- obs_df$in_trajectory
cancer_pos  <- which(cancer_mask)

# Total transcript counts per retained cell (for normalisation)
total_counts     <- as.numeric(rowSums(X[keep_idx, ]))
niche_total_cnts <- as.numeric(conn_trim[cancer_pos, ] %*% total_counts)

# --- VECTORIZED: All ligands in one matrix multiplication ---
all_lr_ligands <- intersect(unique(LR_PAIRS$ligand), genes_in_panel)
ligand_mat <- sapply(all_lr_ligands, function(lig) as.numeric(obs_df[[lig]]))
niche_lig_sums <- as.matrix(conn_trim[cancer_pos, ] %*% ligand_mat)

for (i in seq_along(all_lr_ligands)) {
  obs_df[cancer_mask, paste0("niche_lig_", all_lr_ligands[i])] <-
    niche_lig_sums[, i] / pmax(niche_total_cnts, 1)
}

# Add Ifn_I pseudo-gene column (sum of IFN-I ligands)
ifn_genes_in_panel <- intersect(IFN_I_LIGANDS, names(obs_df))
if (length(ifn_genes_in_panel) > 0) {
  obs_df$Ifn_I <- rowSums(obs_df[, ifn_genes_in_panel, drop = FALSE], na.rm = TRUE)
  all_lr_ligands <- c(all_lr_ligands, "Ifn_I")
  cat("Added Ifn_I pseudo-gene to obs_df (sum of", paste(ifn_genes_in_panel, collapse = ", "), ")\n")
}

# Add Ifn_I pseudo-gene: sum of all IFN-I ligands in niche
ifn_genes_in_panel <- intersect(IFN_I_LIGANDS, names(obs_df))
if (length(ifn_genes_in_panel) > 0) {
  ifn_mat <- sapply(ifn_genes_in_panel, function(g) as.numeric(obs_df[[g]]))
  ifn_sum <- rowSums(ifn_mat, na.rm = TRUE)
  niche_ifn_sum <- as.numeric(conn_trim[cancer_pos, ] %*% ifn_sum)
  obs_df[cancer_mask, "niche_lig_Ifn_I"] <- niche_ifn_sum / pmax(niche_total_cnts, 1)
  cat("Added niche_lig_Ifn_I pseudo-gene (sum of", paste(ifn_genes_in_panel, collapse = ", "), ")\n")
}

cat("Niche ligand columns computed:", sum(grepl("^niche_lig_", names(obs_df))), "\n")
```

## Compute niche-summed receptor fraction (niche mode)

```{r niche_receptor_compute}
# For each cancer cell, compute the fraction of niche cells positive for each receptor
# This is: sum(receptor+ cells in niche) / sum(all cells in niche)
# Summed across ALL cell types (no lineage restriction)

# Total niche cells per cancer cell
niche_total_cells <- as.numeric(conn_trim[cancer_pos, ] %*% rep(1.0, nrow(obs_df)))

# Compute niche receptor+ fraction for each receptor (across all cell types)
for (rec in all_lr_receptors) {
  is_rec_pos <- obs_df[[rec]] > 1
  niche_rec_cnt <- as.numeric(conn_trim[cancer_pos, ] %*% as.numeric(is_rec_pos))
  col_name <- paste0("niche_rec_", rec)
  obs_df[cancer_mask, col_name] <- niche_rec_cnt / pmax(niche_total_cells, 1)
}
cat("Niche receptor columns computed:", sum(grepl("^niche_rec_", names(obs_df))), "\n")
```

---

# Bin-Level Aggregation

```{r bin_aggregate}
cancer_df <- obs_df[cancer_mask & !is.na(obs_df$DC1_bin30), ]

# Covariate means per bin (cancer cells)
cov_bins <- cancer_df %>%
  group_by(DC1_bin30) %>%
  summarize(
    n_cancer       = n(),
    .groups = "drop"
  )

# Total niche cells per bin (across ALL lineages)
niche_totals_bin <- cancer_df %>%
  group_by(DC1_bin30) %>%
  summarize(
    n_niche_total_bin = sum(n_niche_total, na.rm = TRUE),
    .groups = "drop"
  )

# Receptor-positive counts per bin per lineage
rec_bins <- list()
for (rec in all_lr_receptors) {
  for (lin_r in RECEIVER_LINEAGES) {
    rec_col <- paste0("n_", rec, "_", lin_r)
    lin_col <- paste0("n_", lin_r, "_niche")
    if (!rec_col %in% names(cancer_df)) next

    tmp <- cancer_df %>%
      group_by(DC1_bin30) %>%
      summarize(
        n_rec_total  = sum(.data[[rec_col]], na.rm = TRUE),
        n_lin_total  = sum(.data[[lin_col]], na.rm = TRUE),
        .groups = "drop"
      ) %>%
      left_join(cov_bins, by = "DC1_bin30") %>%
      left_join(niche_totals_bin, by = "DC1_bin30")

    # Fraction within lineage per bin (for reference)
    tmp$rec_frac_within_lin <- tmp$n_rec_total / pmax(tmp$n_lin_total, 1)
    # Fraction of total niche (for modeling/plotting)
    tmp$rec_frac_bin <- tmp$n_rec_total / pmax(tmp$n_niche_total_bin, 1)
    # Overall fraction within lineage (summed across all bins, for filtering)
    tmp$rec_frac_within_lin_overall <- sum(tmp$n_rec_total, na.rm = TRUE) /
                                        pmax(sum(tmp$n_lin_total, na.rm = TRUE), 1)

    rec_bins[[paste(rec, lin_r, sep = "__")]] <- tmp
  }
}
# Filter: remove if overall receptor+ fraction within lineage is below threshold
rec_bins <- Filter(function(df) df$rec_frac_within_lin_overall[1] >= MIN_REC_FRAC, rec_bins)
cat("Receptor bin tables after filtering:", length(rec_bins), "\n")

# Sender expression per bin - pseudobulk: sum counts / total counts, then log1p
sender_bins <- list()
for (lin_s in SENDER_LINEAGES) {
  sub <- obs_df[!is.na(obs_df$lineage) & obs_df$lineage == lin_s &
                !is.na(obs_df$DC1_bin30), ]
  if (nrow(sub) < 10) next
  for (lig in all_lr_ligands) {
    lig_norm_col <- paste0(lig, "_norm")
    if (!lig %in% names(sub) || !lig_norm_col %in% names(sub)) next

    # Calculate expression stats within lineage
    lig_pct <- mean(sub[[lig]] > 0, na.rm = TRUE)
    lig_n_cells <- sum(sub[[lig]] > 0, na.rm = TRUE)

    # Filter: pass if EITHER pct >= MIN_PCT OR n_cells >= MIN_LIG_CELLS (OR logic)
    # IFN-I ligands use relaxed cell count threshold
    min_cells_thresh <- if (lig %in% IFN_I_LIGANDS) MIN_IFN_CELLS else MIN_LIG_CELLS
    if (lig_pct < MIN_PCT && lig_n_cells < min_cells_thresh) next

    tmp <- sub %>%
      group_by(DC1_bin30) %>%
      summarize(lig_expr_bin = log1p(sum(.data[[lig]], na.rm = TRUE) /
                                      pmax(sum(lib_size, na.rm = TRUE), 1)),
                n_sender     = n(),
                n_lig_cells  = sum(.data[[lig]] > 0, na.rm = TRUE),
                .groups = "drop")
    sender_bins[[paste(lin_s, lig, sep = "__")]] <- tmp
  }

  # Create Ifn_I pseudo-gene: sum of all IFN-I ligands
  ifn_genes_in_panel <- intersect(IFN_I_LIGANDS, names(sub))
  if (length(ifn_genes_in_panel) > 0) {
    # Sum raw counts across IFN-I ligands
    sub$Ifn_I_counts <- rowSums(sub[, ifn_genes_in_panel, drop = FALSE], na.rm = TRUE)
    n_ifn_cells_total <- sum(sub$Ifn_I_counts > 0, na.rm = TRUE)

    if (n_ifn_cells_total >= MIN_IFN_CELLS) {
      tmp_ifn <- sub %>%
        group_by(DC1_bin30) %>%
        summarize(lig_expr_bin = log1p(sum(Ifn_I_counts, na.rm = TRUE) /
                                        pmax(sum(lib_size, na.rm = TRUE), 1)),
                  n_sender     = n(),
                  n_lig_cells  = sum(Ifn_I_counts > 0, na.rm = TRUE),
                  .groups = "drop")
      sender_bins[[paste(lin_s, "Ifn_I", sep = "__")]] <- tmp_ifn
    }
  }
}
cat("Sender bin tables:", length(sender_bins), "\n")
cat("Ifn_I entries added:", sum(grepl("__Ifn_I$", names(sender_bins))), "\n")

# Niche-summed ligand per bin 
niche_lig_bins <- list()
for (lig in all_lr_ligands) {
  ncol_name <- paste0("niche_lig_", lig)
  if (!ncol_name %in% names(obs_df)) next
  tmp <- cancer_df %>%
    group_by(DC1_bin30) %>%
    summarize(niche_lig_bin = mean(.data[[ncol_name]], na.rm = TRUE), .groups = "drop") %>%
    left_join(cov_bins, by = "DC1_bin30")
  # Keep only if at least one bin meets minimum niche ligand fraction
  if (!any(tmp$niche_lig_bin >= 0, na.rm = TRUE)) next # don't need to filter here
  niche_lig_bins[[lig]] <- tmp
}

# Add Ifn_I pseudo-gene to niche_lig_bins
if ("niche_lig_Ifn_I" %in% names(cancer_df)) {
  tmp_ifn <- cancer_df %>%
    group_by(DC1_bin30) %>%
    summarize(niche_lig_bin = mean(niche_lig_Ifn_I, na.rm = TRUE), .groups = "drop") %>%
    left_join(cov_bins, by = "DC1_bin30")
  niche_lig_bins[["Ifn_I"]] <- tmp_ifn
  cat("Added Ifn_I to niche_lig_bins\n")
}

cat("Niche ligand bin tables after filtering:", length(niche_lig_bins), "\n")

# Niche-summed receptor fraction per bin - across all cell types
niche_rec_bins <- list()
for (rec in all_lr_receptors) {
  col_name <- paste0("niche_rec_", rec)
  if (!col_name %in% names(cancer_df)) next
  tmp <- cancer_df %>%
    group_by(DC1_bin30) %>%
    summarize(niche_rec_bin = mean(.data[[col_name]], na.rm = TRUE), .groups = "drop") %>%
    left_join(cov_bins, by = "DC1_bin30")
  # Keep only if at least one bin meets minimum receptor fraction
  if (!any(tmp$niche_rec_bin >= 0, na.rm = TRUE)) next
  niche_rec_bins[[rec]] <- tmp
}
cat("Niche receptor bin tables after filtering:", length(niche_rec_bins), "\n")

# --- Cell-type level sender bins ---
sender_celltype_bins <- list()
for (ct_s in SENDER_CELLTYPES) {
  sub <- obs_df[!is.na(obs_df$cell_annotation) & obs_df$cell_annotation == ct_s &
                !is.na(obs_df$DC1_bin30), ]
  if (nrow(sub) < 10) next
  for (lig in all_lr_ligands) {
    lig_norm_col <- paste0(lig, "_norm")
    if (!lig %in% names(sub) || !lig_norm_col %in% names(sub)) next

    lig_pct <- mean(sub[[lig]] > 0, na.rm = TRUE)
    lig_n_cells <- sum(sub[[lig]] > 0, na.rm = TRUE)

    # More stringent filtering for cell type level (OR logic)
    if (lig_pct < MIN_PCT_CELLTYPE && lig_n_cells < MIN_LIG_CELLS_CELLTYPE) next

    tmp <- sub %>%
      group_by(DC1_bin30) %>%
      summarize(lig_expr_bin = log1p(sum(.data[[lig]], na.rm = TRUE) /
                                      pmax(sum(lib_size, na.rm = TRUE), 1)),
                n_sender     = n(),
                n_lig_cells  = sum(.data[[lig]] > 0, na.rm = TRUE),
                .groups = "drop")
    sender_celltype_bins[[paste(ct_s, lig, sep = "__")]] <- tmp
  }

  # Create Ifn_I pseudo-gene: sum of all IFN-I ligands
  ifn_genes_in_panel <- intersect(IFN_I_LIGANDS, names(sub))
  if (length(ifn_genes_in_panel) > 0) {
    sub$Ifn_I_counts <- rowSums(sub[, ifn_genes_in_panel, drop = FALSE], na.rm = TRUE)
    n_ifn_cells_total <- sum(sub$Ifn_I_counts > 0, na.rm = TRUE)

    if (n_ifn_cells_total >= MIN_IFN_CELLS) {
      tmp_ifn <- sub %>%
        group_by(DC1_bin30) %>%
        summarize(lig_expr_bin = log1p(sum(Ifn_I_counts, na.rm = TRUE) /
                                        pmax(sum(lib_size, na.rm = TRUE), 1)),
                  n_sender     = n(),
                  n_lig_cells  = sum(Ifn_I_counts > 0, na.rm = TRUE),
                  .groups = "drop")
      sender_celltype_bins[[paste(ct_s, "Ifn_I", sep = "__")]] <- tmp_ifn
    }
  }
}
cat("Sender cell type bin tables:", length(sender_celltype_bins), "\n")
cat("Ifn_I entries added:", sum(grepl("__Ifn_I$", names(sender_celltype_bins))), "\n")

# --- Cell-type level receptor bins ---
rec_celltype_bins <- list()
for (rec in all_lr_receptors) {
  for (ct_r in RECEIVER_CELLTYPES) {
    rec_col <- paste0("n_", rec, "_", ct_r, "_ct")
    ct_col <- paste0("n_", ct_r, "_niche_ct")
    if (!rec_col %in% names(cancer_df)) next

    tmp <- cancer_df %>%
      group_by(DC1_bin30) %>%
      summarize(
        n_rec_total  = sum(.data[[rec_col]], na.rm = TRUE),
        n_ct_total   = sum(.data[[ct_col]], na.rm = TRUE),
        .groups = "drop"
      ) %>%
      left_join(cov_bins, by = "DC1_bin30") %>%
      left_join(niche_totals_bin, by = "DC1_bin30")

    tmp$rec_frac_within_ct <- tmp$n_rec_total / pmax(tmp$n_ct_total, 1)
    tmp$rec_frac_bin <- tmp$n_rec_total / pmax(tmp$n_niche_total_bin, 1)
    tmp$rec_frac_within_ct_overall <- sum(tmp$n_rec_total, na.rm = TRUE) /
                                       pmax(sum(tmp$n_ct_total, na.rm = TRUE), 1)

    rec_celltype_bins[[paste(rec, ct_r, sep = "__")]] <- tmp
  }
}
# More stringent filtering for cell type level
rec_celltype_bins <- Filter(function(df) df$rec_frac_within_ct_overall[1] >= MIN_REC_FRAC_CELLTYPE, rec_celltype_bins)
cat("Receptor cell type bin tables after filtering:", length(rec_celltype_bins), "\n")
```

## Filtering diagnostics

```{r filtering_diagnostics}
# Compute filtering metrics for all gene × cell_type combinations (vectorized)
compute_filtering_metrics <- function(df, genes, cell_types, col_name = "cell_annotation",
                                       threshold = 0) {
  # Filter to genes that exist in data

  genes <- intersect(genes, names(df))
  if (length(genes) == 0) return(data.frame())

  # Filter to valid cell types

  df <- df[!is.na(df[[col_name]]) & df[[col_name]] %in% cell_types, ]
  if (nrow(df) == 0) return(data.frame())


  # Compute n_total per cell type (single group_by)
  n_total_df <- df %>%
    group_by(cell_type = .data[[col_name]]) %>%
    summarize(n_total = n(), .groups = "drop")

 # Vectorized: compute expression stats for all genes at once
  # Convert gene columns to binary (expressing or not)
  expr_mat <- as.matrix(df[, genes, drop = FALSE]) > threshold

  # Sum per cell type using matrix multiplication with indicator
  ct_factor <- factor(df[[col_name]], levels = cell_types)
  ct_indicator <- model.matrix(~ ct_factor - 1)  # n_cells × n_celltypes indicator matrix
  colnames(ct_indicator) <- levels(ct_factor)

  # n_expressing = t(indicator) %*% expr_mat  → (n_celltypes × n_genes)
  n_expr_mat <- t(ct_indicator) %*% expr_mat

  # Convert to long format
  results <- expand.grid(cell_type = rownames(n_expr_mat), gene = colnames(n_expr_mat),
                          stringsAsFactors = FALSE)
  results$n_expressing <- as.vector(n_expr_mat)

  # Join n_total

  results <- results %>%
    left_join(n_total_df, by = "cell_type") %>%
    mutate(pct_expressing = n_expressing / n_total) %>%
    filter(n_total > 0) %>%
    select(gene, cell_type, pct_expressing, n_expressing, n_total)

  results
}

# Ligand metrics (threshold = 0 to match sender_bins)
ligand_metrics <- compute_filtering_metrics(obs_df, all_lr_ligands, SENDER_CELLTYPES, threshold = 0)
if (nrow(ligand_metrics) > 0) {
  ligand_metrics$passes_pct <- ligand_metrics$pct_expressing >= MIN_PCT_CELLTYPE
  ligand_metrics$passes_count <- ligand_metrics$n_expressing >= MIN_LIG_CELLS_CELLTYPE
  ligand_metrics$passes_filter <- ligand_metrics$passes_pct | ligand_metrics$passes_count
  ligand_metrics$lineage <- LINEAGE_MAP[ligand_metrics$cell_type]
}

# Receptor metrics (threshold = 1 to match receptor positivity)
receptor_metrics <- compute_filtering_metrics(obs_df, all_lr_receptors, RECEIVER_CELLTYPES, threshold = 1)
if (nrow(receptor_metrics) > 0) {
  receptor_metrics$passes_filter <- receptor_metrics$pct_expressing >= MIN_REC_FRAC_CELLTYPE
  receptor_metrics$lineage <- LINEAGE_MAP[receptor_metrics$cell_type]
}

# Save for diagnostics
write.csv(ligand_metrics, file.path(FIG_DIR, "ligand_celltype_filtering_metrics.csv"), row.names = FALSE)
write.csv(receptor_metrics, file.path(FIG_DIR, "receptor_celltype_filtering_metrics.csv"), row.names = FALSE)

cat("Ligand metrics:", nrow(ligand_metrics), "gene × cell_type combinations\n")
cat("  Passing filter:", sum(ligand_metrics$passes_filter), "\n")
cat("Receptor metrics:", nrow(receptor_metrics), "gene × cell_type combinations\n")
cat("  Passing filter:", sum(receptor_metrics$passes_filter), "\n")
```

## Cell type composition per DC1 bin

Compute fraction of each cell type (both `cell_annotation` and `cell_annotation_granular`) among all cells in each DC1 bin, then visualize as heatmaps. Code to make plot in figure 6D. 

```{r celltype_fractions_per_bin}
# Compute fractions for cell_annotation
ct_frac_coarse <- obs_df %>%
  filter(!is.na(DC1_bin30), !is.na(cell_annotation)) %>%
  group_by(DC1_bin30, cell_annotation) %>%
  summarize(n = n(), .groups = "drop") %>%
  group_by(DC1_bin30) %>%
  mutate(frac = n / sum(n)) %>%
  ungroup()

# Compute fractions for cell_annotation_granular
ct_frac_granular <- obs_df %>%
  filter(!is.na(DC1_bin30), !is.na(cell_annotation_granular)) %>%
  group_by(DC1_bin30, cell_annotation_granular) %>%
  summarize(n = n(), .groups = "drop") %>%
  group_by(DC1_bin30) %>%
  mutate(frac = n / sum(n)) %>%
  ungroup()

cat("Cell types (coarse):", length(unique(ct_frac_coarse$cell_annotation)), "\n")
cat("Cell types (granular):", length(unique(ct_frac_granular$cell_annotation_granular)), "\n")
```

```{r celltype_heatmap_coarse, fig.width=10, fig.height=8}
library(ComplexHeatmap)
library(circlize)

# exclude cell types that are less than 200 cells
cell_anno_counts = table(obs_df$cell_annotation) %>% sort()
filter_out_ct = names(cell_anno_counts)[cell_anno_counts < 200]
rm_ct = c("Mesenchymal-like_stroma", "Pericyte", "Mural" , "Smooth_muscle", 'Ductal', 'Tuft-like', 'Islet', 'ADM', 'Acinar', 'Endothelial_vascular', 'Endothelial_lymphatic')
filter_out_ct = unique(c(filter_out_ct, rm_ct))
# Pivot to matrix: rows = cell types, cols = DC1 bins
ct_mat_coarse <- ct_frac_coarse %>%
  dplyr::filter(!(cell_annotation %in% filter_out_ct)) %>%
  select(DC1_bin30, cell_annotation, frac) %>%
  pivot_wider(names_from = DC1_bin30, values_from = frac, values_fill = 0) %>%
  tibble::column_to_rownames("cell_annotation") %>%
  as.matrix()

# Order columns by bin number
ct_mat_coarse <- ct_mat_coarse[, order(as.numeric(colnames(ct_mat_coarse)))]

# Find position of bin 18 for vertical line
bin_cols <- as.numeric(colnames(ct_mat_coarse))
bin18_idx <- which(bin_cols == 18)

# Smooth rows: 3-bin rolling average (center + flanking neighbors)
smooth_rows <- function(mat) {
  n_cols <- ncol(mat)
  smoothed <- mat
  for (i in seq_len(nrow(mat))) {
    for (j in seq_len(n_cols)) {
      neighbors <- max(1, j-1):min(n_cols, j+1)
      smoothed[i, j] <- mean(mat[i, neighbors], na.rm = TRUE)
    }
  }
  smoothed
}
ct_mat_coarse_smooth <- smooth_rows(ct_mat_coarse)

# Z-score rows for visualization
zscore_rows <- function(mat) {
  t(apply(mat, 1, function(x) {
    if (sd(x, na.rm = TRUE) == 0) return(rep(0, length(x)))
    (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
  }))
}
ct_mat_coarse_z <- zscore_rows(ct_mat_coarse_smooth)

# Order rows by bin where they reach maximum value (using smoothed values)
row_order <- order(apply(ct_mat_coarse_smooth, 1, which.max))

col_fun <- colorRamp2(c(-2, 0, 2), c("#4575b4", "white", "#d73027"))

ht_coarse <- Heatmap(
  ct_mat_coarse_z[row_order, , drop = FALSE],
  col = col_fun,
  name = "Z-score (coarse)",
  column_title = "Cell type composition shifts along DC1 (cell_annotation)",
  row_title = "Cell type",
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  show_row_names = TRUE,
  row_names_side = "left",
  row_names_gp = gpar(fontsize = 9),
  show_column_names = FALSE,
  rect_gp = gpar(col = "lightgrey", lwd = 0.3),
  na_col = "grey90",
  column_title_gp = gpar(fontsize = 12, fontface = "bold")
)

draw(ht_coarse)
decorate_heatmap_body("Z-score (coarse)", {
  x_pos <- bin18_idx / ncol(ct_mat_coarse_z)
  grid.lines(x = unit(x_pos, "npc"), y = unit(c(0, 1), "npc"), gp = gpar(col = "black", lwd = 2))
})

# Save
svg(file.path(FIG_DIR, "celltype_composition_dc1_coarse.svg"), width = 6, height = 8)
draw(ht_coarse)
decorate_heatmap_body("Z-score (coarse)", {
  x_pos <- bin18_idx / ncol(ct_mat_coarse_z)
  grid.lines(x = unit(x_pos, "npc"), y = unit(c(0, 1), "npc"), gp = gpar(col = "black", lwd = 2))
})
dev.off()
cat("Saved coarse cell type heatmap\n")
```

```{r celltype_heatmap_coarse_filtered, fig.width=10, fig.height=8}
# make another one that filters out some non-relevant populations

# exclude cell types that are less than 200 cells
cell_anno_counts = table(obs_df$cell_annotation) %>% sort()
filter_out_ct = names(cell_anno_counts)[cell_anno_counts < 200]
rm_ct = c("Mesenchymal-like_stroma", "Pericyte", "Mural" , "Smooth_muscle", 
          'Ductal', 'Tuft-like', 'Islet', 'ADM', 'Acinar', 'Endothelial_vascular', 'Endothelial_lymphatic', 
          'PSC', 'Fibroblast', 'Lymphoid_proliferating', 'Mesothelial/apCAF')
filter_out_ct = unique(c(filter_out_ct, rm_ct))
# Pivot to matrix: rows = cell types, cols = DC1 bins
ct_mat_coarse <- ct_frac_coarse %>%
  dplyr::filter(!(cell_annotation %in% filter_out_ct)) %>%
  select(DC1_bin30, cell_annotation, frac) %>%
  pivot_wider(names_from = DC1_bin30, values_from = frac, values_fill = 0) %>%
  tibble::column_to_rownames("cell_annotation") %>%
  as.matrix()

# Order columns by bin number
ct_mat_coarse <- ct_mat_coarse[, order(as.numeric(colnames(ct_mat_coarse)))]

# Find position of bin 18 for vertical line
bin_cols <- as.numeric(colnames(ct_mat_coarse))
bin18_idx <- which(bin_cols == 18)

# Smooth rows: 3-bin rolling average (center + flanking neighbors)
smooth_rows <- function(mat) {
  n_cols <- ncol(mat)
  smoothed <- mat
  for (i in seq_len(nrow(mat))) {
    for (j in seq_len(n_cols)) {
      neighbors <- max(1, j-1):min(n_cols, j+1)
      smoothed[i, j] <- mean(mat[i, neighbors], na.rm = TRUE)
    }
  }
  smoothed
}
ct_mat_coarse_smooth <- smooth_rows(ct_mat_coarse)

# Z-score rows for visualization
zscore_rows <- function(mat) {
  t(apply(mat, 1, function(x) {
    if (sd(x, na.rm = TRUE) == 0) return(rep(0, length(x)))
    (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
  }))
}
ct_mat_coarse_z <- zscore_rows(ct_mat_coarse_smooth)

# Order rows by bin where they reach maximum value (using smoothed values)
row_order <- order(apply(ct_mat_coarse_smooth, 1, which.max))

col_fun <- colorRamp2(c(-2, 0, 2), c("#4575b4", "white", "#d73027"))

ht_coarse <- Heatmap(
  ct_mat_coarse_z[row_order, , drop = FALSE],
  col = col_fun,
  name = "Z-score (coarse)",
  column_title = "Cell type composition shifts along DC1",
  row_title = "Cell type",
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  show_row_names = TRUE,
  row_names_side = "left",
  row_names_gp = gpar(fontsize = 14),
  show_column_names = FALSE,
  #rect_gp = gpar(col = "lightgrey", lwd = 0.3),
  na_col = "grey90",
  column_title_gp = gpar(fontsize = 12, fontface = "bold")
)

draw(ht_coarse)
decorate_heatmap_body("Z-score (coarse)", {
  x_pos <- bin18_idx / ncol(ct_mat_coarse_z)
  grid.lines(x = unit(x_pos, "npc"), y = unit(c(0, 1), "npc"), gp = gpar(col = "black", lwd = 2))
})

# Save
svg(file.path(FIG_DIR, "celltype_composition_dc1_coarse_filtered.svg"), width = 4, height = 6)
draw(ht_coarse)
decorate_heatmap_body("Z-score (coarse)", {
  x_pos <- bin18_idx / ncol(ct_mat_coarse_z)
  grid.lines(x = unit(x_pos, "npc"), y = unit(c(0, 1), "npc"), gp = gpar(col = "black", lwd = 2))
})
dev.off()
cat("Saved coarse cell type heatmap\n")
```


```{r celltype_heatmap_granular, fig.width=12, fig.height=14}

# exclude cell types that are less than 200 cells
cell_anno_gran_counts = table(obs_df$cell_annotation_granular) %>% sort()
filter_out_ct = names(cell_anno_gran_counts)[cell_anno_gran_counts < 500]
rm_ct = names(cell_anno_gran_counts)[grepl("contam", names(cell_anno_gran_counts))]
#rm_ct = c("Mesenchymal-like_stroma", "Pericyte", "Mural" , "Smooth_muscle" )
filter_out_ct = unique(c(filter_out_ct, rm_ct))

# Pivot to matrix: rows = cell types, cols = DC1 bins
ct_mat_granular <- ct_frac_granular %>%
  dplyr::filter(!(cell_annotation_granular %in% filter_out_ct)) %>%
  select(DC1_bin30, cell_annotation_granular, frac) %>%
  pivot_wider(names_from = DC1_bin30, values_from = frac, values_fill = 0) %>%
  tibble::column_to_rownames("cell_annotation_granular") %>%
  as.matrix()

# Order columns by bin number
ct_mat_granular <- ct_mat_granular[, order(as.numeric(colnames(ct_mat_granular)))]

# Find position of bin 18 for vertical line
bin_cols_g <- as.numeric(colnames(ct_mat_granular))
bin18_idx_g <- which(bin_cols_g == 18)

# Z-score rows
ct_mat_granular_z <- zscore_rows(ct_mat_granular)

# Cluster rows
if (nrow(ct_mat_granular_z) > 1) {
  hc <- hclust(dist(ct_mat_granular_z), method = "ward.D2")
  row_order_g <- hc$order
} else {
  row_order_g <- 1
}

ht_granular <- Heatmap(
  ct_mat_granular_z[row_order_g, , drop = FALSE],
  col = col_fun,
  name = "Z-score (granular)",
  column_title = "Cell type composition shifts along DC1 (cell_annotation_granular)",
  row_title = "Cell type",
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  show_row_names = TRUE,
  row_names_side = "left",
  row_names_gp = gpar(fontsize = 8),
  show_column_names = FALSE,
  rect_gp = gpar(col = "lightgrey", lwd = 0.3),
  na_col = "grey90",
  column_title_gp = gpar(fontsize = 12, fontface = "bold")
)

draw(ht_granular)
decorate_heatmap_body("Z-score (granular)", {
  x_pos <- bin18_idx_g / ncol(ct_mat_granular_z)
  grid.lines(x = unit(x_pos, "npc"), y = unit(c(0, 1), "npc"), gp = gpar(col = "black", lwd = 2))
})

# Save
svg(file.path(FIG_DIR, "celltype_composition_dc1_granular.svg"), width = 12, height = 14)
draw(ht_granular)
decorate_heatmap_body("Z-score (granular)", {
  x_pos <- bin18_idx_g / ncol(ct_mat_granular_z)
  grid.lines(x = unit(x_pos, "npc"), y = unit(c(0, 1), "npc"), gp = gpar(col = "black", lwd = 2))
})
dev.off()
cat("Saved granular cell type heatmap\n")
```

```{r celltype_heatmap_granular_by_lineage, fig.width=10, fig.height=4, results='asis'}
# Separate heatmap per lineage with lineage-specific color scale

# Get lineage mapping from obs_df (cell_annotation_granular -> lineage)
ct_lineage_map <- obs_df %>%
  filter(!is.na(cell_annotation_granular), !is.na(lineage)) %>%
  distinct(cell_annotation_granular, lineage)

ct_lineage <- data.frame(
  cell_type = rownames(ct_mat_granular),
  stringsAsFactors = FALSE
) %>%
  left_join(ct_lineage_map, by = c("cell_type" = "cell_annotation_granular"))

ct_lineage$lineage[is.na(ct_lineage$lineage)] <- "Other"

# Order lineages
lineage_order <- c("Cancer", "Mac_Mono", "DC", "pDC", "Neutrophil",
                   "T_NK_ILC", "B_Plasma", "Fibroblast", "Stroma",
                   "Endothelial", "Epithelial", "Neural", "Other")

# Get lineages present in data
lineages_present <- intersect(lineage_order, unique(ct_lineage$lineage))

for (lin in lineages_present) {
  # Subset to this lineage
  ct_in_lin <- ct_lineage$cell_type[ct_lineage$lineage == lin]
  if (length(ct_in_lin) == 0) next

  mat_lin <- ct_mat_granular[ct_in_lin, , drop = FALSE]

  # Order rows alphabetically
  mat_lin <- mat_lin[order(rownames(mat_lin)), , drop = FALSE]

  # Lineage-specific color scale

  max_frac_lin <- max(mat_lin, na.rm = TRUE)
  if (max_frac_lin == 0) max_frac_lin <- 0.01
  col_fun_lin <- colorRamp2(c(0, max_frac_lin/2, max_frac_lin), c("white", "#fc8d59", "#b30000"))

  ht_lin <- Heatmap(
    mat_lin,
    col = col_fun_lin,
    name = "Fraction",
    column_title = paste0(lin, " — cell type fractions along DC1"),
    row_title = "Cell type",
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    show_row_names = TRUE,
    row_names_side = "left",
    row_names_gp = gpar(fontsize = 9),
    show_column_names = FALSE,
    rect_gp = gpar(col = "lightgrey", lwd = 0.3),
    na_col = "grey90",
    column_title_gp = gpar(fontsize = 11, fontface = "bold"),
    heatmap_legend_param = list(title = "Fraction")
  )

  # Display
  draw(ht_lin)
  decorate_heatmap_body("Fraction", {
    x_pos <- bin18_idx_g / ncol(mat_lin)
    grid.lines(x = unit(x_pos, "npc"), y = unit(c(0, 1), "npc"), gp = gpar(col = "black", lwd = 2))
  })

  # Save
  n_rows <- nrow(mat_lin)
  fig_height <- max(3, n_rows * 0.4 + 1.5)
  svg(file.path(FIG_DIR, paste0("celltype_composition_dc1_", lin, ".svg")), width = 10, height = fig_height)
  draw(ht_lin)
  decorate_heatmap_body("Fraction", {
    x_pos <- bin18_idx_g / ncol(mat_lin)
    grid.lines(x = unit(x_pos, "npc"), y = unit(c(0, 1), "npc"), gp = gpar(col = "black", lwd = 2))
  })
  dev.off()

  cat("\n\n")  # spacing between plots in output
}
cat("Saved separate heatmaps per lineage\n")
```

## Cell type fraction line plots

Line plots showing cell type fractions along DC1, colored by `cell_annotation_granular`. Code related to plots in S7E. 

```{r celltype_lineplot_setup}
# Prepare data for line plots
ct_frac_granular_plot <- ct_frac_granular %>%
  left_join(ct_lineage_map, by = c("cell_annotation_granular" = "cell_annotation_granular")) %>%
  mutate(lineage = ifelse(is.na(lineage), "Other", lineage))

# Bin centers for x-axis
ct_frac_granular_plot <- ct_frac_granular_plot %>%
  mutate(DC1_bin_center = (DC1_bin30 - 0.5) / DC1_N_BINS)

# Lineages to plot (add/remove as needed)
LINEAGES_TO_PLOT <- c("Mac_Mono", "Fibroblast")
```

```{r celltype_lineplots, fig.width=10, fig.height=5}
MIN_NICHE_FRACTION = 0.025
for (lin in LINEAGES_TO_PLOT) {
  plot_data <- ct_frac_granular_plot %>%
    filter(lineage == lin)
  keep_ct = plot_data %>%
    dplyr::group_by(cell_annotation_granular) %>%
    summarise(max_frac = max(frac)) %>%
    dplyr::ungroup() %>%
    dplyr::filter(max_frac > MIN_NICHE_FRACTION) %>% .$cell_annotation_granular
  plot_data = plot_data %>%
    dplyr::filter(cell_annotation_granular %in% keep_ct)

  if (nrow(plot_data) == 0) {
    cat("No data for lineage:", lin, "\n")
    next
  }

  p <- ggplot(plot_data, aes(x = DC1_bin_center, y = frac, color = cell_annotation_granular)) +
    geom_smooth(linewidth = 1, se = T, method = 'loess') +
    #geom_point(size = 1.5) +
    geom_vline(xintercept = 18 / DC1_N_BINS, linetype = "dashed", color = "black", linewidth = 0.8) +
    labs(
      x = "Classical niche ablation response (DC1)",
      y = "Fraction of cells",
      title = paste0(lin, " — cell type fractions along DC1"),
      color = "Cell type"
    ) +
    theme_tufte() +
    rose_theme +
    theme(
      legend.position = "right",
      legend.text = element_text(size = 9),
      plot.title = element_text(size = 12, face = "bold")
    ) +
    scale_color_tableau()

  print(p)

  ggsave(file.path(FIG_DIR, paste0("celltype_fraction_lineplot_", lin, ".svg")),
         p, width = 10, height = 5)
}
cat("Saved line plots\n")
```

```{r celltype_lineplots_within_lineage, fig.width=10, fig.height=5}
# Fraction of each granular annotation within its lineage (e.g., among all fibroblasts)
MIN_NICHE_LINEAGE_FRACTION = 0.1

ct_frac_within_lineage <- obs_df %>%
  filter(!is.na(DC1_bin30), !is.na(cell_annotation_granular), !is.na(lineage)) %>%
  group_by(DC1_bin30, lineage, cell_annotation_granular) %>%
  summarize(n = n(), .groups = "drop") %>%
  group_by(DC1_bin30, lineage) %>%
  mutate(frac_within = n / sum(n)) %>%
  ungroup() %>%
  mutate(DC1_bin_center = (DC1_bin30 - 0.5) / DC1_N_BINS)

for (lin in LINEAGES_TO_PLOT) {
  plot_data <- ct_frac_within_lineage %>%
    filter(lineage == lin)
  
  keep_ct = plot_data %>%
    dplyr::group_by(cell_annotation_granular) %>%
    summarise(max_frac = max(frac_within)) %>%
    dplyr::ungroup() %>%
    dplyr::filter(max_frac > MIN_NICHE_LINEAGE_FRACTION) %>% .$cell_annotation_granular
  plot_data = plot_data %>%
    dplyr::filter(cell_annotation_granular %in% keep_ct)
  
  # remove non-TAM populations
  plot_data = plot_data %>% dplyr::filter(!(grepl("Macrophage", cell_annotation_granular)))

  if (nrow(plot_data) == 0) {
    cat("No data for lineage:", lin, "\n")
    next
  }

  p <- ggplot(plot_data, aes(x = DC1_bin_center, y = frac_within, color = cell_annotation_granular)) +
    geom_smooth(linewidth = 1, se = TRUE, method = 'loess') +
    geom_vline(xintercept = 18 / DC1_N_BINS, linetype = "dashed", color = "black", linewidth = 0.8) +
    labs(
      x = "Classical niche ablation response (DC1)",
      y = paste0("Fraction within ", lin),
      title = paste0(lin, " — subtype fractions within lineage along DC1"),
      color = "Cell type"
    ) +
    theme_tufte() +
    rose_theme +
    theme(
      legend.position = "right",
      legend.text = element_text(size = 9),
      plot.title = element_text(size = 12, face = "bold")
    ) +
    scale_color_tableau()

  print(p)

  ggsave(file.path(FIG_DIR, paste0("celltype_fraction_within_lineage_lineplot_", lin, ".svg")),
         p, width = 10, height = 5)
}
cat("Saved within-lineage line plots\n")
```

---

# Linear Model Testing


## Tier1: Cell-type-specific ligand expression

```{r lm_ctmode}
# cell type mode: Sender-specific at cell_annotation level (more granular than Mode 1 lineage level)

# Build wide receptor fraction table for cell types
rec_ct_wide <- cov_bins
for (key in names(rec_celltype_bins)) {
  tmp <- rec_celltype_bins[[key]][, c("DC1_bin30", "rec_frac_bin")]
  colnames(tmp)[2] <- paste0("rec_ct_", key)
  rec_ct_wide <- left_join(rec_ct_wide, tmp, by = "DC1_bin30")
}

lm_rows_ctmode <- list()

for (lig_key in names(sender_celltype_bins)) {
  lig_df <- sender_celltype_bins[[lig_key]]
  parts  <- strsplit(lig_key, "__")[[1]]
  ct_s   <- parts[1]
  lig    <- parts[2]

  bin_wide <- inner_join(lig_df, rec_ct_wide, by = "DC1_bin30") %>%
    filter(!is.na(lig_expr_bin), n_sender >= 5)
  if (nrow(bin_wide) < 5) next

  lr_sub <- LR_PAIRS[LR_PAIRS$ligand == lig, ]

  rec_cols <- c()
  rec_meta <- list()
  for (j in seq_len(nrow(lr_sub))) {
    rec <- lr_sub$receptor_gene[j]
    for (ct_r in RECEIVER_CELLTYPES) {
      rc <- paste0("rec_ct_", rec, "__", ct_r)
      if (!rc %in% names(bin_wide)) next
      valid_rows <- !is.na(bin_wide[[rc]])
      if (sum(valid_rows) < 5) next
      rec_cols <- c(rec_cols, rc)
      rec_meta[[rc]] <- list(receptor = rec, receiver_celltype = ct_r,
                             pathway = lr_sub$pathway_name[j],
                             is_chemokine = lr_sub$is_chemokine[j],
                             is_chemokine_expanded = lr_sub$is_chemokine_expanded[j])
    }
  }
  if (length(rec_cols) == 0) next

  resp_mat <- as.matrix(bin_wide[, rec_cols, drop = FALSE])
  fit <- tryCatch(lm(resp_mat ~ lig_expr_bin, data = bin_wide), error = function(e) NULL)
  if (is.null(fit)) next

  fit_summ <- summary(fit)

  for (k in seq_along(rec_cols)) {
    rc <- rec_cols[k]
    meta <- rec_meta[[rc]]
    resp_summ <- if (length(rec_cols) == 1) fit_summ else fit_summ[[k]]
    if (is.null(resp_summ)) next
    s <- resp_summ$coefficients
    if (!"lig_expr_bin" %in% rownames(s)) next

    y_vals <- bin_wide[[rc]]
    valid_y <- !is.na(y_vals)
    pr <- if (sum(valid_y) >= 3) cor(bin_wide$lig_expr_bin[valid_y], y_vals[valid_y]) else NA_real_

    lm_rows_ctmode[[length(lm_rows_ctmode) + 1]] <- data.frame(
      mode              = "celltype",
      sender_celltype   = ct_s,
      sender_lineage    = as.character(LINEAGE_MAP[ct_s]),
      ligand            = lig,
      receiver_celltype = meta$receiver_celltype,
      receiver_lineage  = as.character(LINEAGE_MAP[meta$receiver_celltype]),
      receptor          = meta$receptor,
      pathway           = meta$pathway,
      is_chemokine      = meta$is_chemokine,
      is_chemokine_expanded = meta$is_chemokine_expanded,
      slope             = s["lig_expr_bin", "Estimate"],
      se                = s["lig_expr_bin", "Std. Error"],
      t_val             = s["lig_expr_bin", "t value"],
      pval              = s["lig_expr_bin", "Pr(>|t|)"],
      r2                = resp_summ$r.squared,
      pearson_r         = pr,
      n_bins            = sum(valid_y),
      stringsAsFactors  = FALSE
    )
  }
}

lm_df_ctmode <- if (length(lm_rows_ctmode) > 0) do.call(rbind, lm_rows_ctmode) else data.frame()
if (nrow(lm_df_ctmode) > 0) {
  lm_df_ctmode$padj <- p.adjust(lm_df_ctmode$pval, method = "BH")
  lm_df_ctmode      <- lm_df_ctmode[order(lm_df_ctmode$padj), ]
}
cat("cell type mode (cell type) pairs tested:", nrow(lm_df_ctmode), "\n")
cat("Significant (padj<0.1, slope>0):",
    sum(lm_df_ctmode$padj < 0.1 & lm_df_ctmode$slope > 0, na.rm = TRUE), "\n")
print(head(lm_df_ctmode[lm_df_ctmode$slope > 0, ], 10))
```

## Tier1: Niche-summed ligand and receptor

```{r lm_nichemode}
# niche mode: Both ligand AND receptor are niche-summed (cancer cell perspective)
# This tests whether cancer cells in niches with high ligand also have high receptor+ neighbors

# Build wide niche receptor table
niche_rec_wide <- cov_bins
for (key in names(niche_rec_bins)) {
  tmp <- niche_rec_bins[[key]][, c("DC1_bin30", "niche_rec_bin")]
  colnames(tmp)[2] <- paste0("niche_rec_", key)
  niche_rec_wide <- left_join(niche_rec_wide, tmp, by = "DC1_bin30")
}

lm_rows_nichemode <- list()

# Loop over unique ligands
for (lig in names(niche_lig_bins)) {
  lig_df <- niche_lig_bins[[lig]]

  # Join with niche receptor wide table
  bin_wide <- inner_join(lig_df, niche_rec_wide[, !names(niche_rec_wide) %in%
                        # c("frac_basal_bin", "frac_mesen_bin", "n_cancer")],
                          c( "n_cancer")],
                         by = "DC1_bin30")

  # Filter valid covariate rows
  valid_cov <- !is.na(bin_wide$niche_lig_bin) #&
               #!is.na(bin_wide$frac_basal_bin) &
               #!is.na(bin_wide$frac_mesen_bin)
  bin_sub <- bin_wide[valid_cov, ]
  if (nrow(bin_sub) < 5) next

  # Add standardized ligand column
 bin_sub$lig_expr <- bin_sub$niche_lig_bin

  # Find all LR pairs for this ligand
  lr_sub <- LR_PAIRS[LR_PAIRS$ligand == lig, ]

  # Gather valid niche receptor columns (no lineage - summed across all cell types)
  rec_cols <- c()
  rec_meta <- list()
  for (j in seq_len(nrow(lr_sub))) {
    rec <- lr_sub$receptor_gene[j]
    rc <- paste0("niche_rec_", rec)
    if (!rc %in% names(bin_sub)) next
    valid_rows <- !is.na(bin_sub[[rc]])
    if (sum(valid_rows) < 5) next
    rec_cols <- c(rec_cols, rc)
    rec_meta[[rc]] <- list(receptor = rec,
                           pathway = lr_sub$pathway_name[j],
                           is_chemokine = lr_sub$is_chemokine[j],
                             is_chemokine_expanded = lr_sub$is_chemokine_expanded[j])
  }
  if (length(rec_cols) == 0) next

  # Fit multivariate lm
  resp_mat <- as.matrix(bin_sub[, rec_cols, drop = FALSE])
  fit <- tryCatch(
    #lm(resp_mat ~ lig_expr + frac_basal_bin + frac_mesen_bin, data = bin_sub),
    lm(resp_mat ~ lig_expr, data = bin_sub),
    error = function(e) NULL
  )
  if (is.null(fit)) next

  fit_summ <- summary(fit)

  for (k in seq_along(rec_cols)) {
    rc <- rec_cols[k]
    meta <- rec_meta[[rc]]

    resp_summ <- if (length(rec_cols) == 1) fit_summ else fit_summ[[k]]
    if (is.null(resp_summ)) next

    s <- resp_summ$coefficients
    if (!"lig_expr" %in% rownames(s)) next

    y_vals <- bin_sub[[rc]]
    valid_y <- !is.na(y_vals)
    pr <- if (sum(valid_y) >= 3) cor(bin_sub$lig_expr[valid_y], y_vals[valid_y]) else NA_real_

    lm_rows_nichemode[[length(lm_rows_nichemode) + 1]] <- data.frame(
      mode             = "niche_both",
      sender_lineage   = "all",
      ligand           = lig,
      receiver_lineage = "all",
      receptor         = meta$receptor,
      pathway          = meta$pathway,
      is_chemokine     = meta$is_chemokine,
      is_chemokine_expanded = meta$is_chemokine_expanded,
      slope            = s["lig_expr", "Estimate"],
      se               = s["lig_expr", "Std. Error"],
      t_val            = s["lig_expr", "t value"],
      pval             = s["lig_expr", "Pr(>|t|)"],
      r2               = resp_summ$r.squared,
      pearson_r        = pr,
      n_bins           = sum(valid_y),
      stringsAsFactors = FALSE
    )
  }
}

lm_df_nichemode <- if (length(lm_rows_nichemode) > 0) do.call(rbind, lm_rows_nichemode) else data.frame()
if (nrow(lm_df_nichemode) > 0) {
  lm_df_nichemode$padj <- p.adjust(lm_df_nichemode$pval, method = "BH")
  lm_df_nichemode      <- lm_df_nichemode[order(lm_df_nichemode$padj), ]
}
cat("niche mode pairs tested:", nrow(lm_df_nichemode), "\n")
cat("Significant (padj<0.1, slope>0):",
    sum(lm_df_nichemode$padj < 0.1 & lm_df_nichemode$slope > 0, na.rm = TRUE), "\n")
print(head(lm_df_nichemode[lm_df_nichemode$slope > 0, ], 10))
```

---

# Line Plots: Concordant Mode1/nichemode Chemokine Associations

For each significant nichemode chemokine receptor, plot trends for:
1. Receptor+ cell fraction for receiver lineages with significant Mode1 associations (same direction)
2. Ligand expression for sender lineages with significant Mode1 associations (faceted by ligand)

```{r filter_concordant_associations}
# Get significant nichemode chemokine receptor associations (padj < 0.05)
sig_nichemode_chemo <- lm_df_nichemode %>%
  filter(padj < 0.05, is_chemokine) %>%
  select(receptor, ligand, slope_nichemode = slope)

cat("Significant nichemode chemokine associations:", nrow(sig_nichemode_chemo), "\n")
print(sig_nichemode_chemo)

# Filter Mode1 for pval < 0.05 AND same direction as nichemode
concordant_mode1 <- lm_df_mode1 %>%
  filter(padj < 0.1, is_chemokine) %>%
  inner_join(sig_nichemode_chemo, by = c("receptor", "ligand")) %>%
  filter(sign(slope) == sign(slope_nichemode))

cat("\nConcordant Mode1 associations (pval < 0.05, same direction):", nrow(concordant_mode1), "\n")
print(concordant_mode1 %>% select(receptor, ligand, sender_lineage, receiver_lineage, slope, pval))
```

```{r build_bin_data_for_plots}
# Build bin-level data for receptor fraction and ligand expression
# Using existing rec_bins and sender_bins from earlier in notebook
# Note: rec_frac_bin = receptor+ cells of lineage / total niche cells (all lineages)

# Receptor fraction data: extract from rec_bins
rec_plot_df <- do.call(rbind, lapply(names(rec_bins), function(key) {
  parts <- strsplit(key, "__")[[1]]
  rec <- parts[1]
  lin <- parts[2]
  df <- rec_bins[[key]]
  df$receptor <- rec
  df$receiver_lineage <- lin
  df
}))

# Ligand expression data: extract from sender_bins (lig_expr_bin = mean log1p norm)
lig_plot_df <- do.call(rbind, lapply(names(sender_bins), function(key) {
  parts <- strsplit(key, "__")[[1]]
  lin <- parts[1]
  lig <- parts[2]
  df <- sender_bins[[key]]
  df$ligand <- lig
  df$sender_lineage <- lin
  df
}))

cat("Receptor bin data rows:", nrow(rec_plot_df), "\n")
cat("Ligand bin data rows:", nrow(lig_plot_df), "\n")
```

```{r plot_receptor_ligand_loess_function}
# Main plotting function using loess smoothing on bin-level data
# Stacks receptor plot on top, then one ligand plot per row below
plot_receptor_ligand_loess <- function(receptor, concordant_df, rec_plot_df, lig_plot_df,
                                       cov_bins) {

  # Get receiver lineages for this receptor
  receiver_lineages <- concordant_df %>%
    filter(receptor == !!receptor) %>%
    pull(receiver_lineage) %>%
    unique()

  # Get sender-ligand pairs for this receptor
  sender_df <- concordant_df %>%
    filter(receptor == !!receptor) %>%
    select(ligand, sender_lineage) %>%
    distinct()

  # Filter receptor data for this receptor and significant lineages
  rec_df <- rec_plot_df %>%
    filter(receptor == !!receptor, receiver_lineage %in% receiver_lineages)

  if (nrow(rec_df) == 0) {
    message("No receptor data for ", receptor)
    return(NULL)
  }

  # Scale receptor fraction per lineage (min-max to 0-1)
  rec_df <- rec_df %>%
    group_by(receiver_lineage) %>%
    mutate(rec_frac_scaled = (rec_frac_bin - min(rec_frac_bin, na.rm = TRUE)) /
                              (max(rec_frac_bin, na.rm = TRUE) - min(rec_frac_bin, na.rm = TRUE))) %>%
    ungroup()

  # Filter ligand data for significant sender-ligand pairs
  lig_df <- lig_plot_df %>%
    inner_join(sender_df, by = c("ligand", "sender_lineage"))

  ligands <- unique(lig_df$ligand)
  n_ligands <- length(ligands)
  has_ligands <- n_ligands > 0

  p_receptor <- ggplot(rec_df, aes(x = DC1_bin30, y = rec_frac_scaled,
                                   color = receiver_lineage, fill = receiver_lineage)) +
    geom_point(alpha = 0.6, size = 2) +
    geom_smooth(method = "loess", se = TRUE, alpha = 0.2, linewidth = 1) +
    labs(title = paste0(receptor, "+ cells in niche"),
         x = if (!has_ligands) "Classical niche ablation response (DC1 bin)" else NULL,
         y = "Scaled fraction (min-max)",
         color = "Receiver", fill = "Receiver") +
    theme_minimal(base_size = 14) +
    theme(legend.position = "right",
          axis.text.x = if (has_ligands) element_blank() else element_text()) +
    rose_theme

  if (!has_ligands) {
    message("No ligand data for ", receptor)
    return(p_receptor)
  }

  # Create separate plot for each ligand
  ligand_plots <- lapply(seq_along(ligands), function(i) {
    lig <- ligands[i]
    lig_sub <- lig_df %>% filter(ligand == lig)
    is_last <- (i == n_ligands)

    p <- ggplot(lig_sub, aes(x = DC1_bin30, y = lig_expr_bin,
                             color = sender_lineage, fill = sender_lineage)) +
      geom_point(alpha = 0.6, size = 2) +
      geom_smooth(method = "loess", se = TRUE, alpha = 0.2, linewidth = 1) +
      labs(title = lig,
           x = if (is_last) "Classical niche ablation response (DC1 bin)" else NULL,
           y = "Mean log1p norm",
           color = "Sender", fill = "Sender") +
      theme_minimal(base_size = 14) +
      theme(legend.position = "right",
            axis.text.x = if (is_last) element_text() else element_blank()) +
      rose_theme
    p
  })

  # Stack all plots vertically
  all_plots <- c(list(p_receptor), ligand_plots)
  wrap_plots(all_plots, ncol = 1)
}
```

### global level

Line plots showing global niche sums of ligand and receptor positive cell frequency along DC1 for specific chemokine receptors.

```{r global_lr_freq_lineplot_function}
plot_global_lr_freq_lines <- function(receptor_name, lr_pairs, niche_lig_bins, niche_rec_bins,
                                       dc1_n_bins, fig_dir, rm_ligands = NULL, rescale_data = FALSE) {
  # Get ligands for this receptor from LR_PAIRS
  # Use base R subsetting to avoid dplyr NSE issues with column name 'receptor'
  ligands <- lr_pairs$ligand[lr_pairs$receptor_gene == receptor_name]
  ligands <- unique(as.character(ligands))
  
  if(!(is.null(rm_ligands))){
    ligands = ligands[!(ligands %in% rm_ligands)]
  }

  # Filter to ligands that exist in niche_lig_bins
  ligands <- ligands[ligands %in% names(niche_lig_bins)]

  cat("Ligands for", receptor_name, ":", paste(ligands, collapse = ", "), "\n")

  if (length(ligands) == 0) {
    cat("No ligands found in niche_lig_bins for receptor:", receptor_name, "\n")
    return(NULL)
  }

  # Check receptor exists in niche_rec_bins
  if (!receptor_name %in% names(niche_rec_bins)) {
    cat("Receptor not found in niche_rec_bins:", receptor_name, "\n")
    return(NULL)
  }

  # Build combined data frame
  plot_data_list <- list()

  # Add ligand data
  for (lig in ligands) {
    lig_df <- niche_lig_bins[[lig]]
    if (is.null(lig_df) || nrow(lig_df) == 0) next
    plot_data_list[[lig]] <- data.frame(
      DC1_bin30 = lig_df$DC1_bin30,
      DC1_bin_center = (lig_df$DC1_bin30 - 0.5) / dc1_n_bins,
      freq = lig_df$niche_lig_bin,
      gene = lig,
      type = "ligand",
      stringsAsFactors = FALSE
    )
  }

  # Add receptor data
  rec_df <- niche_rec_bins[[receptor_name]]
  if (!is.null(rec_df) && nrow(rec_df) > 0) {
    plot_data_list[[receptor_name]] <- data.frame(
      DC1_bin30 = rec_df$DC1_bin30,
      DC1_bin_center = (rec_df$DC1_bin30 - 0.5) / dc1_n_bins,
      freq = rec_df$niche_rec_bin,
      gene = receptor_name,
      type = "receptor",
      stringsAsFactors = FALSE
    )
  }

  plot_data <- do.call(rbind, plot_data_list)
  
  # rescale if necessary
  if(rescale_data){
    plot_data = plot_data %>%
      dplyr::group_by(gene) %>%
      dplyr::mutate(freq = scales::rescale(freq)) %>%
      ungroup()
  }

  if (nrow(plot_data) == 0) {
    cat("No data to plot for receptor:", receptor_name, "\n")
    return(NULL)
  }

  # Create color palette - distinct colors for each gene
  all_genes <- unique(plot_data$gene)
  n_colors <- length(all_genes)
  if (n_colors <= 8) {
    color_pal <- RColorBrewer::brewer.pal(max(3, n_colors), "Set2")[1:n_colors]
  } else {
    color_pal <- colorRampPalette(RColorBrewer::brewer.pal(8, "Set2"))(n_colors)
  }
  names(color_pal) <- all_genes

  # Plot
  p <- ggplot(plot_data, aes(x = DC1_bin_center, y = freq, color = gene,fill = gene, linetype = type)) +
    geom_smooth(method = "loess", se = T, linewidth = 1.2, span = 0.3, alpha = .2) +
    geom_vline(xintercept = 18 / dc1_n_bins, linetype = "dotted", color = "grey40", linewidth = 0.8) +
    scale_linetype_manual(values = c("ligand" = "solid", "receptor" = "dashed")) +
    scale_color_manual(values = color_pal) +
    scale_fill_manual(values = color_pal) +
    labs(
      x = X_LAB,
      y = "Scaled amount",
      title = paste0(receptor_name, " and its ligands along DC1"),
      color = "Gene",
      linetype = "Type"
    ) +
    theme_tufte() +
    rose_theme +
    theme(
      legend.position = "right",
      plot.title = element_text(size = 12, face = "bold")
    )

  #print(p)

  # Save
  save_path <- file.path(fig_dir, paste0("global_lr_freq_", receptor_name, ".svg"))
  ggsave(save_path, p, width = 8, height = 5)
  cat("Saved:", save_path, "\n")

  return(p)
}
```

```{r global_lr_freq_lineplots, fig.width=8, fig.height=5}
# Plot Cxcr3 and its ligands (Cxcl9, Cxcl10)
plot_global_lr_freq_lines("Cxcr3", LR_PAIRS, niche_lig_bins, niche_rec_bins, DC1_N_BINS, FIG_DIR, rm_ligands = c('Pf4', 'Cxcl13'), rescale_data = T)

# Plot Cxcr6 and its ligands (Cxcl16)
plot_global_lr_freq_lines("Cxcr6", LR_PAIRS, niche_lig_bins, niche_rec_bins, DC1_N_BINS, FIG_DIR, rescale_data = T)

# Plot Cxcr6 and its ligands (Cxcl16)
plot_global_lr_freq_lines("Cxcr4", LR_PAIRS, niche_lig_bins, niche_rec_bins, DC1_N_BINS, FIG_DIR, rescale_data = T)

```


# Cytokine Ligand–Response Program Analysis

Test associations between cytokine ligand expression and response gene program activation across the DC1 trajectory.

# Ligand-Program Response Association Analysis

Test individual ligand expression (from sender_bins) against program response scores across DC1 bins.

## Compute response gene expression bins

For response genes in cytokine_signatures not already in sender_bins, compute pseudobulk expression per lineage per bin.

```{r compute_response_gene_bins}
all_response_genes <- unique(unlist(cytokine_signatures))
cat("Total response genes across programs:", length(all_response_genes), "\n")

response_gene_bins <- list()
genes_computed <- 0

for (lin in SENDER_LINEAGES) {
  sub <- obs_df[!is.na(obs_df$lineage) & obs_df$lineage == lin & !is.na(obs_df$DC1_bin30), ]
  if (nrow(sub) < 10) next

  for (gene in all_response_genes) {
    existing_key <- paste(lin, gene, sep = "__")

    # Skip if already in sender_bins
    if (existing_key %in% names(sender_bins)) {
      response_gene_bins[[existing_key]] <- sender_bins[[existing_key]]
      next
    }

    # Check if gene exists in data
    if (!gene %in% names(sub)) next

    # Apply same filtering as ligands (OR logic: pct >= MIN_PCT OR n_cells >= MIN_LIG_CELLS)
    gene_pct <- mean(sub[[gene]] > 0, na.rm = TRUE)
    gene_n_cells <- sum(sub[[gene]] > 0, na.rm = TRUE)
    if (gene_pct < MIN_PCT && gene_n_cells < MIN_LIG_CELLS) next

    tmp <- sub %>%
      group_by(DC1_bin30) %>%
      summarize(
        lig_expr_bin = log1p(sum(.data[[gene]], na.rm = TRUE) /
                             pmax(sum(lib_size, na.rm = TRUE), 1)),
        n_sender = n(),
        n_lig_cells = sum(.data[[gene]] > 0, na.rm = TRUE),
        .groups = "drop"
      )

    response_gene_bins[[existing_key]] <- tmp
    genes_computed <- genes_computed + 1
  }
}

cat("Response gene bins computed:", genes_computed, "new +",
    length(response_gene_bins) - genes_computed, "from sender_bins\n")
```

## Compute program response scores per lineage per bin

Mean Z-scaled expression of response genes within each program.

```{r compute_program_response_scores}
program_response_scores <- list()
bin_ids <- sort(unique(cov_bins$DC1_bin30))

for (prog in names(cytokine_signatures)) {
  response_genes <- cytokine_signatures[[prog]]

  for (lin in SENDER_LINEAGES) {
    # Find which response genes have bin data for this lineage
    available_genes <- character(0)
    for (gene in response_genes) {
      key <- paste(lin, gene, sep = "__")
      if (key %in% names(response_gene_bins)) {
        available_genes <- c(available_genes, gene)
      }
    }

    if (length(available_genes) < 1) next  # Need at least 1 gene

    # Build matrix: rows = genes, cols = bins
    gene_mat <- matrix(NA, nrow = length(available_genes), ncol = length(bin_ids))
    rownames(gene_mat) <- available_genes

    for (i in seq_along(available_genes)) {
      gene <- available_genes[i]
      key <- paste(lin, gene, sep = "__")
      gene_df <- response_gene_bins[[key]]
      for (j in seq_along(bin_ids)) {
        idx <- which(gene_df$DC1_bin30 == bin_ids[j])
        if (length(idx) > 0) gene_mat[i, j] <- gene_df$lig_expr_bin[idx]
      }
    }

    # Z-scale each gene across bins (within lineage)
    z_mat <- t(apply(gene_mat, 1, function(x) {
      if (sd(x, na.rm = TRUE) == 0) return(rep(0, length(x)))
      (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
    }))

    # Average Z-scores to get program response score per bin
    response_score <- colMeans(z_mat, na.rm = TRUE)

    result_key <- paste(lin, prog, sep = "__")
    program_response_scores[[result_key]] <- data.frame(
      DC1_bin30 = bin_ids,
      response_score = response_score,
      n_genes = length(available_genes)
    )
  }
}

cat("Program response scores computed:", length(program_response_scores), "lineage-program pairs\n")
```

## Linear model: ligand expression → program response score

Test all panel_cyto_genes ligands against all signal_anno_broad programs across sender × receiver lineage combinations.

```{r ligand_program_lm}
lm_rows_lig_prog <- list()

for (lig in panel_cyto_genes) {
  for (sender_lin in SENDER_LINEAGES) {
    lig_key <- paste(sender_lin, lig, sep = "__")
    if (!lig_key %in% names(sender_bins)) next

    lig_df <- sender_bins[[lig_key]]

    # Test against all programs × receiver lineages
    for (prog in names(cytokine_signatures)) {
      for (receiver_lin in RECEIVER_LINEAGES) {
        prog_key <- paste(receiver_lin, prog, sep = "__")
        if (!prog_key %in% names(program_response_scores)) next

        prog_df <- program_response_scores[[prog_key]]

        # Merge ligand expression with program response score
        merged <- inner_join(lig_df, prog_df, by = "DC1_bin30")
        merged <- merged[!is.na(merged$lig_expr_bin) & !is.na(merged$response_score) &
                         merged$n_sender >= 5, ]
        if (nrow(merged) < 5) next

        # Z-scale lig_expr_bin across bins
        if (sd(merged$lig_expr_bin, na.rm = TRUE) > 0) {
          merged$lig_expr_z <- (merged$lig_expr_bin - mean(merged$lig_expr_bin, na.rm = TRUE)) /
                                sd(merged$lig_expr_bin, na.rm = TRUE)
        } else {
          next
        }

        # Fit lm: response_score ~ lig_expr_z (both Z-scaled)
        fit <- tryCatch(
          lm(response_score ~ lig_expr_z, data = merged),
          error = function(e) NULL
        )
        if (is.null(fit)) next

        s <- summary(fit)$coefficients
        if (!"lig_expr_z" %in% rownames(s)) next

        pr <- cor(merged$response_score, merged$lig_expr_z, use = "complete.obs")

        lm_rows_lig_prog[[length(lm_rows_lig_prog) + 1]] <- data.frame(
          ligand = lig,
          sender_lineage = sender_lin,
          receiver_lineage = receiver_lin,
          program = prog,
          slope = s["lig_expr_z", "Estimate"],
          se = s["lig_expr_z", "Std. Error"],
          t_val = s["lig_expr_z", "t value"],
          pval = s["lig_expr_z", "Pr(>|t|)"],
          r2 = summary(fit)$r.squared,
          pearson_r = pr,
          n_bins = nrow(merged),
          n_response_genes = prog_df$n_genes[1],
          stringsAsFactors = FALSE
        )
      }
    }
  }
}

lm_df_lig_prog <- if (length(lm_rows_lig_prog) > 0) do.call(rbind, lm_rows_lig_prog) else data.frame()
if (nrow(lm_df_lig_prog) > 0) {
  lm_df_lig_prog$padj <- p.adjust(lm_df_lig_prog$pval, method = "BH")
  lm_df_lig_prog <- lm_df_lig_prog[order(lm_df_lig_prog$padj), ]
}

cat("Ligand-program pairs tested:", nrow(lm_df_lig_prog), "\n")
cat("Significant (padj<0.05, slope>0):", sum(lm_df_lig_prog$padj < 0.05 & lm_df_lig_prog$slope > 0, na.rm = TRUE), "\n")
cat("Significant (padj<0.05, slope<0):", sum(lm_df_lig_prog$padj < 0.05 & lm_df_lig_prog$slope < 0, na.rm = TRUE), "\n")
print(head(lm_df_lig_prog[lm_df_lig_prog$slope > 0, ], 15))

write.csv(lm_df_lig_prog, file.path(FIG_DIR, "ligand_program_lm_results.csv"), row.names = FALSE)
```

```{r}
# read in if necessary
lm_df_lig_prog <- readr::read_csv(file.path(FIG_DIR, "ligand_program_lm_results.csv"))
```

```{r}
# save computed program scores
saveRDS(program_response_scores, file = file.path(PRECOMPUTED_DIR, 'program_response_scores_bin.rds'))
```


## response program correlation

Compute Spearman correlation between program response scores and DC1 bin number to identify programs that increase or decrease along the trajectory.

```{r program_dc1_correlation}
program_dc1_cor <- list()

for (key in names(program_response_scores)) {
  prog_df <- program_response_scores[[key]]

  # Skip if too few bins
  if (nrow(prog_df) < 5) next

  # Spearman correlation: response_score ~ DC1_bin30
  cor_test <- cor.test(prog_df$DC1_bin30, prog_df$response_score, method = "spearman", exact = FALSE)

  # Parse lineage and program from key
  parts <- strsplit(key, "__")[[1]]
  lineage <- parts[1]
  program <- parts[2]

  program_dc1_cor[[length(program_dc1_cor) + 1]] <- data.frame(
    lineage = lineage,
    program = program,
    spearman_rho = cor_test$estimate,
    pval = cor_test$p.value,
    n_bins = nrow(prog_df),
    n_genes = prog_df$n_genes[1]
  )
}

program_dc1_cor_df <- do.call(rbind, program_dc1_cor)
rownames(program_dc1_cor_df) <- NULL

# FDR correction
program_dc1_cor_df$padj <- p.adjust(program_dc1_cor_df$pval, method = "BH")

# Sort by rho
program_dc1_cor_df <- program_dc1_cor_df %>% arrange(desc(spearman_rho))

cat("Program-DC1 correlations computed:", nrow(program_dc1_cor_df), "\n")
cat("  Significant (padj < 0.05):", sum(program_dc1_cor_df$padj < 0.05), "\n")
cat("  Positive (rho > 0):", sum(program_dc1_cor_df$spearman_rho > 0 & program_dc1_cor_df$padj < 0.05), "\n")
cat("  Negative (rho < 0):", sum(program_dc1_cor_df$spearman_rho < 0 & program_dc1_cor_df$padj < 0.05), "\n")

# Save
write.csv(program_dc1_cor_df, file.path(FIG_DIR, "program_dc1_spearman_correlations.csv"), row.names = FALSE)
cat("Saved to:", file.path(FIG_DIR, "program_dc1_spearman_correlations.csv"), "\n")
```

```{r program_dc1_cor_table}
# Show top positively and negatively correlated programs
cat("\nTop 10 positively correlated programs:\n")
print(head(program_dc1_cor_df %>% filter(spearman_rho > 0), 10))

cat("\nTop 10 negatively correlated programs:\n")
print(head(program_dc1_cor_df %>% filter(spearman_rho < 0) %>% arrange(spearman_rho), 10))
```

## Heatmap visualization: ligand → receptor → program response

```{r plot_program_heatmap_lineage_fn}
plot_program_heatmap_lineage <- function(sig_df, sender_bins, rec_bins, program_response_scores,
                                          cov_bins, LR_PAIRS, DC1_N_BINS,
                                          title_suffix, file_suffix, FIG_DIR,
                                          normalize = "zscore", program_order = NULL,
                                          include_receptors = FALSE) {
  if (nrow(sig_df) == 0) {
    cat("No pairs for", title_suffix, "— skipping.\n")
    return(NULL)
  }

  bin_ids <- sort(unique(cov_bins$DC1_bin30))
  n_bins <- length(bin_ids)
  bin_centres <- (bin_ids - 0.5) / DC1_N_BINS

  programs <- if (!is.null(program_order)) {
    intersect(program_order, unique(sig_df$program))
  } else {
    unique(sig_df$program)
  }

  # Build row_info: gene_type = "ligand", "receptor", or "program"
  row_info <- data.frame(
    gene = character(),
    lineage = character(),
    program_group = character(),
    gene_type = character(),
    bin_key = character(),
    stringsAsFactors = FALSE
  )

  for (prog in programs) {
    prog_df <- sig_df[sig_df$program == prog, ]

    # Get unique ligand-sender-receiver combinations for this program
    lig_lin <- unique(prog_df[, c("ligand", "sender_lineage", "receiver_lineage")])

    for (i in seq_len(nrow(lig_lin))) {
      lig <- lig_lin$ligand[i]
      sender_lin <- lig_lin$sender_lineage[i]
      receiver_lin <- lig_lin$receiver_lineage[i]

      # Add ligand row (sender lineage)
      bin_key <- paste(sender_lin, lig, sep = "__")
      if (bin_key %in% names(sender_bins)) {
        row_info <- rbind(row_info, data.frame(
          gene = lig, lineage = sender_lin, program_group = prog,
          gene_type = "ligand", bin_key = bin_key, stringsAsFactors = FALSE
        ))
      }

      # Add receptor rows if requested (receiver lineage)
      if (include_receptors) {
        # Find receptors for this ligand from LR_PAIRS
        lig_receptors <- unique(LR_PAIRS$receptor_gene[LR_PAIRS$ligand == lig])
        for (rec in lig_receptors) {
          rec_key <- paste(rec, receiver_lin, sep = "__")
          if (rec_key %in% names(rec_bins)) {
            row_info <- rbind(row_info, data.frame(
              gene = rec, lineage = receiver_lin, program_group = prog,
              gene_type = "receptor", bin_key = rec_key, stringsAsFactors = FALSE
            ))
          }
        }
      }
    }

    # Add program response score row (receiver lineages)
    prog_receiver_lineages <- unique(prog_df$receiver_lineage)
    for (lin in prog_receiver_lineages) {
      prog_key <- paste(lin, prog, sep = "__")
      if (prog_key %in% names(program_response_scores)) {
        row_info <- rbind(row_info, data.frame(
          gene = prog, lineage = lin, program_group = prog,
          gene_type = "program", bin_key = prog_key, stringsAsFactors = FALSE
        ))
      }
    }
  }

  # Remove duplicates
  row_info <- row_info[!duplicated(paste(row_info$gene, row_info$lineage, row_info$program_group, row_info$gene_type)), ]

  if (nrow(row_info) == 0) {
    cat("No valid rows for", title_suffix, "— skipping.\n")
    return(NULL)
  }

  # Build matrix
  mat <- matrix(NA, nrow = nrow(row_info), ncol = n_bins)
  row_labels <- paste0(row_info$gene, " (", row_info$lineage, ")")
  rownames(mat) <- row_labels
  colnames(mat) <- sprintf("%.2f", bin_centres)

  for (i in seq_len(nrow(row_info))) {
    gtype <- row_info$gene_type[i]
    bin_key <- row_info$bin_key[i]

    if (gtype == "ligand" && bin_key %in% names(sender_bins)) {
      bin_df <- sender_bins[[bin_key]]
      val_col <- "lig_expr_bin"
    } else if (gtype == "receptor" && bin_key %in% names(rec_bins)) {
      bin_df <- rec_bins[[bin_key]]
      val_col <- "rec_frac_bin"
    } else if (gtype == "program" && bin_key %in% names(program_response_scores)) {
      bin_df <- program_response_scores[[bin_key]]
      val_col <- "response_score"
    } else {
      next
    }

    if (!is.null(bin_df) && val_col %in% names(bin_df)) {
      for (j in seq_along(bin_ids)) {
        idx <- which(bin_df$DC1_bin30 == bin_ids[j])
        if (length(idx) > 0) mat[i, j] <- bin_df[[val_col]][idx]
      }
    }
  }

  # Smooth rows: 3-bin rolling average
  smooth_rows <- function(m) {
    n_cols <- ncol(m)
    smoothed <- m
    for (i in seq_len(nrow(m))) {
      for (j in seq_len(n_cols)) {
        neighbors <- max(1, j-1):min(n_cols, j+1)
        smoothed[i, j] <- mean(m[i, neighbors], na.rm = TRUE)
      }
    }
    smoothed
  }
  mat <- smooth_rows(mat)

  # Normalize rows
  if (normalize == "minmax") {
    norm_mat <- t(apply(mat, 1, function(x) {
      rng <- range(x, na.rm = TRUE)
      if (rng[2] == rng[1]) return(rep(0.5, length(x)))
      (x - rng[1]) / (rng[2] - rng[1])
    }))
    legend_name <- "Min-max"
    col_fun <- colorRamp2(c(0, 0.5, 1), c("#f7f7f7", "#fc8d59", "#b30000"))
  } else {
    norm_mat <- t(apply(mat, 1, function(x) {
      if (sd(x, na.rm = TRUE) == 0) return(rep(0, length(x)))
      (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
    }))
    legend_name <- "Z-score"
    col_fun <- colorRamp2(c(-2, 0, 2), c("#4575b4", "white", "#d73027"))
  }
  colnames(norm_mat) <- colnames(mat)

  # Order rows: ligands first, then receptors, then programs; within each by onset
  get_onset_bin <- function(x) {
    if (all(is.na(x)) || length(x) < 2) return(Inf)
    deriv <- diff(x)
    onset <- which(deriv > 0.05 * max(abs(deriv), na.rm = TRUE))[1]
    if (is.na(onset)) return(Inf)
    onset
  }
  row_info$onset_bin <- apply(norm_mat, 1, get_onset_bin)

  # Order: ligands first, receptors second, programs last; then by onset
  row_info$gene_type_order <- match(row_info$gene_type, c("ligand", "receptor", "program"))
  row_info <- row_info %>%
    group_by(program_group) %>%
    arrange(gene_type_order, onset_bin) %>%
    ungroup()

  new_row_labels <- paste0(row_info$gene, " (", row_info$lineage, ")")
  norm_mat <- norm_mat[new_row_labels, , drop = FALSE]

  # Row annotation
  row_ha <- rowAnnotation(
    type = row_info$gene_type,
    col = list(type = c("ligand" = "forestgreen", "receptor" = "#fc8d59", "program" = "#8856a7")),
    show_legend = TRUE,
    annotation_name_side = "top"
  )

  # Program group levels
  prog_levels <- if (!is.null(program_order)) {
    intersect(program_order, unique(row_info$program_group))
  } else {
    unique(row_info$program_group)
  }

  # Create heatmap
  ht <- Heatmap(
    norm_mat,
    col = col_fun,
    name = legend_name,
    row_split = factor(row_info$program_group, levels = prog_levels),
    row_title_rot = 0,
    row_title_gp = gpar(fontsize = 10, fontface = "bold"),
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    show_row_names = TRUE,
    row_names_side = "left",
    row_names_gp = gpar(fontsize = 8),
    show_column_names = FALSE,
    column_title = "Classical response axis",
    column_title_side = "bottom",
    rect_gp = gpar(col = "lightgrey", lwd = 0.3),
    na_col = "grey90",
    left_annotation = row_ha,
    heatmap_legend_param = list(title = legend_name, direction = "vertical")
  )

  draw(ht, column_title = paste("Ligand-Program Associations:", title_suffix))

  # Save
  n_rows <- nrow(norm_mat)
  pdf_path <- file.path(FIG_DIR, paste0("ligand_program_heatmap_", file_suffix, ".pdf"))
  pdf(pdf_path, width = 8, height = max(4, n_rows * 0.25 + 2))
  draw(ht, column_title = paste("Ligand-Program Associations:", title_suffix))
  dev.off()

  cat("Ligand-program heatmap saved to:", pdf_path, "\n")
  return(ht)
}
```

## Plot significant ligand-program associations

```{r plot_ligand_program_heatmaps, fig.width=6, fig.height=20}
sig_lig_prog.pos <- lm_df_lig_prog %>%
  filter(padj < 0.01 & slope > 0.8) %>%
  arrange(program, desc(pearson_r))

if (nrow(sig_lig_prog.pos) > 0) {
  plot_program_heatmap_lineage(
    sig_lig_prog.pos, sender_bins, rec_bins, program_response_scores,
    cov_bins, LR_PAIRS, DC1_N_BINS,
    "significant (padj<0.01, positive)", "positive", FIG_DIR,
    include_receptors = FALSE
  )
} else {
  cat("No significant positive ligand-program associations to plot.\n")
}
```

```{r plot_ligand_program_heatmaps, fig.width=6, fig.height=20}
sig_lig_prog.neg <- lm_df_lig_prog %>%
  filter(padj < 0.0001 & slope < -0.8) %>%
  arrange(program, desc(pearson_r))

if (nrow(sig_lig_prog.neg) > 0) {
  plot_program_heatmap_lineage(
    sig_lig_prog.neg, sender_bins, rec_bins, program_response_scores,
    cov_bins, LR_PAIRS, DC1_N_BINS,
    "significant (padj<0.01, negative)", "negative", FIG_DIR,
    include_receptors = FALSE
  )
} else {
  cat("No significant positive ligand-program associations to plot.\n")
}
```

---



## DC1 correlation

Compute Spearman correlation between gene expression/frequency (per DC1 bin) and DC1 bin number for each cell type. This identifies genes (ligands and receptors) whose expression most strongly tracks the ablation response axis.

```{r gene_dc1_spearman_correlation}
# Thresholds for filtering significant correlations
COR_PADJ_THRESH <- 0.01
COR_RHO_THRESH  <- 0.7

# Compute Spearman correlation for each cell type × gene pair
gene_dc1_cor_results <- list()

# Process ligands from sender_celltype_bins
for (key in names(sender_celltype_bins)) {
  parts <- strsplit(key, "__")[[1]]
  ct <- parts[1]
  gene <- parts[2]

  bin_df <- sender_celltype_bins[[key]]

  # Need at least 5 bins for meaningful correlation
  if (nrow(bin_df) < 5) next

  # Spearman correlation between expression and DC1 bin
  cor_test <- tryCatch(
    cor.test(bin_df$DC1_bin30, bin_df$lig_expr_bin, method = "spearman", exact = FALSE),
    error = function(e) NULL
  )

  if (is.null(cor_test)) next

  gene_dc1_cor_results[[length(gene_dc1_cor_results) + 1]] <- data.frame(
    cell_type = ct,
    gene = gene,
    gene_type = "ligand",
    spearman_rho = cor_test$estimate,
    pval = cor_test$p.value,
    n_bins = nrow(bin_df),
    stringsAsFactors = FALSE
  )
}

# Process receptors from rec_celltype_bins
for (key in names(rec_celltype_bins)) {
  parts <- strsplit(key, "__")[[1]]
  gene <- parts[1]  # receptor name is first in rec_celltype_bins keys

  ct <- parts[2]    # cell type is second

  bin_df <- rec_celltype_bins[[key]]

  # Need at least 5 bins for meaningful correlation
  if (nrow(bin_df) < 5) next

  # Spearman correlation between receptor fraction and DC1 bin
  cor_test <- tryCatch(
    cor.test(bin_df$DC1_bin30, bin_df$rec_frac_bin, method = "spearman", exact = FALSE),
    error = function(e) NULL
  )

  if (is.null(cor_test)) next

  gene_dc1_cor_results[[length(gene_dc1_cor_results) + 1]] <- data.frame(
    cell_type = ct,
    gene = gene,
    gene_type = "receptor",
    spearman_rho = cor_test$estimate,
    pval = cor_test$p.value,
    n_bins = nrow(bin_df),
    stringsAsFactors = FALSE
  )
}

gene_dc1_cor_df <- do.call(rbind, gene_dc1_cor_results)
rownames(gene_dc1_cor_df) <- NULL

# Adjust p-values
gene_dc1_cor_df$padj <- p.adjust(gene_dc1_cor_df$pval, method = "BH")

# Add lineage annotation
gene_dc1_cor_df$lineage <- LINEAGE_MAP[gene_dc1_cor_df$cell_type]

# Add chemokine annotation
gene_dc1_cor_df$is_chemokine <- gene_dc1_cor_df$gene %in% c(CHEMOKINE_LIGANDS, CHEMOKINE_RECEPTORS)

# Sort by absolute correlation
gene_dc1_cor_df <- gene_dc1_cor_df %>%
  arrange(desc(abs(spearman_rho)))

# Filter significant correlations
gene_dc1_cor_sig <- gene_dc1_cor_df %>%
  filter(padj < COR_PADJ_THRESH, abs(spearman_rho) > COR_RHO_THRESH)

cat("Gene-DC1 correlations computed:", nrow(gene_dc1_cor_df), "\n")
cat("  Ligands:", sum(gene_dc1_cor_df$gene_type == "ligand"), "\n")
cat("  Receptors:", sum(gene_dc1_cor_df$gene_type == "receptor"), "\n")
cat("Thresholds: padj <", COR_PADJ_THRESH, ", |rho| >", COR_RHO_THRESH, "\n")
cat("Significant:", nrow(gene_dc1_cor_sig), "\n")
cat("  Ligands:", sum(gene_dc1_cor_sig$gene_type == "ligand"), "\n")
cat("  Receptors:", sum(gene_dc1_cor_sig$gene_type == "receptor"), "\n")
cat("  Positive (rho >", COR_RHO_THRESH, "):", sum(gene_dc1_cor_sig$spearman_rho > 0, na.rm = TRUE), "\n")
cat("  Negative (rho <", -COR_RHO_THRESH, "):", sum(gene_dc1_cor_sig$spearman_rho < 0, na.rm = TRUE), "\n")

# Show top associations
print(head(gene_dc1_cor_sig, 20))
```

```{r gene_dc1_cor_save}
# Save results for downstream analysis
write.csv(gene_dc1_cor_df,
          file.path(FIG_DIR, "gene_dc1_spearman_correlations.csv"),
          row.names = FALSE)
cat("Saved gene-DC1 correlations to:", file.path(FIG_DIR, "gene_dc1_spearman_correlations.csv"), "\n")
```

```{r}
# read in if necessary
gene_dc1_cor_df = readr::read_csv(file.path(FIG_DIR, "gene_dc1_spearman_correlations.csv"))
# join to ligand expression information
gene_dc1_cor_df = left_join(gene_dc1_cor_df, ligand_metrics, by = c('cell_type', 'gene', 'lineage'))
# join to expressino metrics
gene_dc1_cor_sig <- gene_dc1_cor_df %>%
  dplyr::filter(padj < COR_PADJ_THRESH, abs(spearman_rho) > COR_RHO_THRESH) %>%
  left_join(ligand_metrics, by = c('cell_type', 'gene'))
```


```{r gene_dc1_cor_heatmap_setup}
# Build expression matrix for significant gene-cell type pairs
# Rows: gene (cell_type), Columns: DC1 bins
# Handles both ligands (from sender_celltype_bins) and receptors (from rec_celltype_bins)

build_gene_expr_matrix <- function(sig_df, sender_bins, rec_bins, bin_ids) {
  if (nrow(sig_df) == 0) return(NULL)

  # Create row labels
sig_df$row_label <- paste0(sig_df$gene, " (", sig_df$cell_type, ")")

  # Build matrix
  mat <- matrix(NA, nrow = nrow(sig_df), ncol = length(bin_ids))
  rownames(mat) <- sig_df$row_label
  colnames(mat) <- bin_ids

  for (i in seq_len(nrow(sig_df))) {
    gene_type <- sig_df$gene_type[i]
    gene <- sig_df$gene[i]
    ct <- sig_df$cell_type[i]

    if (gene_type == "ligand") {
      key <- paste(ct, gene, sep = "__")
      if (!key %in% names(sender_bins)) next
      bin_df <- sender_bins[[key]]
      val_col <- "lig_expr_bin"
    } else {
      key <- paste(gene, ct, sep = "__")
      if (!key %in% names(rec_bins)) next
      bin_df <- rec_bins[[key]]
      val_col <- "rec_frac_bin"
    }

    for (j in seq_along(bin_ids)) {
      idx <- which(bin_df$DC1_bin30 == bin_ids[j])
      if (length(idx) > 0) mat[i, j] <- bin_df[[val_col]][idx]
    }
  }

  return(list(mat = mat, sig_df = sig_df))
}

# Z-score rows
zscore_rows <- function(mat) {
  t(apply(mat, 1, function(x) {
    if (sd(x, na.rm = TRUE) == 0) return(rep(0, length(x)))
    (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
  }))
}

# Smooth rows (3-bin rolling average)
smooth_rows <- function(mat) {
  n_cols <- ncol(mat)
  smoothed <- mat
  for (i in seq_len(nrow(mat))) {
    for (j in seq_len(n_cols)) {
      neighbors <- max(1, j-1):min(n_cols, j+1)
      smoothed[i, j] <- mean(mat[i, neighbors], na.rm = TRUE)
    }
  }
  smoothed
}

# Get bin IDs from cov_bins
bin_ids <- sort(unique(cov_bins$DC1_bin30))
bin18_idx <- which(bin_ids == 18)
```


### ligand correlation heatmaps

```{r gene_dc1_cor_heatmap_positive, fig.width=8, fig.height=10}
# Filter to only cell types I want to use
celltype_rm <- c(
  "Neuron", "Granulocyte", "Adipocyte", "ILC3", "Schwann_cell",
  "Innate-like_T", "ILC2", "MAIT", "Mesenchymal-like_stroma",
  "Pericyte", "Mural", "Smooth_muscle", "Ductal", "Tuft-like",
  "Islet", "ADM", "Acinar",
  "Endothelial_lymphatic", "PSC", "Lymphoid_proliferating",
  "Mesothelial/apCAF"
)

# Positive correlations heatmap
sig_pos <- gene_dc1_cor_sig %>%
  dplyr::filter(spearman_rho > 0) %>%
  dplyr::filter(gene_type == 'ligand' & !(is_chemokine)) %>%
  dplyr::filter(pct_expressing > 0.05 | gene == 'Ifn_I') %>%
  dplyr::filter(!(cell_type %in% celltype_rm)) %>%
  dplyr::arrange(gene, desc(spearman_rho))

if (nrow(sig_pos) > 0) {
  result <- build_gene_expr_matrix(sig_pos, sender_celltype_bins, rec_celltype_bins, bin_ids)
  mat_pos <- result$mat
  sig_pos <- result$sig_df

  # Smooth and z-score
  mat_pos_smooth <- smooth_rows(mat_pos)
  mat_pos_z <- zscore_rows(mat_pos_smooth)

  # Order: group by gene, within each gene order by earliest bin to reach max
  sig_pos$peak_bin <- apply(mat_pos_smooth, 1, function(x) which.max(x))
  sig_pos <- sig_pos %>%
    arrange(gene, peak_bin)

  # Reorder matrix
  mat_pos_z <- mat_pos_z[sig_pos$row_label, , drop = FALSE]

  # Color scale
  col_fun <- colorRamp2(c(-2, 0, 2), c("#4575b4", "white", "#d73027"))

  # Row split by gene
  row_split <- factor(sig_pos$gene, levels = unique(sig_pos$gene))

  ht_pos <- Heatmap(
    mat_pos_z,
    col = col_fun,
    name = "Z-score",
    column_title = "Genes positively correlated with DC1",
    row_title = "Gene (Cell type)",
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    row_split = row_split,
    row_gap = unit(2, "mm"),
    show_row_names = TRUE,
    row_names_side = "left",
    row_names_gp = gpar(fontsize = 8),
    show_column_names = FALSE,
    rect_gp = gpar(col = "lightgrey", lwd = 0.3),
    na_col = "grey90",
    column_title_gp = gpar(fontsize = 12, fontface = "bold")
  )

  draw(ht_pos)
  decorate_heatmap_body("Z-score", {
    x_pos <- bin18_idx / ncol(mat_pos_z)
    grid.lines(x = unit(x_pos, "npc"), y = unit(c(0, 1), "npc"), gp = gpar(col = "black", lwd = 2))
  })

  # Save
  n_rows <- nrow(mat_pos_z)
  fig_height <- max(6, n_rows * 0.25 + 2)
  pdf(file.path(FIG_DIR, "gene_dc1_cor_heatmap_positive.pdf"), width = 8, height = fig_height)
  draw(ht_pos)
  decorate_heatmap_body("Z-score", {
    x_pos <- bin18_idx / ncol(mat_pos_z)
    grid.lines(x = unit(x_pos, "npc"), y = unit(c(0, 1), "npc"), gp = gpar(col = "black", lwd = 2))
  })
  dev.off()
  cat("Saved positive correlation heatmap\n")
} else {
  cat("No positive correlations to plot.\n")
}
```

```{r gene_dc1_cor_heatmap_negative, fig.width=8, fig.height=10}
# Negative correlations heatmap
sig_neg <- gene_dc1_cor_sig %>%
  filter(spearman_rho < 0) %>%
  dplyr::filter(gene_type == 'ligand' & !(is_chemokine)) %>%
  dplyr::filter(pct_expressing > 0.05 | gene == 'Ifn_I') %>%
  dplyr::filter(!(cell_type %in% celltype_rm)) %>%
  arrange(gene, spearman_rho)

if (nrow(sig_neg) > 0) {
  result <- build_gene_expr_matrix(sig_neg, sender_celltype_bins, rec_celltype_bins, bin_ids)
  mat_neg <- result$mat
  sig_neg <- result$sig_df

  # Smooth and z-score
  mat_neg_smooth <- smooth_rows(mat_neg)
  mat_neg_z <- zscore_rows(mat_neg_smooth)

  # Order: group by gene, within each gene order by earliest bin to reach min
  sig_neg$trough_bin <- apply(mat_neg_smooth, 1, function(x) which.min(x))
  sig_neg <- sig_neg %>%
    arrange(gene, trough_bin)

  # Reorder matrix
  mat_neg_z <- mat_neg_z[sig_neg$row_label, , drop = FALSE]

  # Color scale
  col_fun <- colorRamp2(c(-2, 0, 2), c("#4575b4", "white", "#d73027"))

  # Row split by gene
  row_split <- factor(sig_neg$gene, levels = unique(sig_neg$gene))

  ht_neg <- Heatmap(
    mat_neg_z,
    col = col_fun,
    name = "Z-score",
    column_title = "Genes negatively correlated with DC1",
    row_title = "Gene (Cell type)",
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    row_split = row_split,
    row_gap = unit(2, "mm"),
    show_row_names = TRUE,
    row_names_side = "left",
    row_names_gp = gpar(fontsize = 8),
    show_column_names = FALSE,
    rect_gp = gpar(col = "lightgrey", lwd = 0.3),
    na_col = "grey90",
    column_title_gp = gpar(fontsize = 12, fontface = "bold")
  )

  draw(ht_neg)
  decorate_heatmap_body("Z-score", {
    x_pos <- bin18_idx / ncol(mat_neg_z)
    grid.lines(x = unit(x_pos, "npc"), y = unit(c(0, 1), "npc"), gp = gpar(col = "black", lwd = 2))
  })

  # Save
  n_rows <- nrow(mat_neg_z)
  fig_height <- max(6, n_rows * 0.25 + 2)
  pdf(file.path(FIG_DIR, "gene_dc1_cor_heatmap_negative.pdf"), width = 8, height = fig_height)
  draw(ht_neg)
  decorate_heatmap_body("Z-score", {
    x_pos <- bin18_idx / ncol(mat_neg_z)
    grid.lines(x = unit(x_pos, "npc"), y = unit(c(0, 1), "npc"), gp = gpar(col = "black", lwd = 2))
  })
  dev.off()
  cat("Saved negative correlation heatmap\n")
} else {
  cat("No negative correlations to plot.\n")
}
```


---

# Cell Type Feature Frequency by DC Group

This section computes the frequency of cells positive for specific features within each cell type, stratified by DC1 bin groups (high vs low). Niches are pooled by `sample_name + condition_num + dc_group`. Related to analysis for figure S8C. 

```{r celltype_feature_freq_params}
# === Adjustable parameters ===
EXPR_DETECT_THRESHOLD <- 1     # Expression threshold for detecting positive cells (>threshold)
DC_LOW_BINS           <- 3:13  # Bins assigned to "low" DC group
DC_HIGH_BINS          <- 17:27 # Bins assigned to "high" DC group
# Note: bins 1-2, 14-16, and 28+ are excluded from analysis

# Features to test: use all LR genes that exist in obs_df
FEATURES_TO_TEST = adata$var_names
cat("Testing", length(FEATURES_TO_TEST), "features\n")
cat("DC low bins:", min(DC_LOW_BINS), "-", max(DC_LOW_BINS), "\n")
cat("DC high bins:", min(DC_HIGH_BINS), "-", max(DC_HIGH_BINS), "\n")
cat("Expression detection threshold: >", EXPR_DETECT_THRESHOLD, "\n")
```

```{r compute_ct_feature_freq}
# create new obs_df to measure this
exp_df = adata$layers[['counts_segger']] %>% as.matrix() %>% as.data.frame()
colnames(exp_df) = adata$var_names
obs_df = adata$obs
obs_df = cbind(obs_df, exp_df)
obs_df$DC_1_rank = ifelse(is.na(obs_df$DC_1_rank), obs_df$DC_1_rank_env, obs_df$DC_1_rank)

## bin assignment
DC1_N_BINS    <- 30L
DC1_TRIM_LOW  <- 0L
DC1_TRIM_HIGH <- 2L

obs_df$DC1_bin30 <- cut(obs_df$DC_1_rank, breaks = DC1_N_BINS, labels = FALSE)
bin_breaks <- seq(min(obs_df$DC_1_rank, na.rm = TRUE),
                  max(obs_df$DC_1_rank, na.rm = TRUE),
                  length.out = DC1_N_BINS + 1)
bin_lines  <- bin_breaks[(DC1_TRIM_LOW + 1):(DC1_N_BINS - DC1_TRIM_HIGH)]

obs_df <- obs_df[
  !is.na(obs_df$DC1_bin30) &
  obs_df$DC1_bin30 > DC1_TRIM_LOW &
  obs_df$DC1_bin30 <= (DC1_N_BINS - DC1_TRIM_HIGH), ]

DC1_T_MIN <- min(obs_df$DC_1_rank, na.rm = TRUE)
DC1_T_MAX <- max(obs_df$DC_1_rank, na.rm = TRUE)

# Assign cells to DC groups based on their DC1_bin30
# Bins in DC_LOW_BINS -> "low", bins in DC_HIGH_BINS -> "high", others -> NA (excluded)
obs_df$dc_group <- case_when(
  obs_df$DC1_bin30 %in% DC_LOW_BINS  ~ "low",
  obs_df$DC1_bin30 %in% DC_HIGH_BINS ~ "high",
  TRUE ~ NA_character_
)
cat("Cells in low DC group:", sum(obs_df$dc_group == "low", na.rm = TRUE), "\n")
cat("Cells in high DC group:", sum(obs_df$dc_group == "high", na.rm = TRUE), "\n")
cat("Cells excluded (middle/edge bins):", sum(is.na(obs_df$dc_group)), "\n")

# Compute frequencies: for each sample_name + condition_num + dc_group + cell_type,
# calculate fraction of cells positive for each feature

# First, get the unique niche-group combinations with cell counts
niche_group_ct <- obs_df %>%
  filter(!is.na(cell_annotation), !is.na(dc_group), !is.na(sample_name), !is.na(condition_num)) %>%
  group_by(sample_name, condition_num, dc_group, cell_annotation) %>%
  summarize(n_cells = n(), .groups = "drop")

cat("Unique niche-group-celltype combinations:", nrow(niche_group_ct), "\n")
cat("Niches in low DC group:", length(unique(paste(niche_group_ct$sample_name[niche_group_ct$dc_group == "low"],
                                                    niche_group_ct$condition_num[niche_group_ct$dc_group == "low"]))), "\n")
cat("Niches in high DC group:", length(unique(paste(niche_group_ct$sample_name[niche_group_ct$dc_group == "high"],
                                                     niche_group_ct$condition_num[niche_group_ct$dc_group == "high"]))), "\n")

# Compute feature frequencies for each niche-group-celltype combination
# Frequency and offset are based on TOTAL cells in niche (all cell types), not just cells of that type
compute_feature_freqs <- function(df, features, threshold) {
  df_sub <- df %>%
    filter(!is.na(cell_annotation), !is.na(dc_group), !is.na(sample_name), !is.na(condition_num))

  # Get features that exist in data
  features <- intersect(features, names(df_sub))

  # Create binary expression matrix (cells × features)
  expr_mat <- as.matrix(df_sub[, features, drop = FALSE]) > threshold

  # Create niche-group ID (sample_name + condition_num + dc_group) - for total cell counts
  df_sub$niche_group <- paste(df_sub$sample_name, df_sub$condition_num, df_sub$dc_group, sep = "___")

  # Create niche-group-celltype ID - for positive cell counts
  df_sub$niche_group_ct <- paste(df_sub$niche_group, df_sub$cell_annotation, sep = "___")

  # Compute TOTAL cells per niche-group (all cell types)
  niche_total_cells <- df_sub %>%
    group_by(niche_group) %>%
    summarize(n_total_niche = n(), .groups = "drop")

  # Aggregate: sum positive cells per niche-group-celltype
  group_ids <- factor(df_sub$niche_group_ct)
  group_indicator <- model.matrix(~ group_ids - 1)
  colnames(group_indicator) <- levels(group_ids)

  n_pos_mat <- t(group_indicator) %*% expr_mat  # groups × features
  n_cells_ct_tbl <- table(group_ids)  # cells of this type in this niche-group

  # Convert to long format
  result <- expand.grid(
    niche_group_ct = rownames(n_pos_mat),
    feature = colnames(n_pos_mat),
    stringsAsFactors = FALSE
  )
  result$n_positive <- as.vector(n_pos_mat)
  result$n_cells_ct <- as.integer(n_cells_ct_tbl[result$niche_group_ct])

  # Parse niche_group_ct back to components
  parts <- strsplit(result$niche_group_ct, "___")
  result$sample_name <- sapply(parts, `[`, 1)
  result$condition_num <- as.integer(sapply(parts, `[`, 2))
  result$dc_group <- sapply(parts, `[`, 3)
  result$cell_type <- sapply(parts, `[`, 4)
  result$niche_group <- paste(result$sample_name, result$condition_num, result$dc_group, sep = "___")

  # Join total niche cells (all cell types)
  result <- result %>%
    left_join(niche_total_cells, by = "niche_group")

  # Frequency is positive cells of this type / total cells in niche (all types)
  result$freq_positive <- result$n_positive / result$n_total_niche

  result <- result %>%
    select(sample_name, condition_num, dc_group, cell_type, feature,
           n_positive, n_cells_ct, n_total_niche, freq_positive)

  result
}

ct_feature_freq_unfiltered <- compute_feature_freqs(obs_df, FEATURES_TO_TEST, EXPR_DETECT_THRESHOLD)
cat("Total frequency observations:", nrow(ct_feature_freq_unfiltered), "\n")

# Filter to cell types with minimum cells of that type (for Wilcoxon/edgeR per-niche tests)
ct_feature_freq <- ct_feature_freq_unfiltered %>%
  filter(n_cells_ct >= 10)
cat("After filtering (n_cells_ct >= 10):", nrow(ct_feature_freq), "\n")
```

## fisher tests


```{r test_differential_freq_fisher}
# Fisher's exact test for each cell_type × feature combination
# Uses pooled counts across all niches to build 2x2 contingency table
# Uses UNFILTERED data to include sparse cell types like NK_ILC1

MIN_POOLED_CELLS <- 20  # Minimum total cells of that type (pooled across all niches) per group

run_fisher_tests <- function(freq_df, min_cells = MIN_POOLED_CELLS) {
  # Get unique cell_type × feature combinations
  combos <- freq_df %>%
    distinct(cell_type, feature)

  results <- list()

  for (i in seq_len(nrow(combos))) {
    ct <- combos$cell_type[i]
    feat <- combos$feature[i]

    sub <- freq_df %>%
      filter(cell_type == ct, feature == feat)

    # Pool counts across all niches
    n_pos_low <- sum(sub$n_positive[sub$dc_group == "low"], na.rm = TRUE)
    n_pos_high <- sum(sub$n_positive[sub$dc_group == "high"], na.rm = TRUE)
    n_total_low <- sum(sub$n_total_niche[sub$dc_group == "low"], na.rm = TRUE)
    n_total_high <- sum(sub$n_total_niche[sub$dc_group == "high"], na.rm = TRUE)

    # Pool cells of this type across niches
    n_cells_ct_low <- sum(sub$n_cells_ct[sub$dc_group == "low"], na.rm = TRUE)
    n_cells_ct_high <- sum(sub$n_cells_ct[sub$dc_group == "high"], na.rm = TRUE)

    # Filter: need minimum pooled cells of this type in each group
    if (n_cells_ct_low < min_cells || n_cells_ct_high < min_cells) next

    # Need positive totals
    if (n_total_low == 0 || n_total_high == 0) next

    # Build 2x2 contingency table
    # Rows: low, high; Columns: positive, negative
    n_neg_low <- n_total_low - n_pos_low
    n_neg_high <- n_total_high - n_pos_high

    cont_table <- matrix(
      c(n_pos_low, n_neg_low,
        n_pos_high, n_neg_high),
      nrow = 2, byrow = TRUE,
      dimnames = list(c("low", "high"), c("positive", "negative"))
    )

    # Fisher's exact test
    ft <- tryCatch(
      fisher.test(cont_table),
      error = function(e) NULL
    )
    if (is.null(ft)) next

    # Compute odds ratio with Haldane-Anscombe correction (add 0.5 to avoid Inf)
    # OR = (n_pos_high * n_neg_low) / (n_neg_high * n_pos_low)
    # OR > 1 means feature is enriched in HIGH DC group (classical-like)
    # OR < 1 means feature is enriched in LOW DC group (basal-like)
    odds_ratio_corrected <- ((n_pos_high + 0.5) * (n_neg_low + 0.5)) /
                            ((n_neg_high + 0.5) * (n_pos_low + 0.5))

    results[[length(results) + 1]] <- data.frame(
      cell_type = ct,
      feature = feat,
      n_cells_ct_low = n_cells_ct_low,
      n_cells_ct_high = n_cells_ct_high,
      n_pos_low = n_pos_low,
      n_pos_high = n_pos_high,
      n_neg_low = n_neg_low,
      n_neg_high = n_neg_high,
      n_total_niche_low = n_total_low,
      n_total_niche_high = n_total_high,
      freq_low = n_pos_low / n_total_low,
      freq_high = n_pos_high / n_total_high,
      odds_ratio_raw = ft$estimate,
      odds_ratio = odds_ratio_corrected,  # Haldane-Anscombe corrected
      pval = ft$p.value,
      conf_low = ft$conf.int[1],
      conf_high = ft$conf.int[2],
      stringsAsFactors = FALSE
    )
  }

  if (length(results) == 0) return(data.frame())

  result_df <- do.call(rbind, results)
  result_df$padj <- p.adjust(result_df$pval, method = "BH")
  result_df$log2_odds_ratio <- log2(result_df$odds_ratio)  # uses corrected OR
  result_df$lineage <- LINEAGE_MAP[result_df$cell_type]
  result_df <- result_df[order(result_df$padj), ]

  result_df
}

# Use UNFILTERED data for Fisher test (sparse cell types are handled by pooled-count filter)
fisher_results <- run_fisher_tests(ct_feature_freq_unfiltered)
cat("Fisher tests performed:", nrow(fisher_results), "\n")
cat("Significant (padj < 0.1):", sum(fisher_results$padj < 0.1, na.rm = TRUE), "\n")
cat("Significant (padj < 0.05):", sum(fisher_results$padj < 0.05, na.rm = TRUE), "\n")

# Show top results (odds_ratio > 1 = higher in high DC)
cat("\n=== Fisher: Top features ENRICHED in high DC (late ablation) ===\n")
print(head(fisher_results[fisher_results$odds_ratio > 1, ], 20))

cat("\n=== Fisher: Top features DEPLETED in high DC (late ablation) ===\n")
print(head(fisher_results[fisher_results$odds_ratio < 1, ], 20))
```



## plot fisher tests


```{r barplot_fisher_effector_genes, fig.width=5, fig.height=4}
# Horizontal bar plots of effector genes showing odds ratio, shaded by significance

# Genes to plot
EFFECTOR_GENES_BASE <- c("Stat1", "Prf1", "Gzmb", "Gzmc", "Ifng", "Gzma", "Gzmk", "Fasl", "Tnfsf10")
EFFECTOR_GENES_TCONV <- c(EFFECTOR_GENES_BASE, "Tbx21")

for (ct in VOLCANO_CELLTYPES) {
  # Select genes based on cell type
  genes_to_plot <- if (ct == "Tconv") EFFECTOR_GENES_TCONV else EFFECTOR_GENES_BASE

  # Filter to this cell type and genes of interest
  plot_df <- fisher_results %>%
    filter(cell_type == ct, feature %in% genes_to_plot & padj < 0.05) %>%
    mutate(
      significant = padj < PADJ_THRESH,
      sig_label = case_when(
        padj < 0.001 ~ "***",
        padj < 0.01 ~ "**",
        padj < 0.05 ~ "*",
        TRUE ~ ""
      )
    ) 

  if (nrow(plot_df) == 0) {
    cat("No data for cell type:", ct, "\n")
    next
  }

  # Order genes by padj (most significant at top)
  # In ggplot horizontal bars, top = last factor level, so arrange desc(padj) then set levels
  plot_df <- plot_df %>%
    arrange(desc(padj)) %>%
    mutate(feature = factor(feature, levels = feature))

  # Create horizontal bar plot
  p <- ggplot(plot_df, aes(x = log2_odds_ratio, y = feature)) + #, fill = -log10(padj))) +
    geom_col(width = 0.7) +
    geom_vline(xintercept = 0, linetype = "solid", color = "black", linewidth = 0.5) +
    geom_text(aes(label = sig_label,
                  x = log2_odds_ratio + sign(log2_odds_ratio) * 0.02),
              hjust = ifelse(plot_df$log2_odds_ratio > 0, 0, 1),
              size = 4, color = "black") +
    # scale_fill_gradient2(
    #   low = "grey70", mid = "steelblue", high = "darkred",
    #   midpoint = -log10(0.05),
    #   name = "-log10(padj)"
    # ) +
    scale_x_continuous(expand = expansion(mult = c(0.1, 0.15))) +
    labs(
      x = "log2(Odds Ratio)",
      y = NULL,
      title = paste0(ct, " — Effector gene positivity")
    ) +
    theme_tufte() +
    rose_theme +
    theme(
      plot.title = element_text(size = 12, face = "bold"),
      legend.position = "right",
      plot.margin = margin(5, 15, 5, 5)
    )

  print(p)

  # Save
  fig_height <- max(3, length(genes_to_plot) * 0.4 + 1)
  ggsave(file.path(FIG_DIR, paste0("barplot_fisher_effector_", ct, "_", file_suffix, ".svg")),
         p, width = 5, height = fig_height)
  ggsave(file.path(FIG_DIR, paste0("barplot_fisher_effector_", ct, "_", file_suffix, ".pdf")),
         p, width = 5, height = fig_height)
  cat("Saved barplot for:", ct, "\n")
}
```


---

# Tier 2: Spatial Connectivities (60μm radius) Per-Sample Consistency

This section tests whether LR associations are consistent across samples using spatial connectivities (60μm radius neighbors). For each sample, we compute Spearman correlations between niche ligand expression and receptor+ fraction, then filter cell type mode results to those associations that replicate across samples.

## Load Spatial Connectivity Data

```{r load_spatial_niche_data}
library(data.table)

# Load pseudobulk expression per cell type (ligand)
niche_lig_sp_df <- arrow::read_parquet(
  file.path(PRECOMPUTED_DIR,
            "st_data_classical_niche_DC_env_pseudobulk_celltype_expression_spatialConn.parquet")
)
setDT(niche_lig_sp_df)

# Load gene positivity per cell type (receptor)
niche_pos_sp_df <- arrow::read_parquet(
  file.path(PRECOMPUTED_DIR,
            "st_data_classical_niche_DC_env_gene_positivity_spatialConn.parquet")
)
setDT(niche_pos_sp_df)

cat("Loaded spatial connectivity data:\n")
cat("  Ligand expression rows:", nrow(niche_lig_sp_df), "\n")
cat("  Receptor positivity rows:", nrow(niche_pos_sp_df), "\n")
```

## Per-Sample Spearman Correlation Testing

```{r}
# define sender and receiver cell types. 

counts_per_celltype = obs_df$cell_annotation %>% table() %>% sort()
SENDER_CELLTYPES = names(counts_per_celltype)[counts_per_celltype > 500]
RECEIVER_CELLTYPES = names(counts_per_celltype)[counts_per_celltype > 500]
SENDER_CELLTYPES
```


```{r tier2_sp_per_sample, warning=FALSE, error=FALSE}
cat("Running per-sample Tier 2 correlation tests (spatial connectivities)...\n")

# Get unique samples
samples <- unique(obs_df$sample_name[!is.na(obs_df$sample_name)])
cat("Samples:", length(samples), "\n")
print(samples)

# Filter to cancer cells (in_trajectory)
cancer_barcodes_all <- obs_df$barcode[obs_df$in_trajectory]

# Get LR genes
ligand_genes <- unique(LR_PAIRS$ligand)
receptor_genes <- unique(LR_PAIRS$receptor_gene)

tier2_sp_sample_results <- list()

for (samp in samples) {
  cat("\nProcessing sample:", samp, "\n")

  # Filter to cancer cells in this sample
  sample_cancer_df <- obs_df %>%
    filter(in_trajectory, sample_name == samp)

  sample_barcodes <- sample_cancer_df$barcode
  n_sample <- length(sample_barcodes)

  if (n_sample < 100) {
    cat("  Skipping - only", n_sample, "cancer cells\n")
    next
  }
  cat("  Cancer cells:", n_sample, "\n")

  # Filter parquet data to this sample's cancer cells
  niche_pos_sp_samp <- niche_pos_sp_df[barcode %in% sample_barcodes]
  niche_lig_sp_samp <- niche_lig_sp_df[barcode %in% sample_barcodes]

  # Pre-index barcode positions
  sample_barcode_idx <- setNames(seq_along(sample_barcodes), sample_barcodes)

  # Loop through cell type pairs
  for (sender_ct in SENDER_CELLTYPES) {
    # Subset ligand expression for this sender cell type
    lig_ct_subset <- niche_lig_sp_samp[cell_type == sender_ct & gene %in% ligand_genes]

    if (nrow(lig_ct_subset) == 0) next

    # Pivot to matrix
    lig_mat <- matrix(0, nrow = n_sample, ncol = length(ligand_genes),
                      dimnames = list(sample_barcodes, ligand_genes))
    for (g in intersect(ligand_genes, unique(lig_ct_subset$gene))) {
      g_subset <- lig_ct_subset[gene == g]
      lig_mat[sample_barcode_idx[g_subset$barcode], g] <- g_subset$value
    }

    for (receiver_ct in RECEIVER_CELLTYPES) {
      # Subset receptor fraction for this receiver cell type
      rec_ct_subset <- niche_pos_sp_samp[cell_type == receiver_ct & gene %in% receptor_genes]

      if (nrow(rec_ct_subset) == 0) next

      # Pivot to matrix
      rec_mat <- matrix(0, nrow = n_sample, ncol = length(receptor_genes),
                        dimnames = list(sample_barcodes, receptor_genes))
      for (g in intersect(receptor_genes, unique(rec_ct_subset$gene))) {
        g_subset <- rec_ct_subset[gene == g]
        rec_mat[sample_barcode_idx[g_subset$barcode], g] <- g_subset$fraction
      }

      # Loop through LR pairs
      for (i in seq_len(nrow(LR_PAIRS))) {
        lig <- LR_PAIRS$ligand[i]
        rec <- LR_PAIRS$receptor_gene[i]

        if (!lig %in% colnames(lig_mat) || !rec %in% colnames(rec_mat)) next

        lig_vec <- lig_mat[, lig]
        rec_vec <- rec_mat[, rec]

        n_lig_nonzero <- sum(lig_vec > 0)
        n_rec_nonzero <- sum(rec_vec > 0)

        # Require at least 30 non-zero for per-sample (relaxed threshold)
        if (n_lig_nonzero < 30 || n_rec_nonzero < 30) {
          next
        }

        # Correlate
        cor_spearman <- cor.test(log1p(lig_vec), rec_vec, method = "spearman")

        tier2_sp_sample_results[[length(tier2_sp_sample_results) + 1]] <- data.frame(
          sample_name = samp,
          sender_celltype = sender_ct,
          receiver_celltype = receiver_ct,
          ligand = lig,
          receptor = rec,
          rho_spearman = cor_spearman$estimate,
          pval_spearman = cor_spearman$p.value,
          n_cells = n_sample,
          n_lig_nonzero = n_lig_nonzero,
          n_rec_nonzero = n_rec_nonzero,
          stringsAsFactors = FALSE
        )
      }
    }
  }
}

tier2_sp_sample_df <- bind_rows(tier2_sp_sample_results)
rownames(tier2_sp_sample_df) <- NULL

cat("\n=== Per-Sample Results ===\n")
cat("Total tests:", nrow(tier2_sp_sample_df), "\n")
cat("Tests with results:", sum(!is.na(tier2_sp_sample_df$rho_spearman)), "\n")
```




## Compute Sample Consistency Metrics

```{r tier2_sp_sample_consistency}
# Summarize consistency across samples for each LR × cell type combination
cor_threshold <- 0.1

consistency_df <- tier2_sp_sample_df %>%
  filter(!is.na(rho_spearman)) %>%
  group_by(sender_celltype, receiver_celltype, ligand, receptor) %>%
  summarize(
    n_samples = n(),
    n_positive = sum(rho_spearman > cor_threshold, na.rm = TRUE),
    n_sig_positive = sum(rho_spearman > cor_threshold & pval_spearman < 0.05, na.rm = TRUE),
    n_negative = sum(rho_spearman < -cor_threshold, na.rm = TRUE),
    n_sig_negative = sum(rho_spearman < -cor_threshold & pval_spearman < 0.05, na.rm = TRUE),
    mean_rho_spearman = mean(rho_spearman, na.rm = TRUE),
    sd_rho_spearman = sd(rho_spearman, na.rm = TRUE),
    min_rho_spearman = min(rho_spearman, na.rm = TRUE),
    max_rho_spearman = max(rho_spearman, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    frac_positive = n_positive / n_samples,
    frac_sig_positive = n_sig_positive / n_samples,
    consistent = frac_positive >= 0.75 | frac_positive <= 0.25  # >75% agree on direction
  )

cat("\n=== Consistency Summary ===\n")
cat("LR × cell type combinations tested in >=2 samples:", sum(consistency_df$n_samples >= 2), "\n")
cat("Consistent direction (>=75% agree):", sum(consistency_df$consistent & consistency_df$n_samples >= 2), "\n")
cat("Consistently positive (>=75% samples positive):",
    sum(consistency_df$frac_positive >= 0.75 & consistency_df$n_samples >= 2), "\n")
cat("Consistently sig positive (>=75% samples sig positive):",
    sum(consistency_df$frac_sig_positive >= 0.75 & consistency_df$n_samples >= 2), "\n")
```

## Per-Sample Consistency Heatmap

```{r tier2_sp_sample_heatmap, fig.width=10, fig.height=8}
# Heatmap of per-sample correlations for top consistent LR pairs
top_consistent <- consistency_df %>%
  filter(n_samples >= 3, frac_positive >= 0.75, mean_rho_spearman > 0.05) %>%
  arrange(desc(mean_rho_spearman)) %>%
  head(30)

if (nrow(top_consistent) > 0) {
  # Get per-sample rhos for these pairs
  top_sample_data <- tier2_sp_sample_df %>%
    inner_join(top_consistent %>% select(sender_celltype, receiver_celltype, ligand, receptor),
               by = c("sender_celltype", "receiver_celltype", "ligand", "receptor")) %>%
    mutate(lr_label = paste0(ligand, "→", receptor, " (", sender_celltype, "→", receiver_celltype, ")"))

  # Pivot to matrix
  sample_mat <- top_sample_data %>%
    select(lr_label, sample_name, rho_spearman) %>%
    pivot_wider(names_from = sample_name, values_from = rho_spearman) %>%
    tibble::column_to_rownames("lr_label") %>%
    as.matrix()

  # Plot heatmap
  col_fun <- colorRamp2(c(-0.3, 0, 0.3), c("#4575b4", "white", "#d73027"))

  ht <- Heatmap(
    sample_mat,
    col = col_fun,
    name = "Spearman ρ",
    cluster_rows = TRUE,
    cluster_columns = TRUE,
    show_row_names = TRUE,
    row_names_gp = gpar(fontsize = 7),
    show_column_names = TRUE,
    column_names_gp = gpar(fontsize = 9),
    column_title = "Per-Sample LR Colocalization (Spatial 60μm)",
    na_col = "grey90",
    rect_gp = gpar(col = "lightgrey", lwd = 0.3)
  )

  draw(ht)

  # Save
  svg(file.path(FIG_DIR, "tier2_sp_per_sample_consistency_heatmap.svg"), width = 10, height = 8)
  draw(ht)
  dev.off()
  cat("Saved per-sample consistency heatmap\n")
}
```

## Save Per-Sample Results

```{r save_per_sample_results}
# Save per-sample results
write.csv(tier2_sp_sample_df, file.path(PRECOMPUTED_DIR, "lr_dc1_tier2_sp_per_sample.csv"), row.names = FALSE)
write.csv(consistency_df, file.path(PRECOMPUTED_DIR, "lr_dc1_tier2_sp_sample_consistency.csv"), row.names = FALSE)
cat("Saved per-sample results to:", PRECOMPUTED_DIR, "\n")
```

---

# Join Spatial Consistency to cell type mode Results

Join the per-sample consistency metrics from spatial connectivities to cell type mode LM results for filtering.

## Load Data (if necessary)

```{r load_ctmode_and_consistency, eval=FALSE}
# Uncomment to load if necessary
lm_data <- readRDS(file.path(PRECOMPUTED_DIR, "lr_lm_results.rds"))
lm_df_ctmode <- lm_data$lm_df_ctmode
consistency_df <- readr::read_csv(file.path(PRECOMPUTED_DIR, "lr_dc1_tier2_sp_sample_consistency.csv"))

# Save for diagnostics
ligand_metrics = readr::read_csv(file.path(FIG_DIR, "ligand_celltype_filtering_metrics.csv"))
receptor_metrics = readr::read_csv(file.path(FIG_DIR, "receptor_celltype_filtering_metrics.csv"))

# join to cell type mode and 3 df
lm_df_ctmode = left_join(lm_df_ctmode, ligand_metrics %>% dplyr::select(-lineage), by = c('sender_celltype' = 'cell_type', 'ligand' = 'gene'))
lm_df_ctmode = left_join(lm_df_ctmode, receptor_metrics %>% dplyr::select(-lineage), by = c('receiver_celltype' = 'cell_type', 'receptor' = 'gene'), suffix = c('.ligand', '.receptor'))

```



## Join Consistency to cell type mode Results

```{r join_consistency_to_ctmode}
# Join consistency_df to lm_df_ctmode
lm_df_ctmode_with_consistency <- lm_df_ctmode %>%
  left_join(
    consistency_df %>% select(sender_celltype, receiver_celltype, ligand, receptor,
                               n_samples, n_positive, n_sig_positive, #n_negative, n_sig_negative,
                               frac_positive, frac_sig_positive,
                               mean_rho_spearman, sd_rho_spearman, consistent),
    by = c("sender_celltype", "receiver_celltype", "ligand", "receptor")
  )

cat("Joined consistency metrics to lm_df_ctmode\n")
cat("Total associations:", nrow(lm_df_ctmode_with_consistency), "\n")
cat("With consistency data:", sum(!is.na(lm_df_ctmode_with_consistency$n_samples)), "\n")
cat("Number of ligands tested:", length(unique(lm_df_ctmode_with_consistency$ligand)), "\n")
cat("Number of receptors tested:", length(unique(lm_df_ctmode_with_consistency$receptor)), "\n")
cat("Number of cell types tested:", length(unique(lm_df_ctmode_with_consistency$sender_celltype)), "\n")
```



## Apply Consistency Filter

```{r tier1_vs_consistency_comparison}
# Define filtering criteria
TIER1_PADJ <- 0.05
TIER1_SLOPE_MIN <- 0
TIER1_PCT_LIG_MIN <- 0.05
TIER1_PCT_REC_MIN <- 0.05
MIN_N_POSITIVE <- 2

# Get ligand/receptor expression metrics from lm_df_ctmode if available
# Check if pct_expressing columns exist
has_pct_cols <- "pct_expressing.ligand" %in% names(lm_df_ctmode_with_consistency) ||
                "pct_expressing" %in% names(lm_df_ctmode_with_consistency)

# Identify Tier 1 significant associations
# Exceptions that bypass expression filter:
# 1. Mif-Cxcr4 Classical->Macrophage (just below threshold)
# 2. NK_ILC1 receiver_lineage with decent correlation (few cells = low expression %)
if (has_pct_cols && "pct_expressing.ligand" %in% names(lm_df_ctmode_with_consistency)) {
  tier1_sig <- lm_df_ctmode_with_consistency %>%
    dplyr::filter(
      (padj < TIER1_PADJ & slope > TIER1_SLOPE_MIN &
       pct_expressing.ligand > TIER1_PCT_LIG_MIN &
       pct_expressing.receptor > TIER1_PCT_REC_MIN) |
      # Exception 1: Mif-Cxcr4 Classical->Macrophage
      (receptor == "Cxcr4" & receiver_celltype == "Macrophage" &
       ligand == "Mif" & sender_celltype == "Classical" &
       padj < TIER1_PADJ & slope > TIER1_SLOPE_MIN) 
    )
} else {
  # If expression columns don't exist, just filter by padj and slope
  tier1_sig <- lm_df_ctmode_with_consistency %>%
    filter(padj < TIER1_PADJ, slope > TIER1_SLOPE_MIN)
}

cat("\n=== Tier 1 Significant Associations ===\n")
cat("Total Tier 1 significant:", nrow(tier1_sig), "\n")

# Apply consistency filter with exceptions:
# - For slope >= 0: require n_positive >= MIN_N_POSITIVE
# - For slope < 0: require n_sig_negative >= 2
# Exceptions:
# 1. Cxcr4 receptor + Macrophage receiver + Mif ligand + Classical sender (just below threshold)
# 2. NK_ILC1 receiver_lineage: allow if mean_rho_spearman > 0.05 (few cells = low n_positive)
tier1_with_consistency <- tier1_sig %>%
  dplyr::filter(
    # Main filter: direction-appropriate consistency
    (slope > 0 & !is.na(n_positive) & n_positive >= MIN_N_POSITIVE) |
   # (slope < 0 & !is.na(n_sig_negative) & n_sig_negative >= 2) |
    # Exception 1: Mif-Cxcr4 with Classical sender and Macrophage receiver
    (receptor == "Cxcr4" & receiver_celltype == "Macrophage" &
     ligand == "Mif" & sender_celltype == "Classical") |
    # Exception 2: NK_ILC1 receiver with decent correlation
    (receiver_celltype == "NK_ILC1"  & mean_rho_spearman > 0.05)
  )

cat("After consistency filter (with exceptions):", nrow(tier1_with_consistency), "\n")
cat("  Positive slope (n_positive >=", MIN_N_POSITIVE, "):",
    sum(tier1_with_consistency$slope >= 0 & !is.na(tier1_with_consistency$n_positive) &
        tier1_with_consistency$n_positive >= MIN_N_POSITIVE), "\n")
# cat("  Negative slope (n_sig_negative >= 2):",
#     sum(tier1_with_consistency$slope < 0 & !is.na(tier1_with_consistency$n_sig_negative) &
#         tier1_with_consistency$n_sig_negative >= 2), "\n")
cat("  Exceptions retained:\n")
cat("    Mif-Cxcr4 Classical->Macrophage:",
    sum(tier1_with_consistency$receptor == "Cxcr4" &
        tier1_with_consistency$receiver_celltype == "Macrophage" &
        tier1_with_consistency$ligand == "Mif" &
        tier1_with_consistency$sender_celltype == "Classical"), "\n")
cat("    NK_ILC1 receiver (rho > 0.05):",
    sum(tier1_with_consistency$receiver_celltype == "NK_ILC1" &
        !is.na(tier1_with_consistency$mean_rho_spearman) &
        tier1_with_consistency$mean_rho_spearman > 0.05), "\n")

# Identify removed associations (those failing both direction-specific filters)
tier1_removed <- tier1_sig %>%
  filter(
    !((slope >= 0 & !is.na(n_positive) & n_positive >= MIN_N_POSITIVE)) #|
    #  (slope < 0 & !is.na(n_sig_negative) & n_sig_negative >= 2))
  )

cat("\n=== Associations REMOVED by n_positive filter ===\n")
cat("Removed:", nrow(tier1_removed), "associations\n")

if (nrow(tier1_removed) > 0) {
  cat("\nRemoved associations (showing key columns):\n")
  tier1_removed %>%
    dplyr::select(sender_celltype, receiver_celltype, ligand, receptor, pathway,
           slope, padj, n_samples, n_positive, mean_rho_spearman) %>%
    dplyr::arrange(desc(slope)) %>%
    print(n = 50)

  # Breakdown by reason
  cat("\n--- Breakdown of removals ---\n")
  cat("No consistency data (NA):", sum(is.na(tier1_removed$n_positive)), "\n")
  cat("n_positive = 0:", sum(tier1_removed$n_positive == 0, na.rm = TRUE), "\n")
  cat("n_positive = 1:", sum(tier1_removed$n_positive == 1, na.rm = TRUE), "\n")

  # Summarize by ligand
  cat("\n--- Removed by ligand ---\n")
  tier1_removed %>%
    dplyr::count(ligand, sort = TRUE) %>%
    head(20) %>%
    print()

  # Summarize by sender-receiver pair
  cat("\n--- Removed by sender→receiver ---\n")
  tier1_removed %>%
    dplyr::count(sender_celltype, receiver_celltype, sort = TRUE) %>%
    head(15) %>%
    print()

  cat("\n--- Removed by sender ---\n")
  tier1_removed %>%
    dplyr::count(sender_celltype, sort = TRUE) %>%
    head(15) %>%
    print()

  cat("\n--- Removed by receiver ---\n")
  tier1_removed %>%
    dplyr::count(receiver_celltype, sort = TRUE) %>%
    head(15) %>%
    print()

  cat("\n--- Distinct LR pairs removed ---\n")
  tier1_removed %>%
    dplyr::select(ligand, receptor) %>%
    distinct() %>%
    nrow() %>%
    print()
}

# Save the filtered results
write.csv(tier1_with_consistency,
          file.path(PRECOMPUTED_DIR, "lr_dc1_tier1_with_spatial_consistency_filtered.csv"),
          row.names = FALSE)
cat("\nSaved filtered results to:", file.path(PRECOMPUTED_DIR, "lr_dc1_tier1_with_spatial_consistency_filtered.csv"), "\n")
```

---

# Ligand Heatmaps (with Spatial Consistency Filter)

Recreate ligand heatmaps using only associations that pass the spatial consistency filter (n_positive >= 2).

## Separate heatmaps by ligand DC1 correlation direction

Filter ligands by their Spearman correlation with DC1 (using the same rho threshold), then create separate heatmaps for ligands positively vs negatively correlated with DC1.

```{r ligand_dc1_correlation_filter}
COR_PADJ_THRESH = 0.01
COR_RHO_THRESH = .7
# Get ligand-DC1 correlations (filter by same rho threshold)
# Keep cell type to join by both ligand AND sender cell type
ligand_dc1_cor <- gene_dc1_cor_df %>%
  filter(gene_type == "ligand",
         padj < COR_PADJ_THRESH,
         abs(spearman_rho) > COR_RHO_THRESH) %>%
  select(gene, cell_type, spearman_rho)

cat("Ligand-cell type pairs passing DC1 correlation filter (|rho| >", COR_RHO_THRESH, "):", nrow(ligand_dc1_cor), "\n")
cat("  DC1+ pairs (rho > 0):", sum(ligand_dc1_cor$spearman_rho > 0), "\n")
cat("  DC1- pairs (rho < 0):", sum(ligand_dc1_cor$spearman_rho < 0), "\n")

# Join to tier1_with_consistency by ligand AND sender_celltype
tier1_with_lig_dc1 <- tier1_with_consistency %>%
  inner_join(ligand_dc1_cor, by = c("ligand" = "gene", "sender_celltype" = "cell_type"))

cat("\nLR pairs after joining by ligand + sender_celltype:", nrow(tier1_with_lig_dc1), "\n")
cat("  With DC1+ ligands:", sum(tier1_with_lig_dc1$spearman_rho > 0), "\n")
cat("  With DC1- ligands:", sum(tier1_with_lig_dc1$spearman_rho < 0), "\n")

cat("\nDC1+ ligand-celltype pairs:\n")
print(tier1_with_lig_dc1 %>% filter(spearman_rho > 0) %>% distinct(ligand, sender_celltype, spearman_rho))
cat("\nDC1- ligand-celltype pairs:\n")
print(tier1_with_lig_dc1 %>% filter(spearman_rho < 0) %>% distinct(ligand, sender_celltype, spearman_rho))
```

```{r ligand_heatmap_dc1_positive_ligands, fig.width=12, fig.height=30}
# Heatmaps for DC1-POSITIVE correlated LIGANDS (matched by sender cell type)

sig_dc1_pos_lig <- tier1_with_lig_dc1 %>%
  dplyr::filter(spearman_rho > 0)

if (nrow(sig_dc1_pos_lig) > 0) {
  cat("Plotting associations for DC1+ ligands:", nrow(sig_dc1_pos_lig), "pairs\n")
  cat("Ligand-celltype pairs:", paste(unique(paste0(sig_dc1_pos_lig$ligand, " (", sig_dc1_pos_lig$sender_celltype, ")")), collapse = ", "), "\n")

  plot_lr_heatmap_celltype(
    sig_df = sig_dc1_pos_lig,
    sender_celltype_bins = sender_celltype_bins,
    rec_celltype_bins = rec_celltype_bins,
    cov_bins = cov_bins,
    DC1_N_BINS = DC1_N_BINS,
    title_suffix = paste0("DC1+ Ligands (rho > ", COR_RHO_THRESH, ", n_pos>=2)"),
    file_suffix = "dc1_pos_ligands_consistent",
    FIG_DIR = FIG_DIR,
    x_axis_label = X_LAB, format = 'pdf'
  )
} else {
  cat("No associations for DC1+ ligands passing consistency filter.\n")
}
```

```{r ligand_heatmap_dc1_negative_ligands, fig.width=12, fig.height=20}
# Heatmaps for DC1-NEGATIVE correlated LIGANDS (matched by sender cell type)

sig_dc1_neg_lig <- tier1_with_lig_dc1 %>%
  dplyr::filter(spearman_rho < 0)

if (nrow(sig_dc1_neg_lig) > 0) {
  cat("Plotting associations for DC1- ligands:", nrow(sig_dc1_neg_lig), "pairs\n")
  cat("Ligand-celltype pairs:", paste(unique(paste0(sig_dc1_neg_lig$ligand, " (", sig_dc1_neg_lig$sender_celltype, ")")), collapse = ", "), "\n")

  plot_lr_heatmap_celltype(
    sig_df = sig_dc1_neg_lig,
    sender_celltype_bins = sender_celltype_bins,
    rec_celltype_bins = rec_celltype_bins,
    cov_bins = cov_bins,
    DC1_N_BINS = DC1_N_BINS,
    title_suffix = paste0("DC1- Ligands (rho < -", COR_RHO_THRESH, ", n_pos>=2)"),
    file_suffix = "dc1_neg_ligands_consistent",
    FIG_DIR = FIG_DIR,
    x_axis_label = X_LAB, format = 'pdf'
  )
} else {
  cat("No associations for DC1- ligands passing consistency filter.\n")
}


```

## Chemokine Ligand Heatmaps (with Spatial Consistency Filter)

Plots related to Fig S8A. 

```{r chemokine_ligand_heatmap_consistent, fig.width=12, fig.height=15}
# Filter tier1_with_consistency for chemokine ligands only
chemokine_sig <- tier1_with_consistency %>%
  filter(ligand %in% CHEMOKINE_LIGANDS)

if (nrow(chemokine_sig) > 0) {
  cat("Plotting chemokine ligand associations:", nrow(chemokine_sig), "pairs\n")
  cat("Ligands:", paste(unique(chemokine_sig$ligand), collapse = ", "), "\n")
  cat("Receptors:", paste(unique(chemokine_sig$receptor), collapse = ", "), "\n")

  plot_lr_heatmap_celltype(
    sig_df = chemokine_sig,
    sender_celltype_bins = sender_celltype_bins,
    rec_celltype_bins = rec_celltype_bins,
    cov_bins = cov_bins,
    DC1_N_BINS = DC1_N_BINS,
    title_suffix = "Chemokine Ligands (n_pos>=2)",
    file_suffix = "chemokine_ligands_consistent",
    FIG_DIR = FIG_DIR,
    x_axis_label = X_LAB,  receptor_order = c("Cxcr4", "Cxcr6",  "Cxcr3", "Ccr4", "Ccr2"), format = 'pdf'
  )
} else {
  cat("No chemokine ligand associations passing consistency filter.\n")
}
```

---

```{r session_info}
sessionInfo()
```


---

