---
title: "Ligand–Receptor Recruitment Along the Untreated Niche Trajectory"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 5)
```

# Setup

```{r libraries}
suppressPackageStartupMessages({
  library(anndata)
  library(Matrix)
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(ggplot2)
  library(ggthemes)
  library(RColorBrewer)
  library(patchwork)
  library(here)
  library(ComplexHeatmap)
  library(circlize)
  library(data.table)
})


# Source helper functions
source("basal_ablation/src/xenium_utils/pl/lr_heatmap_helpers.R")
RETICULATE_PYTHON='.conda/envs/xenium-rapids/bin/python'

# Paths
CELLCHAT_PATH <- "ref_data/Mus_musculus/rl/CellChatDB.mouse.v2.rda"
H5AD_PATH     <- "basal_ablation/data/processed/xenium/niche_trajectory_untreated_d24/st_data_untreated_niche_DC_all_cancerOnly_k50_env.h5ad"
FIG_DIR       <- "basal_ablation/results/figures/chemokine_dc1/untreated_d24"
SAVE_DIR      <- "basal_ablation/data/processed/xenium/niche_trajectory_untreated_d24"

dir.create(FIG_DIR, recursive = TRUE, showWarnings = FALSE)

rose_theme <- theme(
  axis.text  = element_text(size = 13, color = "black"),
  axis.title = element_text(size = 14, face = "bold")
)

X_LAB <- "Untreated niche trajectory (DC1)"

# Cell-type level filters (Tier 1)
MIN_PCT_CELLTYPE       <- 0.1    # min fraction cells expressing ligand within cell type
MIN_LIG_CELLS_CELLTYPE <- 300     # min cells expressing ligand within cell type
MIN_REC_FRAC_CELLTYPE  <- 0.05    # min receptor+ fraction within cell type

# Chemokine gene lists
CHEMOKINE_LIGANDS   <- c("Ccl2", "Ccl5", "Ccl17",
                          "Cxcl9", "Cxcl10", "Cxcl12", "Cxcl13", "Cxcl16", "Mif")
CHEMOKINE_RECEPTORS <- c("Ccr2", "Ccr4", "Ccr6", "Ccr7",
                          "Cx3cr1", "Cxcr3", "Cxcr4", "Cxcr5", "Cxcr6", "Xcr1")

# Type I IFN ligands
IFN_I_LIGANDS <- c("Ifna2", "Ifna13", "Ifne", "Ifnb1")
MIN_IFN_CELLS <- 30

# Lineage map
LINEAGE_MAP <- c(
  Classical = 'Cancer',
  Basal = 'Cancer',
  Mesenchymal = 'Cancer',
  Cancer_epithelial              = "Cancer",
  `Cancer_epithelial_neuronal-like` = "Cancer",
  Cancer_mesenchymal             = "Cancer",
  Macrophage             = "Mac_Mono",
  Monocyte               = "Mac_Mono",
  cDC1                   = "DC",
  cDC2                   = "DC",
  mregDC                 = "DC",
  pDC                    = "pDC",
  Neutrophil             = "Neutrophil",
  Granulocyte            = "Neutrophil",
  Fibroblast             = "Fibroblast",
  `Mesothelial/apCAF`    = 'Fibroblast',

  PSC                    = "Fibroblast",
  `Mesenchymal-like_stroma` = "Stroma",
  Mesothelial            = "Stroma",
  Mural                  = "Stroma",
  Smooth_muscle          = "Stroma",
  Adipocyte              = "Stroma",
  Pericyte               = 'Stroma',
  CD8_T                  = "T_NK_ILC",
  Tconv                  = "T_NK_ILC",
  NK                     = "T_NK_ILC",
  Treg                   = "T_NK_ILC",
  gdT                    = "T_NK_ILC",
  gdT_CD8                = "T_NK_ILC",
  MAIT                   = "T_NK_ILC",
  `Innate-like_T`        = "T_NK_ILC",
  ILC1                   = "T_NK_ILC",
  ILC2                   = "T_NK_ILC",
  ILC3                   = "T_NK_ILC",
  NK_ILC1                = 'T_NK_ILC',
  Lymphoid_proliferating = "T_NK_ILC",
  B_cell                 = "B_Plasma",
  Plasma_cell            = 'B_Plasma',
  Plasmablast            = "B_Plasma",
  Endothelial_vascular   = "Endothelial",
  Endothelial_lymphatic  = "Endothelial",
  Acinar                 = "Epithelial",
  ADM                    = "Epithelial",
  Ductal                 = "Epithelial",
  Islet                  = "Epithelial",
  `Tuft-like`            = "Epithelial",
  Neuron                 = "Neural",
  Schwann_cell           = "Neural"
)

# Bins for association analysis
DC1_N_BINS <- 30
ANALYSIS_BINS <- 9:28  # restrict to bins 9-28

set.seed(42)
```

---

# Introduction

This notebook tests ligand → receptor recruitment relationships along the **untreated niche trajectory** (DC1). This axis captures the natural heterogeneity in the tumor microenvironment of untreated PDAC samples.

**Statistical approach:** Tier1 (lm) and tier2 (local niche correlation) used to identify LR relationships. Environment cells are assigned DC1 bins matching cancer cell definitions, and analysis is restricted to bins 5-29.

---

# Data Loading

## Load H5AD

```{r load_h5ad}
adata <- anndata::read_h5ad(H5AD_PATH)
cat("Loaded:", H5AD_PATH, "\n")
cat("Shape:", dim(adata), "\n")

obs_df <- adata$obs
var_names <- adata$var_names

cat("\nCell annotations:\n")
print(table(obs_df$cell_annotation))
```

## Build LR Pairs from CellChatDB

```{r build_lr_pairs}
e_cc   <- new.env()
load(CELLCHAT_PATH, envir = e_cc)
cc_int <- e_cc$CellChatDB.mouse$interaction

# For each interaction, find the first receptor subunit gene present in the panel
cc_int$receptor_gene <- sapply(cc_int$receptor.symbol, function(x) {
  syms <- trimws(unlist(strsplit(x, ",")))
  hit  <- syms[syms %in% adata$var_names]
  if (length(hit) == 0) NA_character_ else hit[1]
})

LR_PAIRS <- cc_int[
  cc_int$ligand %in% adata$var_names &
  !is.na(cc_int$receptor_gene),
  c("ligand", "receptor", "receptor.symbol", "receptor_gene", "pathway_name")
]
LR_PAIRS <- LR_PAIRS[!duplicated(paste(LR_PAIRS$ligand, LR_PAIRS$receptor_gene)), ]
rownames(LR_PAIRS) <- NULL

# Annotate chemokine pairs
LR_PAIRS$is_chemokine <- LR_PAIRS$ligand %in% CHEMOKINE_LIGANDS &
                          LR_PAIRS$receptor_gene %in% CHEMOKINE_RECEPTORS
# Expanded: any pair with a chemokine ligand (regardless of receptor)
LR_PAIRS$is_chemokine_expanded <- LR_PAIRS$ligand %in% CHEMOKINE_LIGANDS

cat("LR pairs to test:", nrow(LR_PAIRS), "\n")
cat("  of which chemokine LR:", sum(LR_PAIRS$is_chemokine), "\n")
cat("  of which chemokine expanded:", sum(LR_PAIRS$is_chemokine_expanded), "\n")
print(LR_PAIRS[LR_PAIRS$is_chemokine, c("ligand", "receptor_gene", "pathway_name")])
```

## Load Cytokine Signatures

```{r load_cytokine_signatures}
SIGNAL_ANNO_PATH <- "basal_ablation/data/features/xenium_panel/44CXV9_mPancreas_480g_gene_list_annotated.xlsx"

gmt_to_list <- function(df) {
  gene_col <- names(df)[1]
  cyto_cols <- names(df)[-1]
  out <- list()
  for (cyto in cyto_cols) {
    genes <- df[[gene_col]][df[[cyto]] == 1]
    out[[cyto]] <- genes[!is.na(genes)]
  }
  out
}

signal_anno <- readxl::read_excel(SIGNAL_ANNO_PATH, sheet = "signal_anno")
cytokine_signatures <- gmt_to_list(signal_anno)
cat("Loaded", length(cytokine_signatures), "cytokine signatures\n")
```

---

# Data Preparation

## Compute DC1 Bins

DC1_bin is computed the same way as in the Python notebook: using quantile binning (`pd.qcut` equivalent) on `DC_1_rank` for cancer cells, then assigning environment cells to bins based on their `DC_1_rank_env`.

```{r compute_dc1_bins}
# First check what columns exist
cat("Available obs columns related to DC1:\n")
print(grep("DC", names(obs_df), value = TRUE))

# Compute DC1_bin using quantile binning on cancer cells' DC_1_rank
# This mirrors: pd.qcut(adata.obs['DC_1_rank'], N_BINS, duplicates='drop')

cancer_mask <- !is.na(obs_df$DC_1_rank)
cancer_ranks <- obs_df$DC_1_rank[cancer_mask]

cat("Cancer cells with DC_1_rank:", sum(cancer_mask), "\n")
cat("DC_1_rank range:", range(cancer_ranks), "\n")

# Create quantile breaks (N_BINS + 1 breaks for N_BINS bins)
quantile_breaks <- quantile(cancer_ranks, probs = seq(0, 1, length.out = DC1_N_BINS + 1))

# Handle duplicate breaks by making them unique (like duplicates='drop')
quantile_breaks <- unique(quantile_breaks)
n_actual_bins <- length(quantile_breaks) - 1

cat("Quantile breaks (", n_actual_bins, "bins):\n")
print(quantile_breaks)

# Assign cancer cells to bins (0-indexed to match Python)
obs_df$DC1_bin <- NA_integer_

cancer_bin_assignments <- cut(
  cancer_ranks,
  breaks = quantile_breaks,
  labels = FALSE,
  include.lowest = TRUE
) - 1  # Convert to 0-indexed

obs_df$DC1_bin[cancer_mask] <- cancer_bin_assignments

cat("\nCancer cell bin distribution:\n")
print(table(obs_df$DC1_bin[cancer_mask]))

# Assign environment cells based on DC_1_rank_env using same breaks
env_mask <- is.na(obs_df$DC_1_rank) & !is.na(obs_df$DC_1_rank_env)
cat("\nEnvironment cells to assign:", sum(env_mask), "\n")

if (sum(env_mask) > 0) {
  env_ranks <- obs_df$DC_1_rank_env[env_mask]

  env_bin_assignments <- cut(
    env_ranks,
    breaks = quantile_breaks,
    labels = FALSE,
    include.lowest = TRUE
  ) - 1  # Convert to 0-indexed

  # Handle values outside the range
  env_bin_assignments[env_ranks < min(quantile_breaks)] <- 0
  env_bin_assignments[env_ranks > max(quantile_breaks)] <- n_actual_bins - 1

  obs_df$DC1_bin[env_mask] <- env_bin_assignments

  cat("Environment cell bin distribution:\n")
  print(table(obs_df$DC1_bin[env_mask]))
}

# Use DC1_bin as the bin column (rename for consistency)
obs_df$DC1_bin30 <- obs_df$DC1_bin

cat("\nTotal cells with DC1_bin:", sum(!is.na(obs_df$DC1_bin)), "\n")
cat("Unique bins:", length(unique(obs_df$DC1_bin[!is.na(obs_df$DC1_bin)])), "\n")
```

Analysis bins 0-4 will be filtered out because these are largely sample specific. 

```{r filter_bins}


ANALYSIS_BINS = 5:29
# Filter to analysis bins (5-29)
in_analysis <- obs_df$DC1_bin30 %in% ANALYSIS_BINS
cat("Cells in analysis bins (", min(ANALYSIS_BINS), "-", max(ANALYSIS_BINS), "):",
    sum(in_analysis, na.rm = TRUE), "/", nrow(obs_df), "\n")

obs_df$in_analysis <- in_analysis
```

```{r extract_expression}
# Extract count and log1p matrices
counts_mat <- adata$layers[["counts_segger"]]
log1p_mat  <- adata$layers[["log1p_norm"]]

# Add receptor columns (raw counts + _pos)
receptor_genes <- unique(LR_PAIRS$receptor_gene)
receptor_in_panel <- receptor_genes[receptor_genes %in% var_names]
receptor_idx <- match(receptor_in_panel, var_names)

receptor_counts <- as.matrix(counts_mat[, receptor_idx])
colnames(receptor_counts) <- receptor_in_panel
receptor_pos <- receptor_counts >= 1
colnames(receptor_pos) <- paste0(receptor_in_panel, "_pos")
obs_df <- cbind(obs_df, receptor_counts, receptor_pos)

# Define ligand genes for later
ligand_genes <- unique(LR_PAIRS$ligand)

# Add all features (log1p + counts)
all_features <- adata$var_names
rm_features <- c('GFP', 'tagBFP2', 'CreER')
all_features <- all_features[!(all_features %in% rm_features)]
feature_idx <- match(all_features, var_names)

log1p_subset <- as.matrix(log1p_mat[, feature_idx])
colnames(log1p_subset) <- paste0(all_features, "_log1p")
counts_subset <- as.matrix(counts_mat[, feature_idx])
colnames(counts_subset) <- paste0(all_features, "_counts")
obs_df <- cbind(obs_df, log1p_subset, counts_subset)


# Add library size
obs_df$lib_size <- rowSums(counts_mat)
obs_df$log_lib_size <- log1p(obs_df$lib_size)

# Add lineage
obs_df$lineage <- LINEAGE_MAP[obs_df$cell_annotation]
obs_df$lineage[is.na(obs_df$lineage)] <- "Other"
```

## Define Sender/Receiver Cell Types

```{r define_celltypes}
# Cell types with >= 500 cells in analysis bins
ct_counts <- obs_df %>%
  dplyr::filter(in_analysis) %>%
  dplyr::count(cell_annotation) %>%
  dplyr::filter(n >= 500)

SENDER_CELLTYPES   <- ct_counts$cell_annotation
RECEIVER_CELLTYPES <- ct_counts$cell_annotation

cat("Sender cell types (n >=500):", length(SENDER_CELLTYPES), "\n")
cat("Receiver cell types (n >=500):", length(RECEIVER_CELLTYPES), "\n")
print(SENDER_CELLTYPES)
```

---

# Niche Quantity Precomputation

## Cell-Type Level Counts

```{r niche_precompute}
# Load mingle connectivities
mingle_conn <- adata$obsp[["mingle_connectivities"]]

# --- Niche cell type counts (unchanged, already efficient) ---
cat("Computing niche cell type counts...\n")
for (ct in SENDER_CELLTYPES) {
  ct_mask <- obs_df$cell_annotation == ct
  ct_mask[is.na(ct_mask)] <- FALSE
  col_name <- paste0("n_", gsub("[^A-Za-z0-9]", "_", ct), "_niche_ct")
  obs_df[[col_name]] <- as.numeric(mingle_conn %*% as.numeric(ct_mask))
}

# --- Receptor+ niche counts by cell type (optimized) ---
cat("Computing receptor+ niche counts by cell type...\n")

# Build receptor_pos matrix (cells x receptors)
receptor_pos_cols <- paste0(receptor_genes, "_pos")
receptor_pos_cols <- receptor_pos_cols[receptor_pos_cols %in% names(obs_df)]
receptor_pos_mat <- as.matrix(obs_df[, receptor_pos_cols, drop = FALSE])
storage.mode(receptor_pos_mat) <- "numeric"

# For each cell type, mask and multiply all receptors at once
for (ct in RECEIVER_CELLTYPES) {
  ct_mask <- obs_df$cell_annotation == ct
  ct_mask[is.na(ct_mask)] <- FALSE

  # Multiply: connectivity * (receptor_pos AND cell_type)
  # Result is n_cells x n_receptors
  masked_receptor <- receptor_pos_mat * ct_mask  # broadcast mask to all columns
  niche_counts <- mingle_conn %*% masked_receptor

  # Add columns with proper names
  col_names <- paste0("n_", sub("_pos$", "", receptor_pos_cols), "_",
                      gsub("[^A-Za-z0-9]", "_", ct), "_ct")
  obs_df[, col_names] <- as.data.frame(as.matrix(niche_counts))
}

# Total niche size
obs_df$n_niche_total <- rowSums(mingle_conn)
cat("Niche precomputation complete.\n")
```

---

# Bin-Level Aggregation

## Sender Cell Type Bins (Ligand Expression)

```{r sender_celltype_bins}
sender_celltype_bins <- list()

# Filter to analysis bins
analysis_df <- obs_df %>% dplyr::filter(in_analysis)

for (ct in SENDER_CELLTYPES) {
  ct_df <- analysis_df %>% filter(cell_annotation == ct)
  if (nrow(ct_df) < 50) next

  # Get ligand count columns that exist
  counts_cols <- paste0(ligand_genes, "_counts")
  counts_cols <- counts_cols[counts_cols %in% names(ct_df)]
  if (length(counts_cols) == 0) next

  # Compute filtering stats for all ligands at once
  lig_stats <- ct_df %>%
    summarize(across(all_of(counts_cols),
                     list(pct = ~mean(. > 0, na.rm = TRUE),
                          n_cells = ~sum(. > 0, na.rm = TRUE))))

  # Aggregate per bin for all ligands at once
  bin_agg_all <- ct_df %>%
    group_by(DC1_bin30) %>%
    summarize(
      n_sender = n(),
      total_lib_size = sum(lib_size, na.rm = TRUE),
      across(all_of(counts_cols),
             list(n_lig_cells = ~sum(. > 0, na.rm = TRUE),
                  sum_counts = ~sum(., na.rm = TRUE))),
      .groups = "drop"
    )

  # Process each ligand's results (filtering + reshaping)
  for (lig in ligand_genes) {
    counts_col <- paste0(lig, "_counts")
    if (!counts_col %in% counts_cols) next

    # Check filtering thresholds
    lig_pct <- lig_stats[[paste0(counts_col, "_pct")]]
    lig_n_cells <- lig_stats[[paste0(counts_col, "_n_cells")]]
    min_cells_thresh <- if (lig %in% IFN_I_LIGANDS) MIN_IFN_CELLS else MIN_LIG_CELLS_CELLTYPE
    if (lig_pct < MIN_PCT_CELLTYPE && lig_n_cells < min_cells_thresh) next

    # Extract this ligand's bin aggregation
    bin_agg <- bin_agg_all %>%
      transmute(
        DC1_bin30,
        n_sender,
        n_lig_cells = .data[[paste0(counts_col, "_n_lig_cells")]],
        lig_expr_bin = log1p(.data[[paste0(counts_col, "_sum_counts")]] / pmax(total_lib_size, 1))
      )

    key <- paste(ct, lig, sep = "__")
    sender_celltype_bins[[key]] <- bin_agg
  }
}

cat("Sender cell type bins:", length(sender_celltype_bins), "\n")
```

## Covariate Bins

```{r cov_bins}
cov_bins <- analysis_df %>%
  filter(!is.na(DC_1_rank) | !is.na(DC_1_rank_env)) %>%
  group_by(DC1_bin30) %>%
  summarize(
    n_cells = n(),
    .groups = "drop"
  )

cat("Covariate bins:\n")
print(cov_bins)
```

## Receptor Cell Type Bins

Uses niche-level receptor counts from cancer cells' neighborhoods.

```{r rec_celltype_bins}
# First compute niche totals per bin (total niche cells across all cancer cells per bin)
# so receptor positive cells near cancer cells
cancer_analysis_df <- analysis_df %>% filter(!is.na(DC_1_rank))

niche_totals_bin <- cancer_analysis_df %>%
  group_by(DC1_bin30) %>%
  summarize(n_niche_total_bin = sum(n_niche_total, na.rm = TRUE), .groups = "drop")

# Receptor cell type bins using niche-level counts
rec_celltype_bins <- list()

for (rec in receptor_genes) {
 for (ct in RECEIVER_CELLTYPES) {
    # Column names for niche counts
    ct_safe <- gsub("[^A-Za-z0-9]", "_", ct)
    rec_col <- paste0("n_", rec, "_", ct_safe, "_ct")
    ct_col <- paste0("n_", ct_safe, "_niche_ct")

    if (!rec_col %in% names(cancer_analysis_df)) next
    if (!ct_col %in% names(cancer_analysis_df)) next

    tmp <- cancer_analysis_df %>%
      group_by(DC1_bin30) %>%
      summarize(
        n_rec_total = sum(.data[[rec_col]], na.rm = TRUE),
        n_ct_total = sum(.data[[ct_col]], na.rm = TRUE),
        .groups = "drop"
      ) %>%
      left_join(cov_bins, by = "DC1_bin30") %>%
      left_join(niche_totals_bin, by = "DC1_bin30")

    tmp$rec_frac_within_ct <- tmp$n_rec_total / pmax(tmp$n_ct_total, 1)
    tmp$rec_frac_bin <- tmp$n_rec_total / pmax(tmp$n_niche_total_bin, 1)
    tmp$rec_frac_within_ct_overall <- sum(tmp$n_rec_total, na.rm = TRUE) /
                                       pmax(sum(tmp$n_ct_total, na.rm = TRUE), 1)

    rec_celltype_bins[[paste(rec, ct, sep = "__")]] <- tmp
  }
}

# Filter based on overall receptor+ fraction within cell type
rec_celltype_bins <- Filter(
  function(df) df$rec_frac_within_ct_overall[1] >= MIN_REC_FRAC_CELLTYPE,
  rec_celltype_bins
)

cat("Receptor cell type bins after filtering:", length(rec_celltype_bins), "\n")
```

---

# Filtering Diagnostics

Compute expression metrics for all gene × cell_type combinations to help identify appropriate filtering thresholds.

```{r filtering_diagnostics}
# Vectorized function to compute filtering metrics for all gene × cell_type combinations
compute_filtering_metrics <- function(df, genes, cell_types, col_name = "cell_annotation",
                                       threshold = 0) {
  # Filter to genes that exist in data
  genes <- intersect(genes, names(df))
  if (length(genes) == 0) return(data.frame())

  # Filter to valid cell types
  df <- df[!is.na(df[[col_name]]) & df[[col_name]] %in% cell_types, ]
  if (nrow(df) == 0) return(data.frame())

  # Compute n_total per cell type
  n_total_df <- df %>%
    group_by(cell_type = .data[[col_name]]) %>%
    summarize(n_total = n(), .groups = "drop")

  # Vectorized: compute expression stats for all genes at once
  # Convert gene columns to binary (expressing or not)
  expr_mat <- as.matrix(df[, genes, drop = FALSE]) > threshold

  # Sum per cell type using matrix multiplication with indicator
  ct_factor <- factor(df[[col_name]], levels = cell_types)
  ct_indicator <- model.matrix(~ ct_factor - 1)
  colnames(ct_indicator) <- levels(ct_factor)

  # n_expressing = t(indicator) %*% expr_mat
  n_expr_mat <- t(ct_indicator) %*% expr_mat

  # Convert to long format
  results <- expand.grid(cell_type = rownames(n_expr_mat), gene = colnames(n_expr_mat),
                          stringsAsFactors = FALSE)
  results$n_expressing <- as.vector(n_expr_mat)

  # Join n_total
  results <- results %>%
    left_join(n_total_df, by = "cell_type") %>%
    mutate(pct_expressing = n_expressing / n_total) %>%
    filter(n_total > 0) %>%
    select(gene, cell_type, pct_expressing, n_expressing, n_total)

  results
}

# Get all LR genes
all_lr_ligands <- unique(LR_PAIRS$ligand)
all_lr_receptors <- unique(LR_PAIRS$receptor_gene)

# Get ALL features with _counts columns (for general filtering)
all_count_cols <- grep("_counts$", names(obs_df), value = TRUE)
all_features <- gsub("_counts$", "", all_count_cols)

# Create a df with all features for general metrics
all_features_df <- obs_df[, c("cell_annotation", all_count_cols), drop = FALSE]
names(all_features_df) <- c("cell_annotation", all_features)

# All cell types in the data
all_cell_types <- unique(obs_df$cell_annotation[!is.na(obs_df$cell_annotation)])

# Compute metrics for ALL features across ALL cell types
all_features_metrics <- compute_filtering_metrics(
  all_features_df,
  all_features,
  all_cell_types,
  threshold = 0
)

if (nrow(all_features_metrics) > 0) {
  all_features_metrics$lineage <- LINEAGE_MAP[all_features_metrics$cell_type]
}

# Ligand metrics - use _counts columns directly
# Build a subset df with just the counts columns, renamed to gene names
ligand_count_cols <- paste0(all_lr_ligands, "_counts")
ligand_count_cols <- ligand_count_cols[ligand_count_cols %in% names(obs_df)]
ligand_genes_present <- gsub("_counts$", "", ligand_count_cols)

# Create a temporary df with just cell_annotation and ligand counts (renamed)
ligand_df <- obs_df[, c("cell_annotation", ligand_count_cols), drop = FALSE]
names(ligand_df) <- c("cell_annotation", ligand_genes_present)

# Ligand metrics (threshold = 0 to match sender_bins)
ligand_metrics <- compute_filtering_metrics(
  ligand_df,
  ligand_genes_present,
  SENDER_CELLTYPES,
  threshold = 0
)

if (nrow(ligand_metrics) > 0) {
  ligand_metrics$passes_pct <- ligand_metrics$pct_expressing >= MIN_PCT_CELLTYPE
  ligand_metrics$passes_count <- ligand_metrics$n_expressing >= MIN_LIG_CELLS_CELLTYPE
  ligand_metrics$passes_filter <- ligand_metrics$passes_pct | ligand_metrics$passes_count
  ligand_metrics$lineage <- LINEAGE_MAP[ligand_metrics$cell_type]
}

# Receptor metrics (threshold = 1 to match receptor positivity)
# Use raw receptor count columns (not _pos)
receptor_genes_present <- all_lr_receptors[all_lr_receptors %in% names(obs_df)]

# Create a temporary df with just cell_annotation and receptor counts
receptor_df <- obs_df[, c("cell_annotation", receptor_genes_present), drop = FALSE]

receptor_metrics <- compute_filtering_metrics(
  receptor_df,
  receptor_genes_present,
  RECEIVER_CELLTYPES,
  threshold = 1
)

if (nrow(receptor_metrics) > 0) {
  receptor_metrics$passes_filter <- receptor_metrics$pct_expressing >= MIN_REC_FRAC_CELLTYPE
  receptor_metrics$lineage <- LINEAGE_MAP[receptor_metrics$cell_type]
}

# Save for diagnostics
write.csv(all_features_metrics, file.path(FIG_DIR, "all_features_celltype_filtering_metrics.csv"), row.names = FALSE)
write.csv(ligand_metrics, file.path(FIG_DIR, "ligand_celltype_filtering_metrics.csv"), row.names = FALSE)
write.csv(receptor_metrics, file.path(FIG_DIR, "receptor_celltype_filtering_metrics.csv"), row.names = FALSE)

cat("All features metrics:", nrow(all_features_metrics), "gene × cell_type combinations\n")
cat("  Unique genes:", length(unique(all_features_metrics$gene)), "\n")
cat("  Unique cell types:", length(unique(all_features_metrics$cell_type)), "\n")
cat("Ligand metrics:", nrow(ligand_metrics), "gene × cell_type combinations\n")
cat("  Passing filter:", sum(ligand_metrics$passes_filter, na.rm = TRUE), "\n")
cat("Receptor metrics:", nrow(receptor_metrics), "gene × cell_type combinations\n")
cat("  Passing filter:", sum(receptor_metrics$passes_filter, na.rm = TRUE), "\n")
```


# Feature Correlation with DC1

## Ligand Correlations

```{r ligand_dc1_correlation}
# Correlate bin-level gene expression with DC bin number for each cell type
dc1_corr_results <- list()

COR_CELLTYPES = c('Cancer_epithelial', 'Fibroblast', 'Macrophage')

for (ct in COR_CELLTYPES) {
  ct_df <- analysis_df %>%
    filter(cell_annotation == ct, !is.na(DC1_bin30))

  if (nrow(ct_df) < 100) next

  # Get all log1p columns
  log1p_cols <- paste0(all_features, "_log1p")
  log1p_cols <- log1p_cols[log1p_cols %in% names(ct_df)]

  # Aggregate expression by bin for all genes at once
  bin_agg <- ct_df %>%
    group_by(DC1_bin30) %>%
    summarize(
      n_cells = n(),
      across(all_of(log1p_cols), ~mean(., na.rm = TRUE)),
      .groups = "drop"
    )

  # Correlate each gene's bin-level expression with bin number
  for (gene in all_features) {
    gene_col <- paste0(gene, "_log1p")
    if (!gene_col %in% names(bin_agg)) next

    bin_expr <- bin_agg[[gene_col]]
    bin_nums <- bin_agg$DC1_bin30

    # Skip if no variance
    if (sd(bin_expr, na.rm = TRUE) == 0) next

    valid_idx <- !is.na(bin_expr) & !is.na(bin_nums)
    if (sum(valid_idx) < 5) next

    cor_test <- cor.test(bin_nums[valid_idx], bin_expr[valid_idx], method = "spearman")

    dc1_corr_results[[length(dc1_corr_results) + 1]] <- data.frame(
      cell_type = ct,
      gene = gene,
      rho = cor_test$estimate,
      pval = cor_test$p.value,
      n_bins = sum(valid_idx),
      stringsAsFactors = FALSE
    )
  }
}

dc1_corr_df <- bind_rows(dc1_corr_results)
dc1_corr_df$padj <- p.adjust(dc1_corr_df$pval, method = "BH")

cat("DC1 bin-level correlation tests:", nrow(dc1_corr_df), "\n")
```

```{r}
# write out results joining to expression statistics
dc1_corr_df = dplyr::left_join(dc1_corr_df, all_features_metrics, by = c('cell_type', 'gene'))
write.csv(dc1_corr_df, file.path(FIG_DIR, 'untreated_dc1_spearman_correlation.csv'), row.names = F, quote = F)
```

```{r}
# read in 
dc1_corr_df = readr::read_csv(file.path(FIG_DIR, 'untreated_dc1_spearman_correlation.csv'))
```

```{r}

dc1_corr_df %>% dplyr::filter(pct_expressing > MIN_PCT_CELLTYPE) %>% View()
```

```{r}
nrow(dc1_corr_df)
```

```{r}
dc1_corr_df %>% dplyr::filter(pct_expressing > MIN_PCT_CELLTYPE & abs(rho) > .7) %>% nrow()
```


## Correlated feature heatmap

```{r correlated_ligand_heatmap_function}
# Function to create DC1-correlated feature heatmap for a given cell type
plot_dc1_correlated_heatmap <- function(ct_name, dc1_corr_df, obs_df, top_n = 20, pct_filter = .1, add_genes = "",
                                        rho_threshold = 0.7,
                                        file_suffix = "_dc1_correlated_features_heatmap.svg") {
  # Get top N positive and top N negative correlated genes
  pos_genes <- dc1_corr_df %>%
    filter(cell_type == ct_name & rho > rho_threshold & pct_expressing > pct_filter) %>%
    arrange(desc(rho)) %>%
    head(top_n)

  neg_genes <- dc1_corr_df %>%
    filter(cell_type == ct_name & rho < -rho_threshold & pct_expressing > pct_filter) %>%
    arrange(rho) %>%
    head(top_n)

  # Handle add_genes separately - assign to pos/neg based on actual rho
  if (length(add_genes) > 0 && !all(add_genes == "")) {
    add_genes_df <- dc1_corr_df %>%
      filter(cell_type == ct_name, gene %in% add_genes)
    pos_genes <- bind_rows(pos_genes, add_genes_df %>% filter(rho >= 0)) %>% distinct(gene, .keep_all = TRUE)
    neg_genes <- bind_rows(neg_genes, add_genes_df %>% filter(rho < 0)) %>% distinct(gene, .keep_all = TRUE)
  }

  corr_genes_df <- bind_rows(pos_genes, neg_genes)

  cat("\n", ct_name, ":\n")
  cat("  Top", top_n, "positive:", nrow(pos_genes), "\n")
  cat("  Top", top_n, "negative:", nrow(neg_genes), "\n")

  if (nrow(corr_genes_df) == 0) {
    cat("  No correlated genes found.\n")
    return(NULL)
  }

  # Get cells in analysis bins
  ct_analysis_df <- obs_df %>%
    filter(cell_annotation == ct_name, in_analysis)

  corr_genes <- corr_genes_df$gene

  # Compute bin-level mean expression
  ct_bin_expr <- ct_analysis_df %>%
    group_by(DC1_bin30) %>%
    summarize(
      n_cells = n(),
      across(all_of(paste0(corr_genes, "_log1p")), ~mean(., na.rm = TRUE)),
      .groups = "drop"
    )

  # Build matrix
  bin_ids <- sort(unique(ct_bin_expr$DC1_bin30))
  n_bins <- length(bin_ids)
  bin_centres <- (bin_ids - 0.5) / DC1_N_BINS

  mat <- matrix(NA, nrow = length(corr_genes), ncol = n_bins)
  rownames(mat) <- corr_genes
  colnames(mat) <- sprintf("%.2f", bin_centres)

  for (gene in corr_genes) {
    col_name <- paste0(gene, "_log1p")
    if (col_name %in% names(ct_bin_expr)) {
      mat[gene, ] <- ct_bin_expr[[col_name]]
    }
  }

  # Row groups by correlation direction
  row_groups <- ifelse(corr_genes_df$rho > 0, "DC1 Positive", "DC1 Negative")
  row_groups <- factor(row_groups, levels = c("DC1 Positive", "DC1 Negative"))
  
  print(head(mat))

  # Z-score normalize
  mat_z <- t(apply(mat, 1, function(x) {
    s <- sd(x, na.rm = TRUE)
    if (is.na(s) || s == 0) return(rep(0, length(x)))
    (x - mean(x, na.rm = TRUE)) / s
  }))

  # Smooth with 3-bin rolling average
  mat_smooth <- mat_z
  for (i in seq_len(nrow(mat_smooth))) {
    for (j in seq_len(ncol(mat_smooth))) {
      neighbors <- max(1, j-1):min(ncol(mat_smooth), j+1)
      mat_smooth[i, j] <- mean(mat_z[i, neighbors], na.rm = TRUE)
    }
  }

  col_fun <- colorRamp2(c(-2, 0, 2), c("#4575b4", "white", "#d73027"))

  ht <- Heatmap(
    mat_smooth,
    col = col_fun,
    name = "Z-score",
    row_split = row_groups,
    row_title_rot = 0,
    row_title_gp = gpar(fontsize = 10, fontface = "bold"),
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    show_row_names = TRUE,
    row_names_side = "left",
    row_names_gp = gpar(fontsize = 8),
    show_column_names = FALSE,
    column_title = X_LAB,
    column_title_side = "bottom",
    rect_gp = gpar(col = "lightgrey", lwd = 0.3),
    na_col = "grey90",
    row_gap = unit(2, "mm"),
    use_raster = FALSE
  )

  plot_title <- paste0("DC1-Correlated Features (", ct_name, ", top ", top_n, " each direction)")

  # Draw

  draw(ht,
       column_title = plot_title,
       column_title_gp = gpar(fontsize = 12, fontface = "bold"),
       padding = unit(c(2, 2, 2, 2), "mm"))

  # Save SVG
  ct_safe <- gsub("[^A-Za-z0-9]", "_", tolower(ct_name))
  svg_path <- file.path(FIG_DIR, paste0(ct_safe, file_suffix))
  n_rows <- nrow(mat_smooth)
  pdf(svg_path, width = 10, height = max(6, n_rows * 0.25 + 2))
  draw(ht,
       column_title = plot_title,
       column_title_gp = gpar(fontsize = 12, fontface = "bold"),
       padding = unit(c(2, 2, 2, 2), "mm"))
  dev.off()
  cat("  Saved:", svg_path, "\n")

  invisible(ht)
}
```

```{r correlated_heatmap_cancer, fig.width=10, fig.height=12}
plot_dc1_correlated_heatmap("Cancer_epithelial", dc1_corr_df, obs_df, top_n = 20, rho_threshold = .7, file_suffix = '_dc1_correlated_features_heatmap.pdf')
```

```{r correlated_heatmap_macrophage, fig.width=10, fig.height=12}
plot_dc1_correlated_heatmap("Macrophage", dc1_corr_df, obs_df, top_n = 20, rho_threshold = .7, file_suffix = '_dc1_correlated_features_heatmap.pdf')
```

```{r correlated_heatmap_fibroblast, fig.width=10, fig.height=12}
plot_dc1_correlated_heatmap("Fibroblast", dc1_corr_df, obs_df, top_n = 20, rho_threshold = .7, file_suffix = '_dc1_correlated_features_heatmap.pdf')
```

### filtered features heatmaps for main

```{r}
fib_genes = c('Pi16', 'Pdgfra', 'Cd34', 'Gli1', 'Eng', 'Igf1', # classical associated
              'Tnc', 'Acta2', 'Fap', 'Thbs1', 'Plaur', 'Pdgfrb') # basal associated

plot_dc1_correlated_heatmap("Fibroblast", dc1_corr_df %>% dplyr::filter(gene %in% fib_genes), obs_df, top_n = 20, pct_filter = .05, file_suffix = "_dc1_correlated_select_features_heatmap.svg")
```

```{r}
mac_genes = c('Folr2', 'C1qb', 'Mafb', 'Clec10a', 'Cx3cr1', 'Trem2',  # classical associated
              'Arg1', 'Axl', 'Fabp4', 'Itgax', 'Plaur', 'Thbs1') # basal associated

plot_dc1_correlated_heatmap("Macrophage", dc1_corr_df %>% dplyr::filter(gene %in% mac_genes), obs_df, top_n = 20, pct_filter = .05, file_suffix = "_dc1_correlated_select_features_heatmap.svg")
```


```{r}
cancer_genes = c('Lgals4', 'Agr2', 'Tspan8', 'Sox9', 'Lrp5', 'Wnt4',  # classical associated
              'Slc4a11', 'Krt17', 'Lamc2', 'Lamb3', 'Itga5', 'Itgb1') # basal associated

plot_dc1_correlated_heatmap("Cancer_epithelial", dc1_corr_df %>% dplyr::filter(gene %in% cancer_genes), obs_df, top_n = 20, pct_filter = .05, file_suffix = "_dc1_correlated_select_features_heatmap.svg")
```

---

---

# Linear Model Testing (Tier 1)

Testing for relationship between expression and receptor positive cells. 


```{r lm_tier1}
# Test cell-type level associations
lm_results <- list()

for (i in seq_len(nrow(LR_PAIRS))) {
  lig <- LR_PAIRS$ligand[i]
  rec <- LR_PAIRS$receptor_gene[i]
  pathway <- LR_PAIRS$pathway_name[i]
  is_chemo <- LR_PAIRS$is_chemokine[i]
  is_chemo_exp <- LR_PAIRS$is_chemokine_expanded[i]

  # Test all sender-receiver cell type combinations
  for (sender_ct in SENDER_CELLTYPES) {
    sender_key <- paste(sender_ct, lig, sep = "__")
    if (!sender_key %in% names(sender_celltype_bins)) next
    sender_df <- sender_celltype_bins[[sender_key]]

    for (receiver_ct in RECEIVER_CELLTYPES) {
      rec_key <- paste(rec, receiver_ct, sep = "__")
      if (!rec_key %in% names(rec_celltype_bins)) next
      rec_df <- rec_celltype_bins[[rec_key]]

      # Merge on bin
      merged <- inner_join(sender_df, rec_df, by = "DC1_bin30", suffix = c("_lig", "_rec"))
      if (nrow(merged) < 5) next

      # Fit LM
      fit <- tryCatch({
        lm(rec_frac_bin ~ lig_expr_bin, data = merged)
      }, error = function(e) NULL)

      if (is.null(fit)) next

      summ <- summary(fit)
      coef_df <- coef(summ)

      if (!"lig_expr_bin" %in% rownames(coef_df)) next

      lm_results[[length(lm_results) + 1]] <- data.frame(
        mode = "tier1",
        sender_celltype = sender_ct,
        receiver_celltype = receiver_ct,
        ligand = lig,
        receptor = rec,
        pathway = pathway,
        is_chemokine = is_chemo,
        is_chemokine_expanded = is_chemo_exp,
        slope = coef_df["lig_expr_bin", "Estimate"],
        se = coef_df["lig_expr_bin", "Std. Error"],
        tval = coef_df["lig_expr_bin", "t value"],
        pval = coef_df["lig_expr_bin", "Pr(>|t|)"],
        r2 = summ$r.squared,
        n_bins = nrow(merged),
        stringsAsFactors = FALSE
      )
    }
  }
}

lm_df_tier1 <- bind_rows(lm_results)
lm_df_tier1$padj <- p.adjust(lm_df_tier1$pval, method = "BH")

# join to receptor and ligand expression information
lm_df_tier1 = left_join(lm_df_tier1, ligand_metrics %>% dplyr::select(gene, cell_type, pct_expressing, n_expressing), 
                        by = c('ligand' = 'gene', 'sender_celltype' = 'cell_type')) %>%
  left_join(receptor_metrics %>% dplyr::select(gene, cell_type, pct_expressing, n_expressing), 
                        by = c('receptor' = 'gene', 'receiver_celltype' = 'cell_type'), 
            suffix= c('.ligand', '.receptor'))


cat("Tier 1 tests:", nrow(lm_df_tier1), "\n")
cat("Significant (padj < 0.05):", sum(lm_df_tier1$padj < 0.05), "\n")
cat("Positive associations:", sum(lm_df_tier1$padj < 0.05 & lm_df_tier1$slope > 0), "\n")
```

```{r save_lm_results}
# Save results
write.csv(lm_df_tier1, file.path(SAVE_DIR, "lr_dc1_lm_results_tier1.csv"), row.names = FALSE)
cat("Saved results to:", file.path(SAVE_DIR, "lr_dc1_lm_results_tier1.csv"), "\n")
```

```{r}
# read in 
lm_df_tier1 = read.csv(file.path(SAVE_DIR, "lr_dc1_lm_results_tier1.csv"), h = T, stringsAsFactors = F)
```

## representative scatterplot

```{r representative_scatterplot, fig.width=6, fig.height=5}
# Representative scatterplot: Cancer_epithelial Pdgfb -> PSC Pdgfrb
sender_ct <- "Cancer_epithelial"
receiver_ct <- "Fibroblast"
lig <- "Tgfb1"
rec <- "Tgfbr2"

# Get sender and receiver bin data
sender_key <- paste(sender_ct, lig, sep = "__")
rec_key <- paste(rec, receiver_ct, sep = "__")

sender_df <- sender_celltype_bins[[sender_key]]
rec_df <- rec_celltype_bins[[rec_key]]

# Merge on bin
plot_df <- inner_join(sender_df, rec_df, by = "DC1_bin30", suffix = c("_lig", "_rec"))

# Create scatterplot with regression line and plasma color palette
ggplot(plot_df, aes(x = lig_expr_bin, y = rec_frac_bin, color = DC1_bin30)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linetype = "dashed") +
  scale_color_viridis_c(option = "plasma", name = "DC1 Bin") +
  labs(
    x = paste0(sender_ct, " ", lig, " expression (log1p)"),
    y = paste0(rec, "+ fraction in niche (", receiver_ct, ")"),
    title = paste0("Ligand-Receptor Association: ", lig, " -> ", rec),
    subtitle = paste0("Sender: ", sender_ct, " | Receiver: ", receiver_ct)
  ) +
  theme_minimal() +
  rose_theme +
  theme(
    legend.position = "right",
    plot.title = element_text(size = 12, face = "bold"),
    plot.subtitle = element_text(size = 10)
  )

# Save figure
ggsave(file.path(FIG_DIR, "representative_lr_scatterplot_tgfb1_tgfbr2.pdf"),
       width = 6, height = 5)
```



---


# Tier 2 with Spatial Connectivities (60μm radius)

Compare niche-level correlations using `spatial_connectivities` (60μm radius neighbors) vs `mingle_connectivities` (32 kNN). Larger neighborhoods may capture longer-range signaling (e.g., chemokine diffusion).


## load niche ligand data

```{r load_niche_ligand}

niche_lig_sp_df = arrow::read_parquet(file.path(SAVE_DIR, 'st_data_untreated_niche_DC_all_cancerOnly_k50_env_pseudobulk_celltype_expression_spatialConn.parquet'))
setDT(niche_lig_sp_df)
```


## Precompute Niche Receptor Fractions

```{r load_niche_receptor}

niche_pos_sp_df = arrow::read_parquet(file.path(SAVE_DIR, 'st_data_untreated_niche_DC_all_cancerOnly_k50_env_gene_positivity_spatialConn.parquet'))
setDT(niche_pos_sp_df)
```

## Tier 2 Correlation Testing (Spatial Connectivities)

```{r tier2_spatial_correlation_test}
cat("Running Tier 2 correlation tests with spatial_connectivities (60μm radius)...\n")

# Filter to cancer cells in analysis bins
cancer_tier2_df <- obs_df %>%
  filter(!is.na(DC_1_rank), in_analysis)

cancer_barcodes <- cancer_tier2_df$barcode
n_cancer <- length(cancer_barcodes)

# Filter parquet data to cancer cells
niche_pos_sp_df.cancer <- niche_pos_sp_df[barcode %in% cancer_barcodes]
niche_lig_sp_df.cancer <- niche_lig_sp_df[barcode %in% cancer_barcodes]

cat("Cancer cells for Tier 2 spatial testing:", n_cancer, "\n")

# Pre-index barcode positions for matrix construction
barcode_idx <- setNames(seq_along(cancer_barcodes), cancer_barcodes)

# Get LR genes
ligand_genes <- unique(LR_PAIRS$ligand)
receptor_genes <- unique(LR_PAIRS$receptor_gene)

tier2_sp_results <- list()

# Loop through cell type pairs first (fewer subsets)
for (sender_ct in SENDER_CELLTYPES) {
  print(sender_ct)
  # Subset ligand expression for this sender cell type (all genes at once)
  lig_ct_subset <- niche_lig_sp_df.cancer[cell_type == sender_ct & gene %in% ligand_genes]

  if (nrow(lig_ct_subset) == 0) next

  # Pivot to matrix: rows = barcodes, cols = genes
  lig_mat <- matrix(0, nrow = n_cancer, ncol = length(ligand_genes),
                    dimnames = list(cancer_barcodes, ligand_genes))
  for (g in intersect(ligand_genes, unique(lig_ct_subset$gene))) {
    g_subset <- lig_ct_subset[gene == g]
    lig_mat[barcode_idx[g_subset$barcode], g] <- g_subset$value
  }

  for (receiver_ct in RECEIVER_CELLTYPES) {
    # Subset receptor fraction for this receiver cell type (all genes at once)
    rec_ct_subset <- niche_pos_sp_df.cancer[cell_type == receiver_ct & gene %in% receptor_genes]

    if (nrow(rec_ct_subset) == 0) {
      # Record NA for all LR pairs with this cell type combo
      for (i in seq_len(nrow(LR_PAIRS))) {
        lig <- LR_PAIRS$ligand[i]
        if (!lig %in% colnames(lig_mat) || sum(lig_mat[, lig] > 0) == 0) next
        tier2_sp_results[[length(tier2_sp_results) + 1]] <- data.frame(
          sender_celltype = sender_ct,
          receiver_celltype = receiver_ct,
          ligand = lig,
          receptor = LR_PAIRS$receptor_gene[i],
          tier2_sp_rho_spearman = NA_real_,
          tier2_sp_pval_spearman = NA_real_,
          tier2_sp_rho_pearson = NA_real_,
          tier2_sp_pval_pearson = NA_real_,
          tier2_sp_n_cells = n_cancer,
          tier2_sp_n_lig_nonzero = sum(lig_mat[, lig] > 0),
          tier2_sp_n_rec_nonzero = 0L,
          stringsAsFactors = FALSE
        )
      }
      next
    }

    # Pivot to matrix: rows = barcodes, cols = genes
    rec_mat <- matrix(0, nrow = n_cancer, ncol = length(receptor_genes),
                      dimnames = list(cancer_barcodes, receptor_genes))
    for (g in intersect(receptor_genes, unique(rec_ct_subset$gene))) {
      g_subset <- rec_ct_subset[gene == g]
      rec_mat[barcode_idx[g_subset$barcode], g] <- g_subset$fraction
    }

    # Now loop through LR pairs - just column extraction
    for (i in seq_len(nrow(LR_PAIRS))) {
      lig <- LR_PAIRS$ligand[i]
      rec <- LR_PAIRS$receptor_gene[i]

      # Check genes exist in matrices
      if (!lig %in% colnames(lig_mat) || !rec %in% colnames(rec_mat)) next

      lig_vec <- lig_mat[, lig]
      rec_vec <- rec_mat[, rec]

      n_lig_nonzero <- sum(lig_vec > 0)
      n_rec_nonzero <- sum(rec_vec > 0)

      if (n_lig_nonzero < 100 || n_rec_nonzero < 100) {
        tier2_sp_results[[length(tier2_sp_results) + 1]] <- data.frame(
          sender_celltype = sender_ct,
          receiver_celltype = receiver_ct,
          ligand = lig,
          receptor = rec,
          tier2_sp_rho_spearman = NA_real_,
          tier2_sp_pval_spearman = NA_real_,
          tier2_sp_rho_pearson = NA_real_,
          tier2_sp_pval_pearson = NA_real_,
          tier2_sp_n_cells = n_cancer,
          tier2_sp_n_lig_nonzero = n_lig_nonzero,
          tier2_sp_n_rec_nonzero = n_rec_nonzero,
          stringsAsFactors = FALSE
        )
        next
      }

      # Correlate across all cancer cells (including zeros)
      cor_spearman <- cor.test(log1p(lig_vec), rec_vec, method = "spearman")
      cor_pearson <- cor.test(log1p(lig_vec), rec_vec, method = "pearson")

      tier2_sp_results[[length(tier2_sp_results) + 1]] <- data.frame(
        sender_celltype = sender_ct,
        receiver_celltype = receiver_ct,
        ligand = lig,
        receptor = rec,
        tier2_sp_rho_spearman = cor_spearman$estimate,
        tier2_sp_pval_spearman = cor_spearman$p.value,
        tier2_sp_rho_pearson = cor_pearson$estimate,
        tier2_sp_pval_pearson = cor_pearson$p.value,
        tier2_sp_n_cells = n_cancer,
        tier2_sp_n_lig_nonzero = n_lig_nonzero,
        tier2_sp_n_rec_nonzero = n_rec_nonzero,
        stringsAsFactors = FALSE
      )
    }
  }
}

tier2_sp_df <- bind_rows(tier2_sp_results)
rownames(tier2_sp_df) <- NULL

cat("Tier 2 spatial tests completed:", nrow(tier2_sp_df), "\n")
cat("Tests with sufficient non-zero observations:", sum(!is.na(tier2_sp_df$tier2_sp_rho_spearman)), "\n")
```

## Per-Sample Consistency Check (Spatial Connectivities)

Test whether LR colocalization is consistent across samples. For each sample, compute the same correlations and report effect sizes.

```{r tier2_sp_per_sample}
cat("Running per-sample Tier 2 correlation tests (spatial connectivities)...\n")

# Get unique samples
samples <- unique(obs_df$sample_id[!is.na(obs_df$sample_id)])
cat("Samples:", length(samples), "\n")
print(samples)

tier2_sp_sample_results <- list()

for (samp in samples) {
  cat("\nProcessing sample:", samp, "\n")

  # Filter to cancer cells in this sample and analysis bins
  sample_cancer_df <- obs_df %>%
    filter(!is.na(DC_1_rank), in_analysis, sample_id == samp)

  sample_barcodes <- sample_cancer_df$barcode
  n_sample <- length(sample_barcodes)

  if (n_sample < 100) {
    cat("  Skipping - only", n_sample, "cancer cells\n")
    next
  }
  cat("  Cancer cells:", n_sample, "\n")

  # Filter parquet data to this sample's cancer cells
  niche_pos_sp_samp <- niche_pos_sp_df[barcode %in% sample_barcodes]
  niche_lig_sp_samp <- niche_lig_sp_df[barcode %in% sample_barcodes]

  # Pre-index barcode positions
  sample_barcode_idx <- setNames(seq_along(sample_barcodes), sample_barcodes)

  # Loop through cell type pairs
  for (sender_ct in SENDER_CELLTYPES) {
    # Subset ligand expression for this sender cell type
    lig_ct_subset <- niche_lig_sp_samp[cell_type == sender_ct & gene %in% ligand_genes]

    if (nrow(lig_ct_subset) == 0) next

    # Pivot to matrix
    lig_mat <- matrix(0, nrow = n_sample, ncol = length(ligand_genes),
                      dimnames = list(sample_barcodes, ligand_genes))
    for (g in intersect(ligand_genes, unique(lig_ct_subset$gene))) {
      g_subset <- lig_ct_subset[gene == g]
      lig_mat[sample_barcode_idx[g_subset$barcode], g] <- g_subset$value
    }

    for (receiver_ct in RECEIVER_CELLTYPES) {
      # Subset receptor fraction for this receiver cell type
      rec_ct_subset <- niche_pos_sp_samp[cell_type == receiver_ct & gene %in% receptor_genes]

      if (nrow(rec_ct_subset) == 0) next

      # Pivot to matrix
      rec_mat <- matrix(0, nrow = n_sample, ncol = length(receptor_genes),
                        dimnames = list(sample_barcodes, receptor_genes))
      for (g in intersect(receptor_genes, unique(rec_ct_subset$gene))) {
        g_subset <- rec_ct_subset[gene == g]
        rec_mat[sample_barcode_idx[g_subset$barcode], g] <- g_subset$fraction
      }

      # Loop through LR pairs
      for (i in seq_len(nrow(LR_PAIRS))) {
        lig <- LR_PAIRS$ligand[i]
        rec <- LR_PAIRS$receptor_gene[i]

        if (!lig %in% colnames(lig_mat) || !rec %in% colnames(rec_mat)) next

        lig_vec <- lig_mat[, lig]
        rec_vec <- rec_mat[, rec]

        n_lig_nonzero <- sum(lig_vec > 0)
        n_rec_nonzero <- sum(rec_vec > 0)

        # Require at least 30 non-zero for per-sample (relaxed threshold)
        if (n_lig_nonzero < 30 || n_rec_nonzero < 30) {
          tier2_sp_sample_results[[length(tier2_sp_sample_results) + 1]] <- data.frame(
            sample_id = samp,
            sender_celltype = sender_ct,
            receiver_celltype = receiver_ct,
            ligand = lig,
            receptor = rec,
            rho_spearman = NA_real_,
            pval_spearman = NA_real_,
            rho_pearson = NA_real_,
            pval_pearson = NA_real_,
            n_cells = n_sample,
            n_lig_nonzero = n_lig_nonzero,
            n_rec_nonzero = n_rec_nonzero,
            stringsAsFactors = FALSE
          )
          next
        }

        # Correlate
        cor_spearman <- cor.test(log1p(lig_vec), rec_vec, method = "spearman")
        cor_pearson <- cor.test(log1p(lig_vec), rec_vec, method = "pearson")

        tier2_sp_sample_results[[length(tier2_sp_sample_results) + 1]] <- data.frame(
          sample_id = samp,
          sender_celltype = sender_ct,
          receiver_celltype = receiver_ct,
          ligand = lig,
          receptor = rec,
          rho_spearman = cor_spearman$estimate,
          pval_spearman = cor_spearman$p.value,
          rho_pearson = cor_pearson$estimate,
          pval_pearson = cor_pearson$p.value,
          n_cells = n_sample,
          n_lig_nonzero = n_lig_nonzero,
          n_rec_nonzero = n_rec_nonzero,
          stringsAsFactors = FALSE
        )
      }
    }
  }
}

tier2_sp_sample_df <- bind_rows(tier2_sp_sample_results)
rownames(tier2_sp_sample_df) <- NULL

cat("\n=== Per-Sample Results ===\n")
cat("Total tests:", nrow(tier2_sp_sample_df), "\n")
cat("Tests with results:", sum(!is.na(tier2_sp_sample_df$rho_spearman)), "\n")
```

```{r tier2_sp_sample_consistency}
# Summarize consistency across samples for each LR × cell type combination
cor_threshold = 0.1
consistency_df <- tier2_sp_sample_df %>%
  filter(!is.na(rho_spearman)) %>%
  group_by(sender_celltype, receiver_celltype, ligand, receptor) %>%
  summarize(
    n_samples = n(),
    n_positive = sum(rho_spearman > cor_threshold, na.rm = TRUE),
    n_sig_positive = sum(rho_spearman > cor_threshold & pval_spearman < 0.05, na.rm = TRUE),
    mean_rho_spearman = mean(rho_spearman, na.rm = TRUE),
    sd_rho_spearman = sd(rho_spearman, na.rm = TRUE),
    min_rho_spearman = min(rho_spearman, na.rm = TRUE),
    max_rho_spearman = max(rho_spearman, na.rm = TRUE),
    mean_rho_pearson = mean(rho_pearson, na.rm = TRUE),
    sd_rho_pearson = sd(rho_pearson, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    frac_positive = n_positive / n_samples,
    frac_sig_positive = n_sig_positive / n_samples,
    consistent = frac_positive >= 0.75 | frac_positive <= 0.25  # >75% agree on direction
  )

cat("\n=== Consistency Summary ===\n")
cat("LR × cell type combinations tested in >=2 samples:", sum(consistency_df$n_samples >= 2), "\n")
cat("Consistent direction (>=75% agree):", sum(consistency_df$consistent & consistency_df$n_samples >= 2), "\n")
cat("Consistently positive (>=75% samples positive):",
    sum(consistency_df$frac_positive >= 0.75 & consistency_df$n_samples >= 2), "\n")
cat("Consistently sig positive (>=75% samples sig positive):",
    sum(consistency_df$frac_sig_positive >= 0.75 & consistency_df$n_samples >= 2), "\n")

# Join consistency metrics to main tier2_sp_df
tier2_sp_df <- tier2_sp_df %>%
  left_join(
    consistency_df %>% select(sender_celltype, receiver_celltype, ligand, receptor,
                               n_samples, frac_positive, frac_sig_positive,
                               mean_rho_spearman, sd_rho_spearman, consistent),
    by = c("sender_celltype", "receiver_celltype", "ligand", "receptor")
  )

cat("\nJoined consistency metrics to tier2_sp_df\n")
```

```{r tier2_sp_sample_heatmap, fig.width=10, fig.height=8}
# Heatmap of per-sample correlations for top consistent LR pairs
top_consistent <- consistency_df %>%
  filter(n_samples >= 3, frac_positive >= 0.75, mean_rho_spearman > 0.05) %>%
  arrange(desc(mean_rho_spearman)) %>%
  head(30)

if (nrow(top_consistent) > 0) {
  # Get per-sample rhos for these pairs
  top_sample_data <- tier2_sp_sample_df %>%
    inner_join(top_consistent %>% select(sender_celltype, receiver_celltype, ligand, receptor),
               by = c("sender_celltype", "receiver_celltype", "ligand", "receptor")) %>%
    mutate(lr_label = paste0(ligand, "→", receptor, " (", sender_celltype, "→", receiver_celltype, ")"))

  # Pivot to matrix
  sample_mat <- top_sample_data %>%
    select(lr_label, sample_id, rho_spearman) %>%
    pivot_wider(names_from = sample_id, values_from = rho_spearman) %>%
    tibble::column_to_rownames("lr_label") %>%
    as.matrix()

  # Plot heatmap
  col_fun <- colorRamp2(c(-0.3, 0, 0.3), c("#4575b4", "white", "#d73027"))

  ht <- Heatmap(
    sample_mat,
    col = col_fun,
    name = "Spearman ρ",
    cluster_rows = TRUE,
    cluster_columns = TRUE,
    show_row_names = TRUE,
    row_names_gp = gpar(fontsize = 7),
    show_column_names = TRUE,
    column_names_gp = gpar(fontsize = 9),
    column_title = "Per-Sample LR Colocalization (Spatial 60μm)",
    na_col = "grey90",
    rect_gp = gpar(col = "lightgrey", lwd = 0.3)
  )

  draw(ht)

  # Save
  svg(file.path(FIG_DIR, "tier2_sp_per_sample_consistency_heatmap.svg"), width = 10, height = 8)
  draw(ht)
  dev.off()
  cat("Saved per-sample consistency heatmap\n")
}
```

```{r save_per_sample_results}
# Save per-sample results
write.csv(tier2_sp_sample_df, file.path(SAVE_DIR, "lr_dc1_tier2_sp_per_sample.csv"), row.names = FALSE)
write.csv(consistency_df, file.path(SAVE_DIR, "lr_dc1_tier2_sp_sample_consistency.csv"), row.names = FALSE)
cat("Saved per-sample results to:", SAVE_DIR, "\n")
```

```{r}
# read if necessary
consistency_df = readr::read_csv(file.path(SAVE_DIR, "lr_dc1_tier2_sp_sample_consistency.csv"))
```


---

# Join Spatial Consistency to Tier 1 Results

Join the per-sample consistency metrics from spatial connectivities to Tier 1 LM results for filtering.

## load data

```{r}
# load if necessary
lm_df_tier1 = read.csv(file.path(SAVE_DIR, "lr_dc1_lm_results_tier1.csv"), h = T, stringsAsFactors = F)
dc1_corr_df = read.csv(file.path(FIG_DIR, 'untreated_dc1_pearson_correlation.csv'), h = T, stringsAsFactors  = F)
# read if necessary
consistency_df = readr::read_csv(file.path(SAVE_DIR, "lr_dc1_tier2_sp_sample_consistency.csv"))
#tier1_data = readRDS(file.path(SAVE_DIR, "lr_bin_summaries_tier1.rds")) # need to run early part of preprocessing up to filtering diagnostics to get this
```


```{r join_consistency_to_tier1}
# Join consistency_df to lm_df_tier1
lm_df_tier1_with_consistency <- lm_df_tier1 %>%
  left_join(
    consistency_df %>% select(sender_celltype, receiver_celltype, ligand, receptor,
                               n_samples, n_positive, n_sig_positive,
                               frac_positive, frac_sig_positive,
                               mean_rho_spearman, sd_rho_spearman, consistent),
    by = c("sender_celltype", "receiver_celltype", "ligand", "receptor")
  )

cat("Joined consistency metrics to lm_df_tier1\n")
cat("Total associations:", nrow(lm_df_tier1_with_consistency), "\n")
cat("With consistency data:", sum(!is.na(lm_df_tier1_with_consistency$n_samples)), "\n")
cat("Number of ligands tested:", length(unique(lm_df_tier1_with_consistency$ligand)), "\n")
cat("Number of receptors tested:", length(unique(lm_df_tier1_with_consistency$receptor)), "\n")
cat("Number of cell types tested:", length(unique(lm_df_tier1_with_consistency$sender_celltype)), "\n") # should be equal between sender and receiver
```
```{r}
# write out for supp
write.csv(lm_df_tier1_with_consistency, file = file.path('basal_ablation/data/processed/xenium/niche_trajectory_classical_pruned_d24_filterv2_d1rm', 'lr_dc1_tier1_with_consistency.csv'), quote = F, row.names = F)
```


```{r tier1_vs_consistency_comparison}
# Define Tier 1 significance criteria
TIER1_PADJ <- 0.01
TIER1_SLOPE_MIN <- 0
TIER1_PCT_LIG_MIN <- 0.1
TIER1_PCT_REC_MIN <- 0.1
MIN_N_POSITIVE <- 2

# Identify Tier 1 significant associations
tier1_sig <- lm_df_tier1_with_consistency %>%
  filter(padj < TIER1_PADJ, slope > TIER1_SLOPE_MIN,
         pct_expressing.ligand > TIER1_PCT_LIG_MIN,
         pct_expressing.receptor > TIER1_PCT_REC_MIN)

cat("\n=== Tier 1 Significant Associations ===\n")
cat("Total Tier 1 significant:", nrow(tier1_sig), "\n")

# Apply n_positive filter
tier1_with_consistency <- tier1_sig %>%
  filter(!is.na(n_positive), n_positive >= MIN_N_POSITIVE)

cat("After n_positive >=", MIN_N_POSITIVE, "filter:", nrow(tier1_with_consistency), "\n")

# Identify removed associations
tier1_removed <- tier1_sig %>%
  filter(is.na(n_positive) | n_positive < MIN_N_POSITIVE)

cat("\n=== Associations REMOVED by n_positive filter ===\n
")
cat("Removed:", nrow(tier1_removed), "associations\n")

if (nrow(tier1_removed) > 0) {
  cat("\nRemoved associations (showing key columns):\n")
  tier1_removed %>%
    dplyr::select(sender_celltype, receiver_celltype, ligand, receptor, pathway,
           slope, padj, n_samples, n_positive, mean_rho_spearman) %>%
    dplyr::arrange(desc(slope)) %>%
    print(n = 50)

  # Breakdown by reason
  cat("\n--- Breakdown of removals ---\n")
  cat("No consistency data (NA):", sum(is.na(tier1_removed$n_positive)), "\n")
  cat("n_positive = 0:", sum(tier1_removed$n_positive == 0, na.rm = TRUE), "\n")
  cat("n_positive = 1:", sum(tier1_removed$n_positive == 1, na.rm = TRUE), "\n")
  cat("n_positive = 2:", sum(tier1_removed$n_positive == 2, na.rm = TRUE), "\n")

  # Summarize by ligand
  cat("\n--- Removed by ligand ---\n")
  tier1_removed %>%
    dplyr::count(ligand, sort = TRUE) %>%
    head(20) %>%
    print()

  # Summarize by sender-receiver pair
  cat("\n--- Removed by sender→receiver ---\n")
  tier1_removed %>%
    dplyr::count(sender_celltype, receiver_celltype, sort = TRUE) %>%
    head(15) %>%
    print()
  
  cat("\n--- Removed by sender ---\n")
  tier1_removed %>%
    dplyr::count(sender_celltype, sort = TRUE) %>%
    head(15) %>%
    print()
  
  cat("\n--- Removed by receiver ---\n")
  tier1_removed %>%
    dplyr::count(receiver_celltype, sort = TRUE) %>%
    head(15) %>%
    print()
  
  cat("\n--- how many distinct ligand pairs ---\n")
  tier1_removed %>%
    dplyr::select(ligand, receptor) %>%
    distinct() %>%
    nrow() %>%
    print()
  
}

# Save the filtered results
write.csv(tier1_with_consistency,
          file.path(SAVE_DIR, "lr_dc1_tier1_with_spatial_consistency_filtered.csv"),
          row.names = FALSE)
cat("\nSaved filtered results to:", file.path(SAVE_DIR, "lr_dc1_tier1_with_spatial_consistency_filtered.csv"), "\n")
```

---

# Ligand heatmaps (with spatial consistency filter)

```{r ligand_heatmaps_dc1_correlated, fig.width=12, fig.height=10}
# Get ligands with |rho| > 0.1 on DC1 from cancer cells
cancer_dc1_corr <- dc1_corr_df %>%
  filter(cell_type == "Cancer_epithelial", abs(rho) > 0.1)

# Get ligands with significant positive Tier 1 associations (cancer as sender)
# NOW USING CONSISTENCY-FILTERED DATA
sig_cancer_sender <- tier1_with_consistency %>%
  filter(sender_celltype == "Cancer_epithelial")

# Find ligands meeting both criteria
dc1_pos_ligands <- cancer_dc1_corr %>% filter(rho > 0.1) %>% pull(gene) %>% unique()
dc1_neg_ligands <- cancer_dc1_corr %>% filter(rho < -0.1) %>% pull(gene) %>% unique()
sig_tier1_ligands <- unique(sig_cancer_sender$ligand)

# Intersect with Tier 1 significant ligands
pos_corr_ligands <- intersect(dc1_pos_ligands, sig_tier1_ligands)
neg_corr_ligands <- intersect(dc1_neg_ligands, sig_tier1_ligands)

cat("DC1 positively correlated ligands (rho > 0.1) with sig Tier 1 + consistency:", length(pos_corr_ligands), "\n")
cat("DC1 negatively correlated ligands (rho < -0.1) with sig Tier 1 + consistency:", length(neg_corr_ligands), "\n")


# how many LR interactions in each direction
sig_pos <- tier1_with_consistency %>%
    filter(sender_celltype == "Cancer_epithelial",
           ligand %in% pos_corr_ligands) %>%
    nrow() %>%
    print()

sig_neg <- tier1_with_consistency %>%
    filter(sender_celltype == "Cancer_epithelial",
           ligand %in% neg_corr_ligands) %>%
    nrow() %>%
    print()

```

### with spatial consistency filter (n_positive >= 2)

```{r ligand_heatmap_dc1_positive_consistent, fig.width=12, fig.height=30}
# Heatmaps for DC1-positive correlated ligands (cancer sender, with consistency filter)
if (length(pos_corr_ligands) > 0) {
  sig_pos <- tier1_with_consistency %>%
    filter(sender_celltype == "Cancer_epithelial",
           ligand %in% pos_corr_ligands) %>%
    arrange(desc(slope))

  if (nrow(sig_pos) > 0) {
    cat("Plotting DC1-positive correlated ligands:", nrow(sig_pos), "pairs\n")

    plot_lr_heatmap_celltype(
      sig_df = sig_pos,
      sender_celltype_bins = sender_celltype_bins,
      rec_celltype_bins = rec_celltype_bins,
      cov_bins = cov_bins,
      DC1_N_BINS = DC1_N_BINS,
      title_suffix = "DC1+ Correlated Ligands (Cancer Sender, n_pos>=2)",
      file_suffix = "ligand_dc1_pos_cancer_sender_consistent",
      FIG_DIR = FIG_DIR,
      x_axis_label = X_LAB, 
      format = 'pdf'
    )
  }
} else {
  cat("No DC1-positive correlated ligands with significant Tier 1 + consistency.\n")
}
```

```{r ligand_heatmap_dc1_negative_consistent, fig.width=12, fig.height=20}
# Heatmaps for DC1-negative correlated ligands (cancer sender, with consistency filter)
if (length(neg_corr_ligands) > 0) {
  sig_neg <- tier1_with_consistency %>%
    filter(sender_celltype == "Cancer_epithelial",
           ligand %in% neg_corr_ligands) %>%
    arrange(desc(slope))

  if (nrow(sig_neg) > 0) {
    cat("Plotting DC1-negative correlated ligands:", nrow(sig_neg), "pairs\n")

    plot_lr_heatmap_celltype(
      sig_df = sig_neg,
      sender_celltype_bins = sender_celltype_bins,
      rec_celltype_bins = rec_celltype_bins,
      cov_bins = cov_bins,
      DC1_N_BINS = DC1_N_BINS,
      title_suffix = "DC1- Correlated Ligands (Cancer Sender, n_pos>=2)",
      file_suffix = "ligand_dc1_neg_cancer_sender_consistent",
      FIG_DIR = FIG_DIR,
      x_axis_label = X_LAB, 
      format = 'pdf'
    )
  }
} else {
  cat("No DC1-negative correlated ligands with significant Tier 1 + consistency.\n")
}
```



---

# Cytokine Ligand-Response Program Analysis

## Compute Program Scores for All Cell Types

```{r compute_program_scores}
# Compute mean program expression for each cell
cat("Computing cytokine program scores for all cell types...\n")

for (prog_name in names(cytokine_signatures)) {
  prog_genes <- cytokine_signatures[[prog_name]]
  prog_genes_in_panel <- prog_genes[prog_genes %in% var_names]

  if (length(prog_genes_in_panel) < 2) next

  # Get gene indices
  gene_idx <- which(var_names %in% prog_genes_in_panel)

  # Mean log1p expression across program genes
  obs_df[[paste0("program_", prog_name)]] <- rowMeans(log1p_mat[, gene_idx, drop = FALSE])
}

program_cols <- grep("^program_", names(obs_df), value = TRUE)
cat("Computed", length(program_cols), "program scores\n")
```



```{r cytokine_heatmap, fig.width=12, fig.height=10}
# Heatmap of significant positive correlations
sig_corr_pos <- corr_df %>%
  dplyr::filter(padj < 0.05, rho > 0.3) %>%
  arrange(program, desc(rho))

if (nrow(sig_corr_pos) > 0) {
  plot_cytokine_correlation_heatmap(
    corr_df = sig_corr_pos,
    expr_bins = sender_celltype_bins,
    cov_bins = cov_bins,
    DC1_N_BINS = DC1_N_BINS,
    title = "Ligand-Program Positive Correlations",
    file_suffix = "positive",
    FIG_DIR = FIG_DIR
  )
}
```

---




# Session Info

```{r session_info}
sessionInfo()
```
